### 2013-04-29 paper/paper.tex
added paper.tex

% TEMPLATE for Usenix papers, specifically to meet requirements of
%  USENIX '05
% originally a template for producing IEEE-format articles using LaTeX.
%   written by Matthew Ward, CS Department, Worcester Polytechnic Institute.
% adapted by David Beazley for his excellent SWIG paper in Proceedings,
%   Tcl 96
% turned into a smartass generic template by De Clarke, with thanks to
%   both the above pioneers
% use at your own risk.  Complaints to /dev/null.
% make it two column with no page numbering, default is 10 point

% Munged by Fred Douglis <douglis@research.att.com> 10/97 to separate
% the .sty file from the LaTeX source template, so that people can
% more easily include the .sty file into an existing document.  Also
% changed to more closely follow the style guidelines as represented
% by the Word sample file.

% Note that since 2010, USENIX does not require endnotes. If you want
% foot of page notes, don't include the endnotes package in the 
% usepackage command, below.

% This version uses the latex2e styles, not the very ancient 2.09 stuff.
\documentclass[letterpaper,twocolumn,10pt]{article}
\usepackage{usenix,epsfig,endnotes}
\begin{document}

%don't want date printed
\date{}

%make title bold and 14 pt font (Latex default is non-bold, 16 pt)
\title{\Large \bf Wonderful : A Terrific Application and Fascinating Paper}

%for single author (just remove % characters)
\author{
{\rm Your N.\ Here}\\
Your Institution
\and
{\rm Second Name}\\
Second Institution
% copy the following lines to add more authors
% \and
% {\rm Name}\\
%Name Institution
} % end author

\maketitle

% Use the following at camera-ready time to suppress page numbers.
% Comment it out when you first submit the paper for review.
\thispagestyle{empty}

\subsection*{Abstract}
Your Abstract Text Goes Here.  Just a few facts.
Whet our appetites.

\section{Introduction}

A paragraph of text goes here.  Lots of text.  Plenty of interesting
text. \\

More fascinating text. Features\endnote{Remember to use endnotes, not footnotes!} galore, plethora of promises.\\

\section{This is Another Section}

Some embedded literal typset code might 
look like the following :

{\tt \small
\begin{verbatim}
int wrap_fact(ClientData clientData,
              Tcl_Interp *interp,
              int argc, char *argv[]) {
    int result;
    int arg0;
    if (argc != 2) {
        interp->result = ``wrong # args'';
        return TCL_ERROR;
    }
    arg0 = atoi(argv[1]);
    result = fact(arg0);
    sprintf(interp->result,''%d'',result);
    return TCL_OK;
}
\end{verbatim}
}

Now we're going to cite somebody.  Watch for the cite tag.
Here it comes~\cite{Chaum1981,Diffie1976}.  The tilde character (\~{})
in the source means a non-breaking space.  This way, your reference will
always be attached to the word that preceded it, instead of going to the
next line.

\section{This Section has SubSections}
\subsection{First SubSection}

Here's a typical figure reference.  The figure is centered at the
top of the column.  It's scaled.  It's explicitly placed.  You'll
have to tweak the numbers to get what you want.\\

% you can also use the wonderful epsfig package...
\begin{figure}[t]
\begin{center}
\begin{picture}(300,150)(0,200)
\put(-15,-30){\special{psfile = fig1.ps hscale = 50 vscale = 50}}
\end{picture}\\
\end{center}
\caption{Wonderful Flowchart}
\end{figure}

This text came after the figure, so we'll casually refer to Figure 1
as we go on our merry way.

\subsection{New Subsection}

It can get tricky typesetting Tcl and C code in LaTeX because they share
a lot of mystical feelings about certain magic characters.  You
will have to do a lot of escaping to typeset curly braces and percent
signs, for example, like this:
``The {\tt \%module} directive
sets the name of the initialization function.  This is optional, but is
recommended if building a Tcl 7.5 module.
Everything inside the {\tt \%\{, \%\}}
block is copied directly into the output. allowing the inclusion of
header files and additional C code.'' \\

Sometimes you want to really call attention to a piece of text.  You
can center it in the column like this:
\begin{center}
{\tt \_1008e614\_Vector\_p}
\end{center}
and people will really notice it.\\

\noindent
The noindent at the start of this paragraph makes it clear that it's
a continuation of the preceding text, not a new para in its own right.

Now this is an ingenious way to get a forced space.
{\tt Real~$*$} and {\tt double~$*$} are equivalent.

Now here is another way to call attention to a line of code, but instead
of centering it, we noindent and bold it.\\

\noindent
{\bf \tt size\_t : fread ptr size nobj stream } \\

And here we have made an indented para like a definition tag (dt)
in HTML.  You don't need a surrounding list macro pair.
\begin{itemize}
\item[]  {\tt fread} reads from {\tt stream} into the array {\tt ptr} at
most {\tt nobj} objects of size {\tt size}.   {\tt fread} returns
the number of objects read. 
\end{itemize}
This concludes the definitions tag.

\subsection{How to Build Your Paper}

You have to run {\tt latex} once to prepare your references for
munging.  Then run {\tt bibtex} to build your bibliography metadata.
Then run {\tt latex} twice to ensure all references have been resolved.
If your source file is called {\tt usenixTemplate.tex} and your {\tt
  bibtex} file is called {\tt usenixTemplate.bib}, here's what you do:
{\tt \small
\begin{verbatim}
latex usenixTemplate
bibtex usenixTemplate
latex usenixTemplate
latex usenixTemplate
\end{verbatim}
}

\subsection{Last SubSection}

Well, it's getting boring isn't it.  This is the last subsection
before we wrap it up.

\section{Acknowledgments}

A polite author always includes acknowledgments.  Thank everyone,
especially those who funded the work.

\section{Availability}

It's great when this section says that MyWonderfulApp is free software, 
available via anonymous FTP from

\begin{center}
{\tt ftp.site.dom/pub/myname/Wonderful}\\
\end{center}

Also, it's even greater when you can write that information is also 
available on the Wonderful homepage at

\begin{center}
{\tt http://www.site.dom/\~{}myname/SWIG}
\end{center}

Now we get serious and fill in those references.  Remember you will
have to run latex twice on the document in order to resolve those
cite tags you met earlier.  This is where they get resolved.
We've preserved some real ones in addition to the template-speak.
After the bibliography you are DONE.

{\footnotesize \bibliographystyle{acm}
\bibliography{../common/bibliography}}

\theendnotes

\end{document}

### 2013-05-12 pysim/samplerate.py
Minor changes

@@ -2,6 +2,8 @@
 # This file attempts to implement the SampleRate bit rate selection algorithm 
 # as outlined in the JBicket MS Thesis.
 
+from __future__ import division
+
 from random import choice 
 from common import ieee80211_to_idx
 import common

### 2013-05-12 raw/collect.sh

@@ -2,7 +2,7 @@
 
 FILE=$1
 
-cat /proc/ath_rate
+cat /proc/ath_rate >/dev/null
 
 while true; do
     cat /proc/ath_rate

### 2013-05-12 pysim/common.py
Minstrel_pavel with balanced_EWMA

@@ -77,14 +77,14 @@ class BalancedEWMA:
             self.denom = denom
             newval = num / denom
         else:
-            avg_block = self.denom / blocks
+            avg_block = self.denom / self.blocks
             block_weight = denom / avg_block
             relweight = self.p / (1 - self.p)
 
             prob = num / denom
 
-            newval = (self.val * relweight + prob * blockweight) / \
-                     (relweight + blockweight)
+            newval = (self.val * relweight + prob * block_weight) / \
+                     (relweight + block_weight)
 
         self.blocks += 1
         self.val = newval

### 2013-05-12 pysim/minstrel_pavel.py

@@ -1,6 +1,8 @@
 # Colleen Josephson, 2013
 # This file attempts to implement the minstrel rate control algorithm
 
+# TODO : seed 1466742691193015552
+
 from random import randint
 from random import choice 
 import common
@@ -21,9 +23,6 @@ cw_min = 15
 cw_max = 1023
 segment_size = 6000
 max_retry = 7 #safe default specified in kernel code
-np = 0
-nd = 0
-probeFlag = False
 
 # The average back-off period, in milliseconds, for up to 8 attempts of a 802.11b unicast packet. 
 # TODO: find g data
@@ -97,7 +96,7 @@ class Rate:
         self.throughput = 0 #in bits per second
         self.last_update = 0.0 #timestamp of last update, ns(?)
         #ewma obj. can return probability of success. if you ask nicely.
-        self.ewma = common.EWMA(0.0, 100e6, 0.75) #100e6 = 100 ms,
+        self.ewma = common.BalancedEWMA(0.0, 100e6, 0.75) #100e6 = 100 ms
         self.succ_hist = 0 #total successes ever
         self.att_hist = 0 #total xmission attempts ever
         self.success = 0  #number of successful xmissions in cur time interval
@@ -153,7 +152,7 @@ rates = dict((r, Rate(r)) for r in [1, 2, 5.5, 6, 9, 11, 12, 18, 24, 36, 48, 54]
 # populating the retry chain during the next 100 ms. Note that the retry chain is 
 # described below.
 def apply_rate(cur_time): #cur_time is in nanoseconds
-    global npkts, nsuccess, nlookaround, NBYTES, currRate, NRETRIES, np, nd
+    global npkts, nsuccess, nlookaround, NBYTES, currRate, NRETRIES
     global bestThruput, nextThruput, bestProb, lowestRate, time_last_called
 
     if cur_time - time_last_called >= 1e8:
@@ -173,31 +172,11 @@ def apply_rate(cur_time): #cur_time is in nanoseconds
     # 2  | Random rate      | Best throughput   | Next best throughput
     # 3  | Best probability | Best probability  | Best probability
     # 4  | Lowest Baserate  | Lowest baserate   | Lowest baserate
-    #if randint(1,100) <= 10:
-    delta = (npkts*.1 - np + nd/2.0)
-    if delta  > 0: #random!
+    if randint(1,100) <= 10:
         #Analysis of information showed that the system was sampling too hard
         #at some rates. For those rates that never work (54mb, 500m range) 
[…]

### 2013-05-12 pysim/samplerate_longer.py
Samplerate fixes and samplerate_longer

@@ -14,6 +14,8 @@ NBYTES = 1500 #constant
 currRate = 54 #current best bitRate
 NRETRIES = 1
 
+MAXFAILS = 25
+
 # The average back-off period, in microseconds, for up to 8 attempts of a 802.11b unicast packet. 
 # TODO: find g data
 backoff = {0:0, 1:155, 2:315, 3:635, 4:1275, 5:2555, 6:5115, 7:5115, 8:5115, 9:5115,
@@ -111,7 +113,7 @@ def apply_rate(cur_time):
     # highest bit-rate that has not had 4 successive failures."
     if nsuccess == 0:
         for i, r in sorted(rates.items(), reverse=True):
-            if r.succFails < 4:
+            if r.succFails < MAXFAILS:
                 currRate = r.rate
                 return [(ieee80211_to_idx(currRate)[0], NRETRIES)]
 
@@ -125,7 +127,7 @@ def apply_rate(cur_time):
         #have a minimum packet transmission time lower than the
         #current bit-rate’s average transmission time."
         eligible = [r for i, r in rates.items()
-                    if r.losslessTX < cavgTX and r.succFails < 4]
+                    if r.losslessTX < cavgTX and r.succFails < MAXFAILS]
 
         if len(eligible) > 0:
             sampleRate = choice(eligible).rate #select random rate from eligible
@@ -250,7 +252,7 @@ def calculateMin():
 
     #set current rate to the one w/ min avg tx time
     c = rates[currRate]
-    if c.succFails > 4:
+    if c.succFails > MAXFAILS:
         c.avgTX = float("inf")
         #c = rates[1]
 
@@ -266,7 +268,7 @@ def calculateMin():
             c = r
             break
         #print("------------------------------------------------")
-        if c.avgTX > r.avgTX and r.succFails < 4:
+        if c.avgTX > r.avgTX and r.succFails < MAXFAILS:
             c = r
 
     currRate = c.rate

### 2013-05-13 raw2/packets.sh
New data got

@@ -1,0 +1,1 @@
+cat "$1" | grep -v : | cut -f10 -d\  | sort -n | uniq -c

### 2013-05-13 pysim/pavel.py
Renamed minstrel_pavel to just pavel; too long to type

@@ -104,6 +104,7 @@ class Rate:
         self.losslessTX = tx_time(rate, 1200) #microseconds, pktsize 1200 in kernel,
         self.ack = tx_time(rate, 10) #microseconds, assumes 1mbps ack rate
         self.window = [] #packets rcvd in last 10s
+        self.tban = 0
 
         #what is the difference between these?
         self.sample_limit = -1
@@ -172,7 +173,14 @@ def apply_rate(cur_time): #cur_time is in nanoseconds
     # 2  | Random rate      | Best throughput   | Next best throughput
     # 3  | Best probability | Best probability  | Best probability
     # 4  | Lowest Baserate  | Lowest baserate   | Lowest baserate
-    if randint(1,100) <= 10:
+
+    if rates[bestThruput].tban > 4:
+        #print("Temp ban on {}, switching to {}!".format(bestThruput, nextThruput))
+        rates[bestThruput].tban = 0
+        bestThruput = nextThruput
+        nextThruput = bestProb
+    
+    if randint(1,100) <= 10 or rates[bestThruput].tban > 4:
         #Analysis of information showed that the system was sampling too hard
         #at some rates. For those rates that never work (54mb, 500m range) 
         #there is no point in sending 10 sample packets (< 6 ms time). Consequently, 
@@ -240,6 +248,10 @@ def process_feedback(status, timestamp, delay, tries):
                 br.success = (br.success + 1) 
                 nsuccess = (nsuccess + 1) 
 
+                br.tban = 0
+            else:
+                br.tban += 1
+
             #instantiate pkt object
             p = Packet(timestamp, status, delay, bitrate)
             
@@ -249,8 +261,15 @@ def process_feedback(status, timestamp, delay, tries):
     if timestamp - time_last_called >= 1e8:
         self.update_stats(timestamp)
 
+USCNT = 0
 def update_stats(timestamp):
-    global bestThruput, nextThruput, bestProb, rates
+    global bestThruput, nextThruput, bestProb, rates, USCNT
+
+    USCNT += 1
+
+    if USCNT % 100 == 0:
+        for br in rates.values():
+            br.tban = 0
 
     for i, br in rates.items():
         if br.attempts: #prevents divide by 0

### 2013-05-13 raw2/spew.py
Allow broadcast sends

@@ -11,9 +11,13 @@ print("Spewing to {}:{}, {} bytes per packet".format(UDP_IP, UDP_PORT,
 
 sock = socket.socket(socket.AF_INET, socket.SOCK_DGRAM)
 
+if UDP_IP == "<broadcast>":
+    sock.setsockopt(socket.SOL_SOCKET, socket.SO_BROADCAST, 1)
+
 i = 0
 while True:
     PREFIX = "Seq {}\n".format(i).encode("utf-8")
     MESSAGE = PREFIX + bytes(random.randint(0, 255) for i in range(MESSAGE_LEN - len(PREFIX)))
     sock.sendto(MESSAGE, (UDP_IP, UDP_PORT))
+    print(PREFIX)
     i += 1

### 2013-05-13 paper/paper.tex
our awesome research paper

@@
 %don't want date printed
-\date{}
+\date{May 2013}
 
 %make title bold and 14 pt font (Latex default is non-bold, 16 pt)
-\title{\Large \bf Wonderful : A Terrific Application and Fascinating Paper}
-
+\title{\Large \bf 6.829 Final Project: A comparison of bit-rate selection algorithms}
 %for single author (just remove % characters)
 \author{
-{\rm Your N.\ Here}\\
-Your Institution
+{\rm Colleen Josephson}\\
+MIT
 \and
-{\rm Second Name}\\
-Second Institution
+{\rm Pavel Panchekha}\\
+MIT
 % copy the following lines to add more authors
@@
 \subsection*{Abstract}
-Your Abstract Text Goes Here.  Just a few facts.
-Whet our appetites.
-
+In this paper we compare the performance SampleRate and Minstrel, two popular bit rate selection algorithms that have widespread real-world usage on the Linux operating system.
+We use a novel trace-based approach that allows us avoid in-depth kernel programming and write simple Python scripts instead.
+We also introduce improvements to the Minstrel algorithm that allow for significant gains in throughput.
 \section{Introduction}
 
-A paragraph of text goes here.  Lots of text.  Plenty of interesting
-text. \\
+One of the key ways that wireless networks differ from wired is that wireless networks have varying link rates.
+Link conditions vary with time due to interference from other devices, changing network geometry, and mobile clients.
+An optimal rate at some time $t$ may be different from the optimal rate just 30 seconds earlier.
+A good bit rate selection algorithm has to detect and adapt to these conditions.
+If the chosen bit rate is too slow, then the throughput will be unnecessarily low.
+If the rate is too high, then failures will be very frequent and throughput will again suffer.
 
-More fascinating text. Features\endnote{Remember to use endnotes, not footnotes!} galore, plethora of promises.\\
+There are three main classes of bit rate adaptiation protocols: frame-based, SNR-based, and cross-layer protocols.
+Frame-based protocols measure the fraction of sucessfully received protocols.
+SNR protocols make decisions based on the estimated Signal to Noise ratio.
+Cross-layer protocols use SoftPHY data from the physical layer.
+The most commonly implemented protocols on today's networks are frame based, because SNR protocols perform poorly [CITATION HERE], and cross-layer protocols cannot be deployed on current networks because they violate the network layering abstration.
 
-\section{This is Another Section}
+The two most popular fram-based protocols are SampleRate and Minstrel.
+Both were implemented in the MadWifi drivers for Linux wirless, and today Minstrel is the default bit-rate selection algorithm for all wireless drivers on Linux.
+Our research used a trace-based approach to analyze the performance of these two algorithms.
+We also created a modified version of Minstrel that provides significant throughput gains over the vanilla Minstrel implementation.
 
-Some embedded literal typset code might 
-look like the following :
+This paper will provide a general overview of SampleRate and Minstrel.
+We then discuss our testing methodology.
+We compare the performance of the two algorithms, and then discuss modifications we made to Minstrel and compare it's performance to the current kernel implementation.
+\section{SampleRate}
 
-{\tt \small
-\begin{verbatim}
-int wrap_fact(ClientData clientData,
-              Tcl_Interp *interp,
-              int argc, char *argv[]) {
-    int result;
-    int arg0;
-    if (argc != 2) {
-        interp->result = ``wrong # args'';
-        return TCL_ERROR;
-    }
-    arg0 = atoi(argv[1]);
-    result = fact(arg0);
-    sprintf(interp->result,''%d'',result);
-    return TCL_OK;
-}
-\end{verbatim}
-}
+Sample all the rates~\cite{samplerate, sprout, minstrel}.
 
-Now we're going to cite somebody.  Watch for the cite tag.
-Here it comes~\cite{Chaum1981,Diffie1976}.  The tilde character (\~{})
-in the source means a non-breaking space.  This way, your reference will
-always be attached to the word that preceded it, instead of going to the
-next line.
+\section{Minstrel}
 
-\section{This Section has SubSections}
-\subsection{First SubSection}
+Derek Smithies, Ph-fucking-D.
 
-Here's a typical figure reference.  The figure is centered at the
-top of the column.  It's scaled.  It's explicitly placed.  You'll
-have to tweak the numbers to get what you want.\\
+\section{Methodology}
 
-% you can also use the wonderful epsfig package...
-\begin{figure}[t]
-\begin{center}
-\begin{picture}(300,150)(0,200)
-\put(-15,-30){\special{psfile = fig1.ps hscale = 50 vscale = 50}}
-\end{picture}\\
-\end{center}
-\caption{Wonderful Flowchart}
-\end{figure}
+We did some things using some methods.
 
-This text came after the figure, so we'll casually refer to Figure 1
-as we go on our merry way.
+\subsection{Motivation}
 
-\subsection{New Subsection}
+MadWifi is old and sucks plus we'd need to buy hardware off e-bay.
+The current Atheros drivers use the linux default, Minstrel.
+There is no kernel implementation of SampleRate and porting it woile have been extremely annoying (possibly impossible) given the time constraints.
+Etc.
+Etc.
+Blah Blah.
 
-It can get tricky typesetting Tcl and C code in LaTeX because they share
-a lot of mystical feelings about certain magic characters.  You
-will have to do a lot of escaping to typeset curly braces and percent
-signs, for example, like this:
-``The {\tt \%module} directive
-sets the name of the initialization function.  This is optional, but is
-recommended if building a Tcl 7.5 module.
-Everything inside the {\tt \%\{, \%\}}
-block is copied directly into the output. allowing the inclusion of
-header files and additional C code.'' \\
+\subsection{Trace Collection}
 
-Sometimes you want to really call attention to a piece of text.  You
-can center it in the column like this:
-\begin{center}
-{\tt \_1008e614\_Vector\_p}
-\end{center}
-and people will really notice it.\\
+Yeee haw.
 
-\noindent
-The noindent at the start of this paragraph makes it clear that it's
-a continuation of the preceding text, not a new para in its own right.
+\subsection{Testing framework}
 
-Now this is an ingenious way to get a forced space.
-{\tt Real~$*$} and {\tt double~$*$} are equivalent.
+Harness.py, etc.
 
-Now here is another way to call attention to a line of code, but instead
-of centering it, we noindent and bold it.\\
-
-\noindent
-{\bf \tt size\_t : fread ptr size nobj stream } \\
-
-And here we have made an indented para like a definition tag (dt)
-in HTML.  You don't need a surrounding list macro pair.
-\begin{itemize}
-\item[]  {\tt fread} reads from {\tt stream} into the array {\tt ptr} at
-most {\tt nobj} objects of size {\tt size}.   {\tt fread} returns
-the number of objects read. 
-\end{itemize}
-This concludes the definitions tag.
-
-\subsection{How to Build Your Paper}
-
-You have to run {\tt latex} once to prepare your references for
-munging.  Then run {\tt bibtex} to build your bibliography metadata.
-Then run {\tt latex} twice to ensure all references have been resolved.
-If your source file is called {\tt usenixTemplate.tex} and your {\tt
-  bibtex} file is called {\tt usenixTemplate.bib}, here's what you do:
-{\tt \small
-\begin{verbatim}
-latex usenixTemplate
-bibtex usenixTemplate
-latex usenixTemplate
-latex usenixTemplate
-\end{verbatim}
-}
-
-\subsection{Last SubSection}
+\section{Analysis}
 
@@
 
-\section{Acknowledgments}
+\subsection{Improvements to Minstrel}
+Minproved.
 
-A polite author always includes acknowledgments.  Thank everyone,
-especially those who funded the work.
+\section{Future Work}
 
-\section{Availability}
+802.11n, SampleRate with EWMA, take data using broadcast approach,  submitting minstrel changes as kernel patch?
 
-It's great when this section says that MyWonderfulApp is free software, 
-available via anonymous FTP from
+\section{Conclusion}
 
-\begin{center}
-{\tt ftp.site.dom/pub/myname/Wonderful}\\
-\end{center}
-
-Also, it's even greater when you can write that information is also 
-available on the Wonderful homepage at
-
-\begin{center}
-{\tt http://www.site.dom/\~{}myname/SWIG}
-\end{center}
-
-Now we get serious and fill in those references.  Remember you will
-have to run latex twice on the document in order to resolve those
-cite tags you met earlier.  This is where they get resolved.
-We've preserved some real ones in addition to the template-speak.
-After the bibliography you are DONE.
+The end.
 
 {\footnotesize \bibliographystyle{acm}
-\bibliography{../common/bibliography}}
+\bibliography{paper}}
 
-\theendnotes
+\section{Acknowledgements}
+Jonathan Perry, Hari Balikrishnan, and Derek Smithies.
 

### 2013-05-13 pysim/samplerate.py

@@ -123,7 +123,7 @@ def apply_rate(cur_time):
 
         #" that have not failed four successive times and that
         #have a minimum packet transmission time lower than the
-        #current bit-rate’s average transmission time."
+        #current bit-rate's average transmission time."
         eligible = [r for i, r in rates.items()
                     if r.losslessTX < cavgTX and r.succFails < 4]
 
@@ -193,7 +193,7 @@ def process_feedback(status, timestamp, delay, tries):
     p = Packet(timestamp, status, tx, bitrate)
     br.window.append(p)
 
-#"SampleRate’s remove stale results() function removes results from
+#"SampleRate's remove stale results() function removes results from
 # the transmission results queue that were obtained longer than ten
 # seconds ago."
 def remove_stale_results(cur_time):

### 2013-05-13 pysim/harness.py
neatened up the repo, possibly fit for public consumption now?

@@ -1,4 +1,5 @@
-import numpy
+from __future__ import print_function
+from __future__ import division
 import time
 import common
 import os, sys
@@ -6,38 +7,49 @@ import random
 
 DEBUG = "DEBUG" in os.environ
 
+WINDOW = 1e7 # 10ms
+
 def load_data(source):
-    return numpy.load(source) / 1. # Convert to float
+    return eval(open(source, "rt").read())
 
-def packet_stats(data, time, rate):
-    time = time % len(data)
+LUOPS = 0
+def packet_stats(data, cache, time, rate):
+    global LUOPS
     
-    # First, we try to find a prior instance of such transmission
-    for i in range(time, -1, -1):
-        if data[i, rate].any():
-            delay, tries = data[i, rate]
-            return (delay, tries)
+    txs = []
 
-    # Next, try to find a future instance
-    for i in range(len(data)-1, time, -1):
-        if data[i, rate].any():
-            delay, tries = data[i, rate]
-            return (delay, tries)
-    
-    # Otherwise, this seems like a big problem, so we cause an error
-    raise ValueError("No data available at bit-rate", rate)
+    while not txs:
+        idx, w = cache[rate]
+        cache[rate][0] = 0
+        
+        for i in range(idx, len(data[rate])):
+            t, success, delay = data[rate][i]
+            LUOPS += 1
+            
+            if abs(t - time) < w:
+                txs.append((success, delay))
+                if not cache[rate][0]:
+                    cache[rate][0] = i
+                    cache[rate][1] = min(w / 1.5, WINDOW)
 
-def baseline(data):
-    s = data.sum(0)[0]
-    return s[0] / s[1]
+            if t > w + time:
+                break
+
+        cache[rate][1] *= 2
+
+    successful = [tx[1] for tx in txs if tx[0]]
+
+    return (len(successful) / len(txs), 
+            sum(successful) / sum([tx[1] for tx in txs]))
 
 class Harness:
     def __init__(self, data, choose_rate, push_statistics):
-        # Keep the time somewhat believable
-        #self.clock = time.time()
-        self.clock = 0
-        self.packet_num = 0
-        self.baseline = baseline(data)
+        self.start = data[0]
+        self.data = data[1]
+        self.cache = [[0, WINDOW] for i in self.data]
+        self.end = data[2]
+
+        self.clock = data[0]
         self.choose_rate = choose_rate
         self.push_statistics = push_statistics
 
@@ -56,22 +68,24 @@ class Harness:
         tot_tries = []
         tot_status = None
[…]

### 2013-05-13 pysim/harness2.py

@@ -1,167 +1,0 @@
-from __future__ import print_function
-from __future__ import division
-import time
-import common
-import os, sys
-import random
-
-DEBUG = "DEBUG" in os.environ
-
-WINDOW = 1e7 # 10ms
-
-def load_data(source):
-    return eval(open(source, "rt").read())
-
-LUOPS = 0
-def packet_stats(data, cache, time, rate):
-    global LUOPS
-    
-    txs = []
-
-    while not txs:
-        idx, w = cache[rate]
-        cache[rate][0] = 0
-        
-        for i in range(idx, len(data[rate])):
-            t, success, delay = data[rate][i]
-            LUOPS += 1
-            
-            if abs(t - time) < w:
-                txs.append((success, delay))
-                if not cache[rate][0]:
-                    cache[rate][0] = i
-                    cache[rate][1] = min(w / 1.5, WINDOW)
-
-            if t > w + time:
-                break
-
-        cache[rate][1] *= 2
-
-    successful = [tx[1] for tx in txs if tx[0]]
-
-    return (len(successful) / len(txs), 
-            sum(successful) / sum([tx[1] for tx in txs]))
-
-class Harness:
-    def __init__(self, data, choose_rate, push_statistics):
-        self.start = data[0]
-        self.data = data[1]
-        self.cache = [[0, WINDOW] for i in self.data]
-        self.end = data[2]
-
-        self.clock = data[0]
-        self.choose_rate = choose_rate
-        self.push_statistics = push_statistics
-
-        self.histogram = [0] * len(common.RATES)
-
-    def send_packet(self):
-        rate_arr = self.choose_rate(self.clock)
-
-        if DEBUG:
-            print("Sending packet at:", end=" ")
-            for (rate, tries) in rate_arr:
-                print("Rate {}/{}".format(rate, tries), end=" ")
-            print()
-
-        tot_delay = 0
-        tot_tries = []
-        tot_status = None
-        for (rate, tries) in rate_arr:
-            p_success, a_delay = packet_stats(self.data, self.cache, self.clock, rate)
-
-            s_tries = 0
-            succeeded = False
-            for i in range(tries):
-                success = random.random() < p_success
-                s_tries += 1
-
-                if success:
-                    succeeded = True
-                    break
-
-            tot_tries.append((rate, s_tries))
-            self.histogram[rate] += s_tries
-            tot_delay += s_tries * common.tx_time(rate, 1500)
-
-            if succeeded:
[…]

### 2013-05-13 pysim/minstrel.py

@@ -1,5 +1,7 @@
 # Colleen Josephson, 2013
-# This file attempts to implement the minstrel rate control algorithm
+# This file attempts to implement the minstrel rate control algorithm from the
+# 3.3.8 linux kernel. We assume multi-rate retry capabilities, so we omit 
+# the code for the non-mrr case. 
 
 from random import randint
 from random import choice 

### 2013-05-13 pysim/samplerate2.py

@@ -1,212 +1,0 @@
-# Colleen Josephson, 2013
-# This file implements a slightly modified version (I think) of the JBicket MS thesis
-# It uses retry chain for the initial send, and seems to get a slightly higher throughput
-# interestingly, adding retry chains for normal sends seems to decrease throughput
-
-from random import choice 
-from common import ieee80211_to_idx
-import common
-
-npkts = 0 #number of packets sent over link
-nsuccess = 0 #number of packets sent successfully 
-NBYTES = 1500 #constant
-currRate = 54 #current best bitRate
-NRETRIES = 20
-
-# The average back-off period, in microseconds, for up to 8 attempts of a 802.11b unicast packet. 
-# TODO: find g data
-backoff = {0:0, 1:155, 2:315, 3:635, 4:1275, 5:2555, 6:5115, 7:5115, 8:5115, 9:5115,
-           10:5115, 11:5115, 12:5115, 13:5115, 14:5115, 15:5115, 16:5115, 17:5115,
-           18:5115, 19:5115, 20:5115}
-
-#To calculate the transmission time of a n-byte unicast packet given the bit-rate b and
-#number of retries r, SampleRate uses the following equation based on the 802.11 unicast
-# tx_time(b, r, n) =  difs + backoff[r] + (r + 1)*(sifs + ack + header + (n * 8/b))
-def tx_time(bitrate, retries, nbytes):
-    global currRate, npkts, nsuccess, NBYTES
-    difs = 28 #DCF Interframe Space (DIFS), 28 microseconds in 802.11g
-    sifs = 9 #Short Interframe Space (SIFS), 9 microseconds for 802.11g
-    ack = 200 #in microseconds, for 6 megabit acknowledgements
-    header = 20 #in microseconds, for 802.11 a/g bitrates
-    return difs + backoff[retries] + (retries+1)*(sifs + ack + header + (nbytes * 8/bitrate))
-
-class Packet:
-    def __init__(self, time_sent, success, txTime, rate):
-        self.time_sent = time_sent
-        self.success = success
-        self.txTime = txTime
-        self.rate = rate
-
-    def __repr__(self):
-        return ("Pkt sent at time %r, rate %r was successful: %r\n" 
-                % (self.time_sent, self.rate, self.success))
-
-
-class Rate:
-    def __init__(self, rate):
-        self.rate = rate #in mbps
-        self.success = 0
-        self.tries = 0
-        self.pktAcked = 0
[…]

### 2013-05-13 pysim/samplerate_retry.py

@@ -1,0 +1,276 @@
+# Colleen Josephson, 2013
+# This file attempts to implement the SampleRate bit rate selection algorithm 
+# as outlined in the JBicket MS Thesis.
+
+from __future__ import division
+
+from random import choice 
+from common import ieee80211_to_idx
+import common
+
+npkts = 0 #number of packets sent over link
+nsuccess = 0 #number of packets sent successfully 
+NBYTES = 1500 #constant
+currRate = 54 #current best bitRate
+NRETRIES = 1
+
+# The average back-off period, in microseconds, for up to 8 attempts of a 802.11b unicast packet. 
+# TODO: find g data
+backoff = {0:0, 1:155, 2:315, 3:635, 4:1275, 5:2555, 6:5115, 7:5115, 8:5115, 9:5115,
+           10:5115, 11:5115, 12:5115, 13:5115, 14:5115, 15:5115, 16:5115, 17:5115,
+           18:5115, 19:5115, 20:5115}
+
+def bitrate_type(bitrate):
+    return common.RATES[ieee80211_to_idx(bitrate)[0]].phy
+
+
+
+#"To calculate the transmission time of a n-byte unicast packet given the bit-rate b and
+# number of retries r, SampleRate uses the following equation based on the 802.11 unicast
+# retransmission mechanism detailed in Section 2.2"
+#
+# tx_time(b, r, n) =  difs + backoff[r] + (r + 1)*(sifs + ack + header + (n * 8/b))
+def tx_time(bitrate, retries, nbytes):
+    # bitrate in MBPS, since 1*10^6 bps / 10-6 seconds/microseconds = 1 bit per microsecond
+    global currRate, npkts, nsuccess, NBYTES
+
+    brtype = bitrate_type(bitrate)
+    if bitrate == 1:
+        difs = 50
+        sifs = 10
+        ack = 304
+        header = 192
+    elif brtype == "cck":
+        difs = 50
+        sifs = 10
+        ack = 304
+        header = 96
+    elif brtype == "ofdm":
+        difs = 28
+        sifs = 9
+        ack = 304 # Somehow 6mb acks aren't used
+        header = 20
+    else:
+        raise ValueError("Unknown bitrate type", brtype, bitrate)
+
+    return difs + backoff[retries] + (retries+1)*(sifs + ack + header + (nbytes * 8/(bitrate)))
+
+class Packet:
+    def __init__(self, time_sent, success, txTime, rate):
+        self.time_sent = time_sent
[…]

### 2013-05-13 raw/collect.sh

@@ -2,7 +2,7 @@
 
 FILE=$1
 
-cat /proc/ath_rate >/dev/null
+cat /proc/ath_rate
 
 while true; do
     cat /proc/ath_rate

### 2013-05-13 raw/raw_old/spew.py

@@ -11,13 +11,6 @@ print("Spewing to {}:{}, {} bytes per packet".format(UDP_IP, UDP_PORT,
 
 sock = socket.socket(socket.AF_INET, socket.SOCK_DGRAM)
 
-if UDP_IP == "<broadcast>":
-    sock.setsockopt(socket.SOL_SOCKET, socket.SO_BROADCAST, 1)
-
-i = 0
 while True:
-    PREFIX = "Seq {}\n".format(i).encode("utf-8")
-    MESSAGE = PREFIX + bytes(random.randint(0, 255) for i in range(MESSAGE_LEN - len(PREFIX)))
+    MESSAGE = bytes(random.randint(0, 255) for i in range(MESSAGE_LEN))
     sock.sendto(MESSAGE, (UDP_IP, UDP_PORT))
-    print(PREFIX)
-    i += 1

### 2013-05-13 raw/spew.py

@@ -11,6 +11,13 @@ print("Spewing to {}:{}, {} bytes per packet".format(UDP_IP, UDP_PORT,
 
 sock = socket.socket(socket.AF_INET, socket.SOCK_DGRAM)
 
+if UDP_IP == "<broadcast>":
+    sock.setsockopt(socket.SOL_SOCKET, socket.SO_BROADCAST, 1)
+
+i = 0
 while True:
-    MESSAGE = bytes(random.randint(0, 255) for i in range(MESSAGE_LEN))
+    PREFIX = "Seq {}\n".format(i).encode("utf-8")
+    MESSAGE = PREFIX + bytes(random.randint(0, 255) for i in range(MESSAGE_LEN - len(PREFIX)))
     sock.sendto(MESSAGE, (UDP_IP, UDP_PORT))
+    print(PREFIX)
+    i += 1

### 2013-05-14 data/bar.py
plts

@@ -1,0 +1,34 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+N = 12
+throughput = (.806, 1.632, 4.716, 5.118, 7.888, 8.784, 9.865, 15.156, 1.174, 1.960, 0, 1.189)
+
+ind = np.arange(N)  # the x locations for the groups
+width = 0.5       # the width of the bars
+
+fig = plt.figure()
+ax = fig.add_subplot(111)
+rects1 = ax.bar(ind, throughput, width, color=['red', 'orange', 'DarkGoldenRod', 'yellow', 'lawngreen', 'DarkGreen', 'blue', 'DarkBlue', 'indigo',  'purple', 'DarkMagenta', 'gray'] )
+
+# add some
+ax.set_ylim([0,17])
+ax.set_xlim([-0.5, 12])
+ax.set_ylabel('Throughput in Mbps')
+ax.set_title('Throughput for Constant Bitrate')
+ax.set_xticks(ind+width)
+ax.set_xticklabels( ('1', '2', '5.5', '6', '9', '11', '12', '18', '24', '36', '48', '54') )
+
+#ax.legend( (rects1[0], rects2[0]), ('Men', 'Women') )
+
+def autolabel(rects):
+    # attach some text labels
+    for rect in rects:
+        height = rect.get_height()
+        ax.text(rect.get_x()+rect.get_width()/2., 1.05*height, '%r'% (height),
+                ha='center', va='bottom')
+
+autolabel(rects1)
+
+
+plt.show()

### 2013-05-14 paper/paper.tex
some more changes
blah

@@
 \documentclass[letterpaper,twocolumn,10pt]{article}
-\usepackage{usenix,epsfig,endnotes}
+\usepackage{epsfig,graphicx,usenix,fullpage}
+
+%\usepackage{endnotes}
 \begin{document}
@@
 \author{
-{\rm Colleen Josephson}\\
+{\rm Colleen Josephson\footnote{Supported in part by Lindt Chocolatier.}}\\
 MIT
@@
 % Comment it out when you first submit the paper for review.
-\thispagestyle{empty}
+%\thispagestyle{empty}
 
@@
 Yeee haw.
+\begin{figure}[htb]
+  \hspace{-4em}\includegraphics[width=3.75in]{constant.png}\vspace{-2em}
+  \caption{dfg}
+\end{figure}
 
@@
 
-Well, it's getting boring isn't it.  This is the last subsection
-before we wrap it up.
+Minstrel actually is better than SampleRate (except on short traces). Eit.
 
@@
 Minproved.
+
+\subsection{Availability}
+We should make our github repo or something like that available here.
 

### 2013-05-14 paper/paper.tex
updated the paper

@@
 \documentclass[letterpaper,twocolumn,10pt]{article}
-\usepackage{epsfig,graphicx,usenix,fullpage}
+\usepackage{epsfig,graphicx,usenix}
 
@@
 
+SampleRate sends at the rate with the lowest average transmission time.
+Every ten packets, it sends out a sample packet at a random rate that has a smaller predicted transmission time than the current rate.
+SampleRate updates its transmission time estimates every ten seconds based on feedback from the card about which packets succeeded and at which rate they were sent.
+If the current best rate experiences four successive failures, SampleRate stops using that rate.
+
 Sample all the rates~\cite{samplerate, sprout, minstrel}.
@@
 
-Derek Smithies, Ph-fucking-D.
+Minstrel utilizes the multi-rate retry chain (mrr), which is an array that tells the card which rates to try before reporting a failure.
+The retry chain makes failures extremely unlikely.
+Minstrel has a mrr of [highest throughput, next highest throughput, best probability, lowest base rate] for most packets.
+In essence, Minstrel tells the card to send at the highest throughput rates but to fall back to reliable rates upon failure.
+The throughput and success probability are calculated for each rate every 100ms.
+To keep an accurate estimate of throughput and probability for each rate, Minstrel sends sample frames ten percent of the time.
+If the randomly chosen rate has a higher lossless throughput than the current optimal rate, the mrr is as follows: [random, highest throughput, best probability, lowest base rate].
+If the random rate is slower than the optimal rate, it is moved further back in the retry chain: [best throughput, random, best probability, lowest base rate].
+This ensures that Minstrel never samples rates worse than the current optimal rate unless the optimal rate experiences a failure.
 
@@
 
+\section{Acknowledgements}
+Jonathan Perry, Hari Balikrishnan, and Derek Smithies.
+
 {\footnotesize \bibliographystyle{acm}
@@
 
-\section{Acknowledgements}
-Jonathan Perry, Hari Balikrishnan, and Derek Smithies.
-
 \end{document}

### 2013-05-14 paper/paper.tex
done for the night

@@
 
-We did some things using some methods.
+We used a trace-based approach to collect data, and then replayed that data to the rate-selection algorithms using a Python framework. Th
 
@@
 
-MadWifi is old and sucks plus we'd need to buy hardware off e-bay.
-The current Atheros drivers use the linux default, Minstrel.
-There is no kernel implementation of SampleRate and porting it woile have been extremely annoying (possibly impossible) given the time constraints.
-Etc.
-Etc.
-Blah Blah.
+Our initial plan was to use the algorithms as implemented in the MadWifi drivers for the Atheros chipset.
+However, MadWifi is depricated, and only runs on quite old network cards.
+In addition to the lack of compatible hardware available to us, MadWifi is not an accurate reflection of what a modern user would use.
+We looked into the ath5k and ath9k projects, which were created to replace MadWifi.
+Both of these did not port the old bit-rate selection system from MadWifi and instead opted to use the linux kernel's rate selection software, which only offers Minstrel.
+Porting SampleRate over to the new drivers would not have been trivial, as there are vast differences in the assumed interfaces.
+
+At this point, we decided to take a different approach and use an approach similar to that used in Sprout.
 

### 2013-05-14 paper/paper.tex
added to paper

@@
 \documentclass[letterpaper,twocolumn,10pt]{article}
-\usepackage{epsfig,graphicx,usenix}
+\usepackage{epsfig,graphicx,usenix,fullpage, hyperref}
 
@@
 %don't want date printed
-\date{May 2013}
 
@@
 \author{
-{\rm Colleen Josephson\footnote{Supported in part by Lindt Chocolatier.}}\\
-MIT
+{\rm Colleen Josephson}\\
+cjoseph@mit.edu
 \and
 {\rm Pavel Panchekha}\\
-MIT
+pavpan@mit.edu
 % copy the following lines to add more authors
@@
 
+SampleRate has three main functions: \texttt{ApplyRate()}, which returns a the bit-rate to send a packet at, \texttt{ProcessFeedback()}, which updates the statistics for a bitrate after an attempted packet send,  and \texttt{RemoveStaleResults()}, which removes all statistics older than 10 seconds.
+
 SampleRate sends at the rate with the lowest average transmission time.
@@
 
-Sample all the rates~\cite{samplerate, sprout, minstrel}.
-
 \section{Minstrel}
@@
 The retry chain makes failures extremely unlikely.
-Minstrel has a mrr of [highest throughput, next highest throughput, best probability, lowest base rate] for most packets.
+Minstrel has a mrr of [\emph{highest throughput, next highest throughput, best probability, lowest base rate}] for most packets.
 In essence, Minstrel tells the card to send at the highest throughput rates but to fall back to reliable rates upon failure.
@@
 To keep an accurate estimate of throughput and probability for each rate, Minstrel sends sample frames ten percent of the time.
-If the randomly chosen rate has a higher lossless throughput than the current optimal rate, the mrr is as follows: [random, highest throughput, best probability, lowest base rate].
-If the random rate is slower than the optimal rate, it is moved further back in the retry chain: [best throughput, random, best probability, lowest base rate].
+If the randomly chosen rate has a higher lossless throughput than the current optimal rate, the mrr is as follows: [\emph{random, highest throughput, best probability, lowest base rate}].
+If the random rate is slower than the optimal rate, it is moved further back in the retry chain: [\emph{best throughput, random, best probability, lowest base rate}].
 This ensures that Minstrel never samples rates worse than the current optimal rate unless the optimal rate experiences a failure.
+See table ~\ref{table:1} for an overview of the mrr's used in Minsterl.
+
+Minstrel's implementation is based on SampleRate's, so it also has \texttt{ApplyRate()} and \texttt{ProcessFeedback()}, as well as a \texttt{UpdateStats()} method that runs every 100ms.
+\texttt{UpdateStats()} is home to some of the key differences between Minstrel and SampleRate.
+Instead of making decisions based on the average
+transmission time, Minstrel uses throughput: \begin{equation} T = \frac{p}{tx\_time}\end{equation} where $T$ is the throughput of some rate $r$, $p$ is the probability of successfully transmitting at $r$, and $tx\_time$ is the lossless transmission time at $r$.
+
+The probability $p$ of a successful transmission is calculated from statistics collected by the \texttt{ProcessFeedback()} method, and is done using an exponential weighted moving average, or EWMA.
+The EWMA creates a weighted average that weighs recent data more heavily than old data.
+The weighting for old data decreases exponentially.
+An EWMA is beneficial because it ensures that a sudden degredation in link quality will create a rapid response in the probabilities due to the increased weight of recent data.
+Minstrel will waste less time sending packets at a rate that no longer works, as compared to a simple probability calculation.
+
+\begin{table*}[htb]
+    \begin{tabular}[htb]{l|l|l|l}
+    Try & Normal               & Random (slower than best) & Random (faster than best) \\ \hline
+    1   & Best throughput      & Best throughput           & Random                    \\
+    2   & Next best throughput & Random                    & Best throughput           \\
+    3   & Highest probability  & Highest probability       & Highest probability       \\
+    4   & Lowest rate          & Lowest rate               & Lowest rate               \\
+    \end{tabular}
+    \caption{Multi-rate retry chains for Minstrel}
+\label{table:1}
+\end{table*}
 
@@
 
+\begin{figure}[hbt]
+  \hspace{-0.5em}\includegraphics[width=3.5in]{bsmt13floorplan.png}\vspace{-0em}
+  \caption{We collected most traces at a remote access point in the basement of Building 13 at MIT.}
+\end{figure}
+
 Yeee haw.
 \begin{figure}[htb]
-  \hspace{-4em}\includegraphics[width=3.75in]{constant.png}\vspace{-2em}
-  \caption{dfg}
+  \hspace{-.75em}\includegraphics[width=3.5in]{constant.png}\vspace{-0em}
+  \caption{Throughput of using a constant 802.11b/g bitrate in an MIT basement.}
 \end{figure}
@@
 Harness.py, etc.
+
+We implemented SampleRate as outlined in John Bicket's Master's thesis~\cite{samplerate}.
+This implementation is slightly different from how it was implemented in the MadWifi kernel driver.
+The primary difference is that the kernel implementatiion uses a EWMA instead of a 10 second window.
+The thesis implmenentation throws out old rate data and begins anew every 10 seconds.
+It would be worth looking at the performance of EWMA-based SampleRate in the future.
+
+\texttt{samplerate.py} contains \texttt{ApplyRate()}, \texttt{ProcessFeedback()} and  \texttt{RemoveStaleResults()}, as well as a few helper functions and data structures for tracking rate statistics.
+
+We implemented Minstrel by porting the C code from the 3.3.8 version of the Linux kernel into Python.
+\texttt{minstrel.py} contains \texttt{ApplyRate()}, \texttt{ProcessFeedback()} and  \texttt{UpdateStats()}, as well as a few helper functions and data structures for tracking rate statistics.
 
@@
 \subsection{Availability}
-We should make our github repo or something like that available here.
+All of our code and collected trace data are available on GitHub:
+
+\noindent
+\url{https://github.com/pavpanchekha/6.829-project/tree/3.8.6}
+
+This repository contains the modified ath9k driver used to collect the traces, the traces we collected, as well as the Python simulation framework.
 
@@
 
-802.11n, SampleRate with EWMA, take data using broadcast approach,  submitting minstrel changes as kernel patch?
+Very little analysis has been done of the performance of bit rate selection algorithms on 802.11n networks.
+802.11n introduces many new rates, as well as multiple-input/multiple output, or MIMO, capabilities.
+There is an 802.11n version of Minstrel, but there is no implmentation of SampleRate.
+
+We implemented SampleRate as outlined in the thesis, but the MadWifi implementation has a few key differences from the thesis.
+We suspect that the MadWifi use of an EWMA instead of a window has more accurate rate statistics, which will potentially lead to better throughput.
+We did not have the time to implement this alternative version, but it is worth further examination.
+
+Hari Balikrishnan suggested that we use a broadcast approach when collecting our traces, instead of link layer acknowledgements.
+This would entail having multiple computers listening at the broadcast address, and compiling successes and failures from that information.
+This approach would free us from certain idiosyncracies of reading data from the driver, and lead to possibly more accurate traces.
+We were unable to implement this due to a lack of time and ability to acquire compatible hardware.
+
+Finally, we are considering submitting our improvements to the Minstrel algorithm, namely the BEWMA, as a kernel patch. Currently is implemented in Python, so we would have to port the code into C.
 
@@
 \section{Acknowledgements}
-Jonathan Perry, Hari Balikrishnan, and Derek Smithies.
+We would like to thank Jonathan Perry and Hari Balikrishnan for their valuable guidance, as well as Derek Smithies for answering our questions about Minstrel.
 

### 2013-05-14 paper/paper.tex
added plots

@@
 Cross-layer protocols use SoftPHY data from the physical layer.
-The most commonly implemented protocols on today's networks are frame based, because SNR protocols perform poorly [CITATION HERE], and cross-layer protocols cannot be deployed on current networks because they violate the network layering abstration.
+The most commonly implemented protocols on today's networks are frame based, because SNR protocols perform poorly~\cite{samplerate}, and cross-layer protocols cannot be deployed on current networks because they violate the network layering abstration.
 
@@
 
-We used a trace-based approach to collect data, and then replayed that data to the rate-selection algorithms using a Python framework. Th
+We used a trace-based approach to collect data, and then replayed that data to the rate-selection algorithms using a Python framework.
 
@@
 
-At this point, we decided to take a different approach and use an approach similar to that used in Sprout.
+At this point, we decided to take a different approach and use an approach similar to that used in Sprout~\cite{sprout}.
 
@@
 
-Yeee haw.
 \begin{figure}[htb]

### 2013-05-14 data/bar.py

@@ -1,34 +1,44 @@
 import numpy as np
 import matplotlib.pyplot as plt
 
-N = 12
-throughput = (.806, 1.632, 4.716, 5.118, 7.888, 8.784, 9.865, 15.156, 1.174, 1.960, 0, 1.189)
+N = 4
+samplerate1 = (16.474, 13.585, 5.42, 16.138, 7.455)
+minstrel1 = (12.653, 10.208, 7.587, 10.867, 8.430)
+minproved1 = (17.037, 14.879, 11.107, 15.846, 12.162)
+
+samplerate2 = (13.107, 9.688, 7.982, 13.894)
+minstrel2 = (11.575, 10.837, 8.320, 11.729)
+minproved2 =(16.869, 15.156, 12.570, 16.292)
+
+
 
 ind = np.arange(N)  # the x locations for the groups
-width = 0.5       # the width of the bars
+width = 0.25       # the width of the bars
 
 fig = plt.figure()
 ax = fig.add_subplot(111)
-rects1 = ax.bar(ind, throughput, width, color=['red', 'orange', 'DarkGoldenRod', 'yellow', 'lawngreen', 'DarkGreen', 'blue', 'DarkBlue', 'indigo',  'purple', 'DarkMagenta', 'gray'] )
+sr1 = ax.bar(ind, samplerate2, width, color='r')
+mn1 = ax.bar(ind+width, minstrel2, width, color='b')
+mp1 = ax.bar(ind+2*width, minproved2, width, color='y')
 
 # add some
-ax.set_ylim([0,17])
-ax.set_xlim([-0.5, 12])
+ax.set_ylim([0,20])
+ax.set_xlim([-0.5, 6])
 ax.set_ylabel('Throughput in Mbps')
-ax.set_title('Throughput for Constant Bitrate')
-ax.set_xticks(ind+width)
-ax.set_xticklabels( ('1', '2', '5.5', '6', '9', '11', '12', '18', '24', '36', '48', '54') )
+ax.set_xticks(ind+width+width)
+ax.set_xticklabels( ('clear', 'moving', 'corner', 'interference') )
 
-#ax.legend( (rects1[0], rects2[0]), ('Men', 'Women') )
+ax.legend( (mn1[0], sr1[0], mp1[0]), ('Minstrel', 'SampleRate', 'Minproved') )
 
 def autolabel(rects):
     # attach some text labels
     for rect in rects:
         height = rect.get_height()
[…]

### 2013-05-15 paper/paper.tex
save

@@
 } % end author
+
+\date{May 2013}
 
@@
 \begin{table*}[htb]
+  \centering
     \begin{tabular}[htb]{l|l|l|l}
@@
 
+TODO
+
 \begin{figure}[hbt]
@@
 
-\begin{figure}[htb]
-  \hspace{-.75em}\includegraphics[width=3.5in]{constant.png}\vspace{-0em}
-  \caption{Throughput of using a constant 802.11b/g bitrate in an MIT basement.}
-\end{figure}
-
 \subsection{Testing framework}
 
-Harness.py, etc.
+TODO: Harness.py, etc.
 
@@
 
-Minstrel actually is better than SampleRate (except on short traces). Eit.
+Minstrel and SampleRate perform similarly.
+Minstrel sometimes has very poor performance on short traces--as low as 2mbps.
+This is because it does not stop sending after four successive failures, but must wait until the end of the 100ms window for statistics to be compiled.
+
+\begin{figure}[htb]
+  \hspace{-1em}\includegraphics[width=3.4in]{constant.png}\vspace{-0em}
+  \caption{Throughput of using a constant 802.11b/g bitrate in clear line-of-sight from the access point.}
+\end{figure}
+
+\begin{figure}[htb]
+  \hspace{-.75em}\includegraphics[width=3.5in]{MinVSSam1.png}\vspace{-0em}
+  \caption{Throughput of Minstrel and SampleRate on various trace situations.}
+\end{figure}
 
@@
 Minproved.
+
+\begin{figure}[htb]
+  \hspace{-2em}\includegraphics[width=3.5in]{mnVSspVSmp1.png}\vspace{-0em}
+  \caption{Throughput of Minproved as compared to SampleRate and vanilla Minstrel.}
+\end{figure}
 
@@
 
-The end.
+Minstrel and SampleRate perform similarly.
+The results suggest that SampleRate tends to do slightly better, except in very lossy links such as the client being far awat from the AP or actively moving.
+Minproved, our improved version of Minstrel, solidly outperforms both Minstrel and SampleRate.
+
+It seems like the abandonment of SampleRate in the Linux kernel may have been a bit hasty, since it seems to perform similarly to Minstrel and it is a much simpler algorithm and thus less bug prone.
+We are not sure if there are adjustments to SampleRate that will bring it's performance near Minproved, though.
 

### 2013-05-15 paper/paper.tex
added to paper

@@
     \begin{tabular}[htb]{l|l|l|l}
-    Try & Normal               & Random (slower than best) & Random (faster than best) \\ \hline
+    \textbf{Try} & \textbf{Normal}               & \textbf{Random (slower than best)} & \textbf{Random (faster than best)} \\ \hline
     1   & Best throughput      & Best throughput           & Random                    \\
@@
   \caption{We collected most traces at a remote access point in the basement of Building 13 at MIT.}
+\label{figure:1}
 \end{figure}
@@
 
-Minstrel and SampleRate perform similarly.
-Minstrel sometimes has very poor performance on short traces--as low as 2mbps.
+Minstrel and SampleRate perform similarly, usually within 3 Mbps of each other.
+In especially noisy situations, such as when the client is moving or when the client is around the corner from the access point, Minstrel consistently outperforms SampleRate.
+Figure~\ref{figure:3} shows the results of a single run of both Minstrel and SampleRate over all the different types of traces we collected.
+This is consisten with the literature--the authors of Minstrel noted that they focused on making Minstrel robust to poor conditions~\cite{minstrel}, which means losing potential througput gains in stable situations.
+Conversely, John Bicket noted that SampleRate performs better in stable situations than noisy or highly varying conditions~\cite{samplerate}.
+We were slightly surprised to see how significantly SampleRate outperformed Minstrel in certain situations, though, with throughput being up to 67\% higher.
+
+In Figure~\ref{figure:2} we show the performance of constant bitrates on the \texttt{clear\_1.dat} trace, which is a 35 second trace in clear view of the access point.
+The optimal rate here was 18 Mbps, which achieved an average throughput of about 15 Mbps.
+The next highest bitrate, 24 Mbps, shows a large plummet in performance, down to a throughput of 1.17 Mbps.
+This steep drop is charateristic, and is also talked about in the SampleRate thesis~\cite{samplerate}.
+It demonstrates just how important bit rate selection is--the optimal bit rate has typically much higher throughput than the next-best rate, but if the algorithm strays too high the results can be devistating.
+
+Bit rate selection algorithms strive to have a throughput near or greater than the optimal constant rate.
+For this \texttt{clear\_1.dat} trace, SampleRate clearly meets this goal, achieving an average throughput more than 1 Mbps higher than the optimal constant rate.
+In noisier situations, though, Minstrel tends to be closer to the optimal constant rate.
+
+We did notice that Minstrel occasionally has very poor performance on traces 1 seconds or shorter--sometimes as low as 2mbps.
 This is because it does not stop sending after four successive failures, but must wait until the end of the 100ms window for statistics to be compiled.
+Thus it wastes a lot of time sending at non-optimal rates, and a short trace only has a few windows.
+This behavior does not show up in long traces at all, however, as Minstrel has had time to both smooth out bad windows and collect more statistics (making it less likely to choose a bad rate).
+
+Bit rate selection algorithms naturally show variation between runs on the same data because probe rates are chosen randomly.
+The variations were small, though.
+On ten runs on the \texttt{clear\_1.dat} trace, the average of Minstrel over 10 runs was 7.52 Mbps, with a maximum variation of +/- 0.12 Mbps.
+SampleRate was similarly stable, it averaged 5.284 Mbps over 10 runs, with a maximum variation of about +/- 0.3 Mbps
+
+TODO: talk about histograms (table 2)
+
+Some interesting patterns we noticed were that Minstrel had much fewer failures than SampleRate, usually between 0 and 10 for each run.
+SampleRate usually had one to two thousand failures on 30s.
+Minstrel's comparative lack of failures is due to the multi-rate retry chain.
+TODO: finish thought, talk about EWMA weighting problem
 
@@
   \caption{Throughput of using a constant 802.11b/g bitrate in clear line-of-sight from the access point.}
+\label{figure:2}
 \end{figure}
@@
   \caption{Throughput of Minstrel and SampleRate on various trace situations.}
+\label{figure:3}
 \end{figure}
@@
 
+\begin{table*}[htb]
+    \centering
+    \begin{tabular}{l|l|l|l}
+    \textbf{Rate in Mbps}     & \textbf{SampleRate} & \textbf{Minstrel}   & \textbf{Minproved}   \\ \hline
+    1                & 770        & 39         & 29          \\
+    2                & 298        & 82         & 20          \\
+    5.5              & 5114       & 110        & 805         \\
+    6                & 1412       & 137        & 276         \\
+    9                & 12         & 327        & 57          \\
+    11               & 3480       & 4121       & 3356        \\
+    12               & 4616       & 9515       & 14285       \\
+    18               & 16         & 13295      & 15795       \\
+    24               & 16         & 6909       & 1918        \\
+    36               & 16         & 6944       & 2030        \\
+    48               & 16         & 6888       & 2009        \\
+    54               & 16         & 7308       & 1953        \\ \hline
+    \textbf{Avg. Throughput:} & \emph{5.42 Mbps}  & \emph{7.578 Mbps} & \emph{11.107 Mbps} \\
+    \end{tabular}
+    \caption{Histograms of the number of packets sent at each rate.
+This trace was recorded around the corner from the access point.
+The best fixed rate for this trace was 11 Mbps, which achieved a throughput of 9.625 Mbps.
+Minproved achieves a higher throughput than the best fixed rate.}
+\label{table:2}
+\end{table*}
+
 \begin{figure}[htb]
@@
   \caption{Throughput of Minproved as compared to SampleRate and vanilla Minstrel.}
+\label{figure:4}
 \end{figure}

### 2013-05-16 paper/paper.tex
paper at 5pgs

@@
 On ten runs on the \texttt{clear\_1.dat} trace, the average of Minstrel over 10 runs was 7.52 Mbps, with a maximum variation of +/- 0.12 Mbps.
-SampleRate was similarly stable, it averaged 5.284 Mbps over 10 runs, with a maximum variation of about +/- 0.3 Mbps
+SampleRate was similarly stable, it averaged 5.284 Mbps over 10 runs, with a maximum variation of about +/- 0.3 Mbps.
+The variation of results on other traces was similar.
 
-TODO: talk about histograms (table 2)
+The histograms of which rates the algorithms deided to send at provided some useful insight into the behavior.
+Table~\ref{table:2} is a histogram of a 35 second corner trace.
+In this situation Minstrel outperforms SampleRate.
+The constant rate throughputs demonstrate that 11, 12 and 18 Mbps were the optimal rates to send at, and that no packets were sucessful at any higher rates.
+SampleRate does not use these rates except for probe packets, which happens once every ten seconds.
+Minstrel does not have this measure, however, spends a lot of time sending probe packets at these high rates that happen to never be successful.
+The only reason low rates have fewer probe packets is because they are second in the retry chain.
+Despite the fact that SampleRate spends less time probing high rates, it is less successful.
+This is because Minstrel send the most packets at the three most successful bitrates.
+SampleRate, however, spends too much time sampling low bit rates.
+Additionally, it spends a lot of time sending at 5.5 Mbps.
+TODO: why does SampleRate spend time at 5.5 Mbps???
+We suspect that this might be because the equation SampleRate uses to estimate average transmission time needs to be tweaked.
 
-Some interesting patterns we noticed were that Minstrel had much fewer failures than SampleRate, usually between 0 and 10 for each run.
-SampleRate usually had one to two thousand failures on 30s.
-Minstrel's comparative lack of failures is due to the multi-rate retry chain.
-TODO: finish thought, talk about EWMA weighting problem
+Our examination of the histograms showed us that Minstrel spends a lot of time sampling at high rates.
+In the specific case we examined, they were all probe packets.
+But sometimes Minstrel would choose 54Mbps as the rate with the best throughput, which made little sense.
+A closer investigation revealed that Minstrel would sometimes luck out and get one success sending a probe at 54Mbps.
+The probability of success would be 1, so then 54Mbps would be chosen as the rate with the best throughput.
+This is unfortunate, but the real harm came in the next window: nearly all of the packets at 54Mbps would fail, but Minstrel was still considering 54 Mbps to be the best rate!
+This is because the EWMA did not account for the fact that the second block having sent hunreds more packets at 54 Mbps than the first.
+Thus the probability of success would drop from 1 to about 0.70, which is still high enough for the throughput equation to rank 54 Mbps highly.
+In essense, the probabilities were not declining fast enough after a large number of failures at some rate.
+
+In the next section we will talk about Minproved, an improved version of Minstrel where we attempt to fix the probability problem.
 
@@
 \begin{figure}[htb]
-  \hspace{-.75em}\includegraphics[width=3.5in]{MinVSSam1.png}\vspace{-0em}
+  \hspace{-.75em}\includegraphics[width=3.5in]{MinVSSam1.png}\vspace{-0.75em}
   \caption{Throughput of Minstrel and SampleRate on various trace situations.}
@@
 \subsection{Improvements to Minstrel}
-Minproved.
-
+TODO: Minproved.
+-make sure we mention how Minproved almost always does better than highest fixed rate
 \begin{table*}[htb]
     \centering
-    \begin{tabular}{l|l|l|l}
-    \textbf{Rate in Mbps}     & \textbf{SampleRate} & \textbf{Minstrel}   & \textbf{Minproved}   \\ \hline
-    1                & 770        & 39         & 29          \\
-    2                & 298        & 82         & 20          \\
-    5.5              & 5114       & 110        & 805         \\
-    6                & 1412       & 137        & 276         \\
-    9                & 12         & 327        & 57          \\
-    11               & 3480       & 4121       & 3356        \\
-    12               & 4616       & 9515       & 14285       \\
-    18               & 16         & 13295      & 15795       \\
-    24               & 16         & 6909       & 1918        \\
-    36               & 16         & 6944       & 2030        \\
-    48               & 16         & 6888       & 2009        \\
-    54               & 16         & 7308       & 1953        \\ \hline
+    \begin{tabular}{l|l|l|l|l}
+    \textbf{Rate in Mbps}     & \textbf{SampleRate} & \textbf{Minstrel}   & \textbf{Minproved}   & \textbf{Constant Rate Throughput}\\ \hline
+    1                & 631        & 39         & 29      & 0.639 Mbps    \\
+    2                & 863        & 82         & 20      & 1.425 Mbps    \\
+    5.5              & 4814       & 110        & 805     & 4.406 Mbps    \\
+    6                & 423       & 137        & 276     & 4.603 Mbps    \\
+    9                & 553         & 327        & 57      & 4.630 Mbps   \\
+    11               & 3326       & 4121       & 3356    & 9.627 Mbps   \\
+    12               & 4649       & 9515       & 14285   & 9.444 Mbps   \\
+    18               & 1154         & 13295      & 15795   & 8.458 Mbps    \\
+    24               & 16         & 6909       & 1918    & 0 Mbps   \\
+    36               & 16         & 6944       & 2030    & 0 Mbps   \\
+    48               & 16         & 6888       & 2009    & 0 Mbps   \\
+    54               & 16         & 7308       & 1953    & 0 Mbps   \\ \hline
     \textbf{Avg. Throughput:} & \emph{5.42 Mbps}  & \emph{7.578 Mbps} & \emph{11.107 Mbps} \\
@@
     \caption{Histograms of the number of packets sent at each rate.
-This trace was recorded around the corner from the access point.
-The best fixed rate for this trace was 11 Mbps, which achieved a throughput of 9.625 Mbps.
+The trace was \texttt{corner\_1.dat} and was recorded around the corner from the access point.
+The best fixed rate for this trace was 11 Mbps, which achieved a throughput of 9.627 Mbps.
 Minproved achieves a higher throughput than the best fixed rate.}
@@
 \noindent
-\url{https://github.com/pavpanchekha/6.829-project/tree/3.8.6}
+\small{\url{https://github.com/pavpanchekha/6.829-project/tree/3.8.6}}
 

### 2013-05-16 pysim/common.py

@@ -57,7 +57,7 @@ class EWMA:
         self.time = time
 
     def read(self):
-        if self.val:
+        if self.val is not None:
             return self.val
         else: 
             return None

### 2013-05-16 pysim/minstrel.py

@@ -302,6 +302,7 @@ def update_stats(timestamp):
             br.att_hist += br.attempts
             br.ewma.feed(timestamp, p)
         p = br.ewma.read()
+        
 
         if p and (p > 17100 or p < 1800):
             br.adjusted_retry_count = br.retry_count >> 1
@@ -327,7 +328,7 @@ def update_stats(timestamp):
 
     #print("(rate, throughput, probability)")
     #for r in rates_:
-    #    print("(%r, %r mbps, p = %r, succ = %r, att = %r)"%(r.rate, round(r.throughput/1e6, 3), round(r.ewma.read()/18000.0, 3) if r.ewma.read() else None, r.succ_hist, r.att_hist)) 
+    #    print("(%r, %r mbps, p = %r, succ = %r, att = %r)"%(r.rate, round(r.throughput/1e6, 3), round(r.ewma.read()/18000.0, 3) if r.ewma.read() is not None else None, r.succ_hist, r.att_hist)) 
     #print()
 
     nextThruput = rates_[1].rate

### 2013-05-16 paper/paper.tex
paper tweak
paper tweak2

@@
 Thus the probability of success would drop from 1 to about 0.70, which is still high enough for the throughput equation to rank 54 Mbps highly.
-In essense, the probabilities were not declining fast enough after a large number of failures at some rate.
+In essense, the probabilities were not declining fast enough after a large number of failures.
 
@@
 
-Hari Balikrishnan suggested that we use a broadcast approach when collecting our traces, instead of link layer acknowledgements.
+Hari Balakrishnan suggested that we use a broadcast approach when collecting our traces, instead of link layer acknowledgements.
 This would entail having multiple computers listening at the broadcast address, and compiling successes and failures from that information.
@@
 \section{Acknowledgements}
-We would like to thank Jonathan Perry and Hari Balikrishnan for their valuable guidance, as well as Derek Smithies for answering our questions about Minstrel.
+We would like to thank Jonathan Perry and Hari Balakrishnan for their valuable guidance, as well as Derek Smithies for answering our questions about Minstrel.
 

### 2013-05-16 paper/paper.tex
paper updates

@@
 
+\subsection{Experimental parameters}
+
+There are a few parameters that can be adjusted in Minstrel and SampleRate, but our implementation is consistent with what the respective authors originally chose.
+
+In SampleRate, the number of successive retries, the frequency of probe packets, the window size, and the lossless $tx\_time$ estimations are all parameters that can be changed.
+The SampleRate thesis chose 4 successive retries, a probe frequency of every ten packets, and a window size of ten seconds.
+We used the same values in our implementation.
+Additionally, we use the same equation to calculate $tx\_time$i.
+
+In Minstrel, the frequency of probe packets, the window size, and the EWMA weighting are the adjustable parameters.
+The 3.3.8 Linux kernel uses a probe frequency of 10\%, a window size of 100ms, and a EWMA weighing of 0.75.
+We use the same values.
+Additionally, Minstrel also estimates $tx\_time$ using the \texttt{ieee80211\_frame\_duration()} method from the \texttt{util.c} file in \texttt{net/mac80211}.
+We implemented the same function in Python.
+
 \section{Analysis}
@@
 Figure~\ref{figure:3} shows the results of a single run of both Minstrel and SampleRate over all the different types of traces we collected.
-This is consisten with the literature--the authors of Minstrel noted that they focused on making Minstrel robust to poor conditions~\cite{minstrel}, which means losing potential througput gains in stable situations.
+This is consisten with the literature---the authors of Minstrel noted that they focused on making Minstrel robust to poor conditions~\cite{minstrel}, which means losing potential througput gains in stable situations.
 Conversely, John Bicket noted that SampleRate performs better in stable situations than noisy or highly varying conditions~\cite{samplerate}.
@@
 This steep drop is charateristic, and is also talked about in the SampleRate thesis~\cite{samplerate}.
-It demonstrates just how important bit rate selection is--the optimal bit rate has typically much higher throughput than the next-best rate, but if the algorithm strays too high the results can be devistating.
+It demonstrates just how important bit rate selection is---the optimal bit rate has typically much higher throughput than the next-best rate, but if the algorithm strays too high the results can be devistating.
 
@@
 
-We did notice that Minstrel occasionally has very poor performance on traces 1 seconds or shorter--sometimes as low as 2mbps.
+We did notice that Minstrel occasionally has very poor performance on traces 1 seconds or shorter---sometimes as low as 2 Mbps.
 This is because it does not stop sending after four successive failures, but must wait until the end of the 100ms window for statistics to be compiled.
@@
 \noindent
-\small{\url{https://github.com/pavpanchekha/6.829-project/tree/3.8.6}}
+{\small\url{https://github.com/pavpanchekha/6.829-project/tree/3.8.6}}
 

### 2013-05-16 paper/paper.tex
spell checked

@@
 \begin{document}
-
 %don't want date printed
@@
 
-There are three main classes of bit rate adaptiation protocols: frame-based, SNR-based, and cross-layer protocols.
-Frame-based protocols measure the fraction of sucessfully received protocols.
+There are three main classes of bit rate adaptation protocols: frame-based, SNR-based, and cross-layer protocols.
+Frame-based protocols measure the fraction of successfully received packets.
 SNR protocols make decisions based on the estimated Signal to Noise ratio.
 Cross-layer protocols use SoftPHY data from the physical layer.
-The most commonly implemented protocols on today's networks are frame based, because SNR protocols perform poorly~\cite{samplerate}, and cross-layer protocols cannot be deployed on current networks because they violate the network layering abstration.
+The most commonly implemented protocols on today's networks are frame based, because SNR protocols perform poorly~\cite{samplerate}, and cross-layer protocols cannot be deployed on current networks because they violate the network layering abstraction.
 
-The two most popular fram-based protocols are SampleRate and Minstrel.
-Both were implemented in the MadWifi drivers for Linux wirless, and today Minstrel is the default bit-rate selection algorithm for all wireless drivers on Linux.
+The two most popular frame-based protocols are SampleRate and Minstrel.
+Both were implemented in the MadWifi drivers for Linux wireless, and today Minstrel is the default bit-rate selection algorithm for all wireless drivers on Linux.
 Our research used a trace-based approach to analyze the performance of these two algorithms.
@@
 
-SampleRate has three main functions: \texttt{ApplyRate()}, which returns a the bit-rate to send a packet at, \texttt{ProcessFeedback()}, which updates the statistics for a bitrate after an attempted packet send,  and \texttt{RemoveStaleResults()}, which removes all statistics older than 10 seconds.
+SampleRate has three main functions: \texttt{ApplyRate()}, which returns a the bit-rate to send a packet at, \texttt{ProcessFeedback()}, which updates the statistics for a bit rate after an attempted packet send,  and \texttt{RemoveStaleResults()}, which removes all statistics older than 10 seconds.
 
@@
 This ensures that Minstrel never samples rates worse than the current optimal rate unless the optimal rate experiences a failure.
-See table ~\ref{table:1} for an overview of the mrr's used in Minsterl.
+See table ~\ref{table:1} for an overview of the mrr's used in Minstrel.
 
@@
 The weighting for old data decreases exponentially.
-An EWMA is beneficial because it ensures that a sudden degredation in link quality will create a rapid response in the probabilities due to the increased weight of recent data.
+An EWMA is beneficial because it ensures that a sudden degradation in link quality will create a rapid response in the probabilities due to the increased weight of recent data.
 Minstrel will waste less time sending packets at a rate that no longer works, as compared to a simple probability calculation.
@@
 Our initial plan was to use the algorithms as implemented in the MadWifi drivers for the Atheros chipset.
-However, MadWifi is depricated, and only runs on quite old network cards.
+However, MadWifi is deprecated, and only runs on quite old network cards.
 In addition to the lack of compatible hardware available to us, MadWifi is not an accurate reflection of what a modern user would use.
 We looked into the ath5k and ath9k projects, which were created to replace MadWifi.
-Both of these did not port the old bit-rate selection system from MadWifi and instead opted to use the linux kernel's rate selection software, which only offers Minstrel.
+Both of these did not port the old bit-rate selection system from MadWifi and instead opted to use the Linux kernel's rate selection software, which only offers Minstrel.
 Porting SampleRate over to the new drivers would not have been trivial, as there are vast differences in the assumed interfaces.
@@
 This implementation is slightly different from how it was implemented in the MadWifi kernel driver.
-The primary difference is that the kernel implementatiion uses a EWMA instead of a 10 second window.
-The thesis implmenentation throws out old rate data and begins anew every 10 seconds.
+The primary difference is that the kernel implementation uses a EWMA instead of a 10 second window.
+The thesis implementation throws out old rate data and begins anew every 10 seconds.
 It would be worth looking at the performance of EWMA-based SampleRate in the future.
@@
 Figure~\ref{figure:3} shows the results of a single run of both Minstrel and SampleRate over all the different types of traces we collected.
-This is consisten with the literature---the authors of Minstrel noted that they focused on making Minstrel robust to poor conditions~\cite{minstrel}, which means losing potential througput gains in stable situations.
+This is consistent with the literature---the authors of Minstrel noted that they focused on making Minstrel robust to poor conditions~\cite{minstrel}, which means losing potential throughput gains in stable situations.
 Conversely, John Bicket noted that SampleRate performs better in stable situations than noisy or highly varying conditions~\cite{samplerate}.
@@
 
-In Figure~\ref{figure:2} we show the performance of constant bitrates on the \texttt{clear\_1.dat} trace, which is a 35 second trace in clear view of the access point.
+In Figure~\ref{figure:2} we show the performance of constant bit rates on the \texttt{clear\_1.dat} trace, which is a 35 second trace in clear view of the access point.
 The optimal rate here was 18 Mbps, which achieved an average throughput of about 15 Mbps.
 The next highest bitrate, 24 Mbps, shows a large plummet in performance, down to a throughput of 1.17 Mbps.
-This steep drop is charateristic, and is also talked about in the SampleRate thesis~\cite{samplerate}.
-It demonstrates just how important bit rate selection is---the optimal bit rate has typically much higher throughput than the next-best rate, but if the algorithm strays too high the results can be devistating.
+This steep drop is characteristic, and is also talked about in the SampleRate thesis~\cite{samplerate}.
+It demonstrates just how important bit rate selection is---the optimal bit rate has typically much higher throughput than the next-best rate, but if the algorithm strays too high the results can be devastating.
 
@@
 
-The histograms of which rates the algorithms deided to send at provided some useful insight into the behavior.
+The histograms of which rates the algorithms decided to send at provided some useful insight into the behavior.
 Table~\ref{table:2} is a histogram of a 35 second corner trace.
 In this situation Minstrel outperforms SampleRate.
-The constant rate throughputs demonstrate that 11, 12 and 18 Mbps were the optimal rates to send at, and that no packets were sucessful at any higher rates.
+The constant rate throughputs demonstrate that 11, 12 and 18 Mbps were the optimal rates to send at, and that no packets were successful at any higher rates.
 SampleRate does not use these rates except for probe packets, which happens once every ten seconds.
@@
 In the specific case we examined, they were all probe packets.
-But sometimes Minstrel would choose 54Mbps as the rate with the best throughput, which made little sense.
-A closer investigation revealed that Minstrel would sometimes luck out and get one success sending a probe at 54Mbps.
-The probability of success would be 1, so then 54Mbps would be chosen as the rate with the best throughput.
-This is unfortunate, but the real harm came in the next window: nearly all of the packets at 54Mbps would fail, but Minstrel was still considering 54 Mbps to be the best rate!
-This is because the EWMA did not account for the fact that the second block having sent hunreds more packets at 54 Mbps than the first.
+But sometimes Minstrel would choose 54 Mbps as the rate with the best throughput, which made little sense.
+A closer investigation revealed that Minstrel would sometimes luck out and get one success sending a probe at 54 Mbps.
+The probability of success would be 1, so then 54 Mbps would be chosen as the rate with the best throughput.
+This is unfortunate, but the real harm came in the next window: nearly all of the packets at 54 Mbps would fail, but Minstrel was still considering 54 Mbps to be the best rate!
+This is because the EWMA did not account for the fact that the second block having sent hundreds more packets at 54 Mbps than the first.
 Thus the probability of success would drop from 1 to about 0.70, which is still high enough for the throughput equation to rank 54 Mbps highly.
-In essense, the probabilities were not declining fast enough after a large number of failures.
+In essence, the probabilities were not declining fast enough after a large number of failures.
 
@@
 802.11n introduces many new rates, as well as multiple-input/multiple output, or MIMO, capabilities.
-There is an 802.11n version of Minstrel, but there is no implmentation of SampleRate.
+There is an 802.11n version of Minstrel, but there is no implementation of SampleRate.
 
@@
 
-Hari Balakrishnan suggested that we use a broadcast approach when collecting our traces, instead of link layer acknowledgements.
+Hari Balakrishnan suggested that we use a broadcast approach when collecting our traces, instead of link layer acknowledgments.
 This would entail having multiple computers listening at the broadcast address, and compiling successes and failures from that information.
-This approach would free us from certain idiosyncracies of reading data from the driver, and lead to possibly more accurate traces.
+This approach would free us from certain idiosyncrasies of reading data from the driver, and lead to possibly more accurate traces.
 We were unable to implement this due to a lack of time and ability to acquire compatible hardware.
@@
 Minstrel and SampleRate perform similarly.
-The results suggest that SampleRate tends to do slightly better, except in very lossy links such as the client being far awat from the AP or actively moving.
+The results suggest that SampleRate tends to do slightly better, except in very lossy links such as the client being far away from the AP or actively moving.
 Minproved, our improved version of Minstrel, solidly outperforms both Minstrel and SampleRate.
@@
 
-\section{Acknowledgements}
+\section{Acknowledgments}
 We would like to thank Jonathan Perry and Hari Balakrishnan for their valuable guidance, as well as Derek Smithies for answering our questions about Minstrel.

### 2013-05-16 paper/paper.tex
standardized to bit-rate

@@
 \subsection*{Abstract}
-In this paper we compare the performance SampleRate and Minstrel, two popular bit rate selection algorithms that have widespread real-world usage on the Linux operating system.
+In this paper we compare the performance SampleRate and Minstrel, two popular bit-rate selection algorithms that have widespread real-world usage on the Linux operating system.
 We use a novel trace-based approach that allows us avoid in-depth kernel programming and write simple Python scripts instead.
@@
 An optimal rate at some time $t$ may be different from the optimal rate just 30 seconds earlier.
-A good bit rate selection algorithm has to detect and adapt to these conditions.
-If the chosen bit rate is too slow, then the throughput will be unnecessarily low.
+A good bit-rate selection algorithm has to detect and adapt to these conditions.
+If the chosen bit-rate is too slow, then the throughput will be unnecessarily low.
 If the rate is too high, then failures will be very frequent and throughput will again suffer.
 
-There are three main classes of bit rate adaptation protocols: frame-based, SNR-based, and cross-layer protocols.
+There are three main classes of bit-rate adaptation protocols: frame-based, SNR-based, and cross-layer protocols.
 Frame-based protocols measure the fraction of successfully received packets.
@@
 Cross-layer protocols use SoftPHY data from the physical layer.
-The most commonly implemented protocols on today's networks are frame based, because SNR protocols perform poorly~\cite{samplerate}, and cross-layer protocols cannot be deployed on current networks because they violate the network layering abstraction.
+The most commonly implemented protocols on today's networks are frame based, because SNR protocols perform poorly~\cite{samplerate}, and cross-layer protocols cannot be deployed on current networks as they violate the network layering abstraction.
 
@@
 
-SampleRate has three main functions: \texttt{ApplyRate()}, which returns a the bit-rate to send a packet at, \texttt{ProcessFeedback()}, which updates the statistics for a bit rate after an attempted packet send,  and \texttt{RemoveStaleResults()}, which removes all statistics older than 10 seconds.
+SampleRate has three main functions: \texttt{ApplyRate()}, which returns a the bit-rate to send a packet at, \texttt{ProcessFeedback()}, which updates the statistics for a bit-rate after an attempted packet send,  and \texttt{RemoveStaleResults()}, which removes all statistics older than 10 seconds.
 
@@
 
-In Figure~\ref{figure:2} we show the performance of constant bit rates on the \texttt{clear\_1.dat} trace, which is a 35 second trace in clear view of the access point.
+In Figure~\ref{figure:2} we show the performance of constant bit-rates on the \texttt{clear\_1.dat} trace, which is a 35 second trace in clear view of the access point.
 The optimal rate here was 18 Mbps, which achieved an average throughput of about 15 Mbps.
-The next highest bitrate, 24 Mbps, shows a large plummet in performance, down to a throughput of 1.17 Mbps.
+The next highest bit-rate, 24 Mbps, shows a large plummet in performance, down to a throughput of 1.17 Mbps.
 This steep drop is characteristic, and is also talked about in the SampleRate thesis~\cite{samplerate}.
-It demonstrates just how important bit rate selection is---the optimal bit rate has typically much higher throughput than the next-best rate, but if the algorithm strays too high the results can be devastating.
+It demonstrates just how important bit-rate selection is---the optimal bit-rate has typically much higher throughput than the next-best rate, but if the algorithm strays too high the results can be devastating.
 
-Bit rate selection algorithms strive to have a throughput near or greater than the optimal constant rate.
+bit-rate selection algorithms strive to have a throughput near or greater than the optimal constant rate.
 For this \texttt{clear\_1.dat} trace, SampleRate clearly meets this goal, achieving an average throughput more than 1 Mbps higher than the optimal constant rate.
@@
 
-Bit rate selection algorithms naturally show variation between runs on the same data because probe rates are chosen randomly.
+bit-rate selection algorithms naturally show variation between runs on the same data because probe rates are chosen randomly.
 The variations were small, though.
@@
 Despite the fact that SampleRate spends less time probing high rates, it is less successful.
-This is because Minstrel send the most packets at the three most successful bitrates.
-SampleRate, however, spends too much time sampling low bit rates.
+This is because Minstrel send the most packets at the three most successful bit-rates.
+SampleRate, however, spends too much time sampling low bit-rates.
 Additionally, it spends a lot of time sending at 5.5 Mbps.
@@
   \hspace{-1em}\includegraphics[width=3.4in]{constant.png}\vspace{-0em}
-  \caption{Throughput of using a constant 802.11b/g bitrate in clear line-of-sight from the access point.}
+  \caption{Throughput of using a constant 802.11b/g bit-rate in clear line-of-sight from the access point.}
 \label{figure:2}
@@
 
-Very little analysis has been done of the performance of bit rate selection algorithms on 802.11n networks.
+Very little analysis has been done of the performance of bit-rate selection algorithms on 802.11n networks.
 802.11n introduces many new rates, as well as multiple-input/multiple output, or MIMO, capabilities.

### 2013-05-16 paper/paper.tex
Added trace collection section

@@
 
-We used a trace-based approach to collect data, and then replayed that data to the rate-selection algorithms using a Python framework.
+We collected traces about the success rates of each bit rate through a modified Linux wireless driver; Minstrel and SampleRate were then re-implemented in Python and replayed on this collected data.
 
@@
 Our initial plan was to use the algorithms as implemented in the MadWifi drivers for the Atheros chipset.
-However, MadWifi is deprecated, and only runs on quite old network cards.
-In addition to the lack of compatible hardware available to us, MadWifi is not an accurate reflection of what a modern user would use.
-We looked into the ath5k and ath9k projects, which were created to replace MadWifi.
-Both of these did not port the old bit-rate selection system from MadWifi and instead opted to use the Linux kernel's rate selection software, which only offers Minstrel.
+However, MadWifi is deprecated, and only runs on now-ancient network cards.
+In addition to the unavailability of compatible hardware, MadWifi would not be an accurate reflection of contemporary users.
+Both ath5k and ath9k, which were created to replace MadWifi, did not port the old bit-rate selection system from MadWifi and instead opted to use the Linux kernel's rate selection framework, which only implements Minstrel.
 Porting SampleRate over to the new drivers would not have been trivial, as there are vast differences in the assumed interfaces.
@@
 At this point, we decided to take a different approach and use an approach similar to that used in Sprout~\cite{sprout}.
+We modified the ath9k driver to try bit rates uniformly and report packet successes and failures, then analyze this data in user-space with re-implementations of SampleRate and Minstrel.
+Not only did this lessen the re-implementation cost (since the algorithms were re-implemented in user-space in Python, not in kernel-mode in C), but it also provided reproducibility and allowed us to investigate modifications to SampleRate and to Minstrel.
 
@@
 
-TODO
+Traces were collected from an ath9k driver modified to sample bit-rates uniformly (forgoing the usual bit rate selection algorithm) and to report successes and failure from each packet.
+Each bit-rate was sampled uniformly in time -- thus, more packets were sent at higher bit-rates, since those bit-rates otherwise deliver packets faster.
+For each packet sent, the time to send was recorded, and the bit rate for the next packet was chosen to be the bit-rate that had the smallest total transmissions time.
+This guarantees approximately uniform selection of bit-rates.
+The wireless card used was an Qualcomm Atheros \texttt{AR9285} Wireless Network Adapter.
+
+Traces were collected at an unused access point in a basement of MIT.
+Before initiating trace collection, the wireless card was put in monitor mode for a minute to check that no other users were connected to the access point.
+None were, guaranteeing that traces would not include delays from carrier sensing or run into problems due to congestion at the access point.
+Traffic was simulated by sending 1500 byte packets to the access point's IP address.
+Packets were sent over UDP to avoid the TCP congestion window and retry loop; since no users were using the access point, there was no real risk of deteriorating service for other users.
+Figure~\ref{figure:1} shows a floor plan of the access point and basement.
+
+Eighteen total traces were collected, grouped into a number of scenarios.
+Each trace was 30 seconds to a minute long, and each scenario was repeated twice.
+One scenario placed the wireless card with line of sight to the access point, two meters from it.
+Another created solid body interference by having one of the authors stand between the wireless device and the access point.
+One placed the device around a corner from the access point, creating some potential for multi-path interference and fading.
+In one scenario, we moved the wireless device throughout the trace, pacing back and forth in a 10 meter line in front of the access point.
+Finally, one five-minute trace was taken, which mixed the above scenarios, first moving around a corner, then standing still, then moving to line-of-sight with the access point.
+A few shorter (3 to 15 second long) traces were collected to investigate the startup behavior of bit-rate selection.
 
 \begin{figure}[hbt]
-  \hspace{-0.5em}\includegraphics[width=3.5in]{bsmt13floorplan.png}\vspace{-0em}
+  \centering
+  \includegraphics[width=3.5in]{bsmt13floorplan.png}
   \caption{We collected most traces at a remote access point in the basement of Building 13 at MIT.}
@@
 
-TODO: Harness.py, etc.
+Once collected, traces were fed to a bit rate selection simulator to analyze the performance of Minstrel and SampleRate.
+Both algorithms were represented by two functions: \texttt{ApplyRate}, which returned a multi-rate retry chain of rates to attempt transmission at, and \texttt{ProcessFeedback}, which was passed the success or failure of the transmission and how many attempts were made at each bit rate (along with the current time and the total time to send the packet); this parallels the interface used by the Linux kernel itself.
+Helper functions, like SampleRate's \texttt{RemoveStaleResults} function, were instead called from one of \verb|apply_rate| or \verb|process_feedback|.
+For every packet transmission, the simulator would call \texttt{ApplyRate}; compute the probability of success based on packets sent in 10-millisecond window around the current time; and from that determine the success or failure of the packet, and the number of retries required.
+The results of a test run are thus non-deterministic; both algorithms tested are non-deterministic anyway, so this is no handicap.
+To improve re-implementation fidelity, both implementations were meticulously checked by both authors.
 
@@
 This implementation is slightly different from how it was implemented in the MadWifi kernel driver.
-The primary difference is that the kernel implementation uses a EWMA instead of a 10 second window.
-The thesis implementation throws out old rate data and begins anew every 10 seconds.
+The primary difference is that the kernel implementatiion uses a EWMA instead of a 10 second window.
+The thesis implementation computes success probabilities based on a 10-second sliding window.
 It would be worth looking at the performance of EWMA-based SampleRate in the future.
-
-\texttt{samplerate.py} contains \texttt{ApplyRate()}, \texttt{ProcessFeedback()} and  \texttt{RemoveStaleResults()}, as well as a few helper functions and data structures for tracking rate statistics.
+Our implementation, \texttt{samplerate.py} contains \texttt{ApplyRate()}, \texttt{ProcessFeedback()} and  \texttt{RemoveStaleResults()}, as well as a few helper functions and data structures for tracking rate statistics.
 
@@
 We used the same values in our implementation.
-Additionally, we use the same equation to calculate $tx\_time$i.
+Additionally, we use the same equation to calculate $tx\_time$.
 
 In Minstrel, the frequency of probe packets, the window size, and the EWMA weighting are the adjustable parameters.
-The 3.3.8 Linux kernel uses a probe frequency of 10\%, a window size of 100ms, and a EWMA weighing of 0.75.
+The 3.3.8 Linux kernel uses a probe frequency of 10\%, a window size of 100ms, and a EWMA weighting of 0.75.
 We use the same values.
@@
 Conversely, John Bicket noted that SampleRate performs better in stable situations than noisy or highly varying conditions~\cite{samplerate}.
-We were slightly surprised to see how significantly SampleRate outperformed Minstrel in certain situations, though, with throughput being up to 67\% higher.
+We were slightly surprised to see how significantly SampleRate outperformed Minstrel in certain situations, sometimes with throughput as much as 67\% higher.
 
 In Figure~\ref{figure:2} we show the performance of constant bit-rates on the \texttt{clear\_1.dat} trace, which is a 35 second trace in clear view of the access point.
-The optimal rate here was 18 Mbps, which achieved an average throughput of about 15 Mbps.
+The optimal constant rate here was 18 Mbps, which achieved an average throughput of about 15 Mbps.
 The next highest bit-rate, 24 Mbps, shows a large plummet in performance, down to a throughput of 1.17 Mbps.
@@
 
-bit-rate selection algorithms strive to have a throughput near or greater than the optimal constant rate.
-For this \texttt{clear\_1.dat} trace, SampleRate clearly meets this goal, achieving an average throughput more than 1 Mbps higher than the optimal constant rate.
-In noisier situations, though, Minstrel tends to be closer to the optimal constant rate.
-
 We did notice that Minstrel occasionally has very poor performance on traces 1 seconds or shorter---sometimes as low as 2 Mbps.
 This is because it does not stop sending after four successive failures, but must wait until the end of the 100ms window for statistics to be compiled.
-Thus it wastes a lot of time sending at non-optimal rates, and a short trace only has a few windows.
-This behavior does not show up in long traces at all, however, as Minstrel has had time to both smooth out bad windows and collect more statistics (making it less likely to choose a bad rate).
+Thus it wastes a lot of time sending at non-optimal rates; and a short trace has only a few 100ms windows.
+This behavior is not evident in long traces, however, as Minstrel has had time to both smooth out bad windows and collect more statistics (making it less likely to choose a bad rate).
 
-bit-rate selection algorithms naturally show variation between runs on the same data because probe rates are chosen randomly.
+Bit-rate selection algorithms naturally show variation between runs on the same data because probe rates are chosen randomly.
 The variations were small, though.
@@
 
-The histograms of which rates the algorithms decided to send at provided some useful insight into the behavior.
-Table~\ref{table:2} is a histogram of a 35 second corner trace.
-In this situation Minstrel outperforms SampleRate.
-The constant rate throughputs demonstrate that 11, 12 and 18 Mbps were the optimal rates to send at, and that no packets were successful at any higher rates.
-SampleRate does not use these rates except for probe packets, which happens once every ten seconds.
-Minstrel does not have this measure, however, spends a lot of time sending probe packets at these high rates that happen to never be successful.
-The only reason low rates have fewer probe packets is because they are second in the retry chain.
-Despite the fact that SampleRate spends less time probing high rates, it is less successful.
-This is because Minstrel send the most packets at the three most successful bit-rates.
-SampleRate, however, spends too much time sampling low bit-rates.
-Additionally, it spends a lot of time sending at 5.5 Mbps.
-TODO: why does SampleRate spend time at 5.5 Mbps???
-We suspect that this might be because the equation SampleRate uses to estimate average transmission time needs to be tweaked.
+Additionally, Minstrel uses a complex method of calculating when to send a sample packet that involves a heuristic to make sure that 10\% of packets were actually samples.
+This is because the mrr places sample packets second in the retry chain when the sample rate is slower than the rate with the best throughput.
+So even if 10\% of probes have a sample rate in the retry chain, a sample rate isn't actually sent 10\% of the time.
+We implemented this heuristic and it tended to cause Minstrel to send too many probe packets.
 
-Our examination of the histograms showed us that Minstrel spends a lot of time sampling at high rates.
-In the specific case we examined, they were all probe packets.
-But sometimes Minstrel would choose 54 Mbps as the rate with the best throughput, which made little sense.
-A closer investigation revealed that Minstrel would sometimes luck out and get one success sending a probe at 54 Mbps.
-The probability of success would be 1, so then 54 Mbps would be chosen as the rate with the best throughput.
-This is unfortunate, but the real harm came in the next window: nearly all of the packets at 54 Mbps would fail, but Minstrel was still considering 54 Mbps to be the best rate!
-This is because the EWMA did not account for the fact that the second block having sent hundreds more packets at 54 Mbps than the first.
-Thus the probability of success would drop from 1 to about 0.70, which is still high enough for the throughput equation to rank 54 Mbps highly.
-In essence, the probabilities were not declining fast enough after a large number of failures.
-
-In the next section we will talk about Minproved, an improved version of Minstrel where we attempt to fix the probability problem.
+In the next section we will talk about Minproved, an improved version of Minstrel where we attempt to fix problems with EWMA and too-frequent probing
 

### 2013-05-16 paper/paper.tex
paper updates

--- 2013-05-16 paper/paper.tex
@@
 
-At this point, we decided to take a different approach and use an approach similar to that used in Sprout~\cite{sprout}.
+At this point, we decided to take a different approach and use an trace-based method similar to that used in Sprout~\cite{sprout}.
 
@@
 
-bit-rate selection algorithms strive to have a throughput near or greater than the optimal constant rate.
+Bit-rate selection algorithms strive to have a throughput near or greater than the optimal constant rate.
 For this \texttt{clear\_1.dat} trace, SampleRate clearly meets this goal, achieving an average throughput more than 1 Mbps higher than the optimal constant rate.
@@
 
-We did notice that Minstrel occasionally has very poor performance on traces 1 seconds or shorter---sometimes as low as 2 Mbps.
-This is because it does not stop sending after four successive failures, but must wait until the end of the 100ms window for statistics to be compiled.
-Thus it wastes a lot of time sending at non-optimal rates, and a short trace only has a few windows.
-This behavior does not show up in long traces at all, however, as Minstrel has had time to both smooth out bad windows and collect more statistics (making it less likely to choose a bad rate).
-
-bit-rate selection algorithms naturally show variation between runs on the same data because probe rates are chosen randomly.
+Bit-rate selection algorithms naturally show variation between runs on the same data because probe rates are chosen randomly.
 The variations were small, though.
@@
 
-The histograms of which rates the algorithms decided to send at provided some useful insight into the behavior.
+The histograms of which rates the algorithms decided to send at provided some useful insight into their behavior.
 Table~\ref{table:2} is a histogram of a 35 second corner trace.
@@
 
-Our examination of the histograms showed us that Minstrel spends a lot of time sampling at high rates.
+Our examination of the histograms showed us that Minstrel spends a lot of time sending at high rates.
 In the specific case we examined, they were all probe packets.
@@
 
-In the next section we will talk about Minproved, an improved version of Minstrel where we attempt to fix the probability problem.
+Additionally, Minstrel uses a complex method of calculating when to send a sample packet that involves a heuristic to make sure that 10\% of packets were actually samples.
+This is because the mrr places sample packets second in the retry chain when the sample rate is slower than the rate with the best throughput.
+So even if 10\% of probes have a sample rate in the retry chain, a sample rate isn't actually sent 10\% of the time.
+We implemented this heuristic and it tended to cause Minstrel to send too many probe packets.
+
+In the next section we will talk about Minproved, an improved version of Minstrel where we attempt to fix problems with EWMA and too-frequent probing
 

### 2013-05-16 pysim/minstrel.py

@@ -176,7 +176,7 @@ def apply_rate(cur_time): #cur_time is in nanoseconds
     # 3  | Best probability | Best probability  | Best probability
     # 4  | Lowest Baserate  | Lowest baserate   | Lowest baserate
     #if randint(1,100) <= 10:
-    delta = (npkts*.1 - np + nd/2.0)
+    delta = (npkts*.1 - (np + nd/2.0))
     if delta  > 0: #random!
         #Analysis of information showed that the system was sampling too hard
         #at some rates. For those rates that never work (54mb, 500m range) 

### 2013-05-16 paper/paper.tex
tweak

--- 2013-05-16 paper/paper.tex
@@
 This implementation is slightly different from how it was implemented in the MadWifi kernel driver.
-The primary difference is that the kernel implementatiion uses a EWMA instead of a 10 second window.
-The thesis implementation computes success probabilities based on a 10-second sliding window.
+The primary difference is that the kernel implementatiion uses a EWMA, while the thesis implementation computes average transmission times based on a 10-second sliding window.
 It would be worth looking at the performance of EWMA-based SampleRate in the future.

### 2013-05-16 paper/paper.tex
Minproved section.

@@
 \begin{document}
+
 %don't want date printed
@@
 
-SampleRate has three main functions: \texttt{ApplyRate()}, which returns a the bit-rate to send a packet at, \texttt{ProcessFeedback()}, which updates the statistics for a bit-rate after an attempted packet send,  and \texttt{RemoveStaleResults()}, which removes all statistics older than 10 seconds.
+SampleRate has three main functions: \texttt{ApplyRate()}, which returns a the bit-rate to send a packet at, \texttt{ProcessFeedback()}, which updates the statistics for a bitrate after an attempted packet send,  and \texttt{RemoveStaleResults()}, which removes all statistics older than 10 seconds.
 
@@
 
-\section{Analysis}
+\section{Analysis} \label{sec:analysis}
 
@@
 Figure~\ref{figure:3} shows the results of a single run of both Minstrel and SampleRate over all the different types of traces we collected.
-This is consistent with the literature---the authors of Minstrel noted that they focused on making Minstrel robust to poor conditions~\cite{minstrel}, which means losing potential throughput gains in stable situations.
+This is consisten with the literature---the authors of Minstrel noted that they focused on making Minstrel robust to poor conditions~\cite{minstrel}, which means losing potential througput gains in stable situations.
 Conversely, John Bicket noted that SampleRate performs better in stable situations than noisy or highly varying conditions~\cite{samplerate}.
-We were slightly surprised to see how significantly SampleRate outperformed Minstrel in certain situations, sometimes with throughput as much as 67\% higher.
+We were slightly surprised to see how significantly SampleRate outperformed Minstrel in certain situations, though, with throughput being up to $67\%$ higher.
 
-In Figure~\ref{figure:2} we show the performance of constant bit-rates on the \texttt{clear\_1.dat} trace, which is a 35 second trace in clear view of the access point.
-The optimal constant rate here was 18 Mbps, which achieved an average throughput of about 15 Mbps.
-The next highest bit-rate, 24 Mbps, shows a large plummet in performance, down to a throughput of 1.17 Mbps.
-This steep drop is characteristic, and is also talked about in the SampleRate thesis~\cite{samplerate}.
-It demonstrates just how important bit-rate selection is---the optimal bit-rate has typically much higher throughput than the next-best rate, but if the algorithm strays too high the results can be devastating.
+In Figure~\ref{figure:2} we show the performance of constant bitrates on the \texttt{clear\_1.dat} trace, which is a 35 second trace in clear view of the access point.
+The optimal rate here was 18 Mbps, which achieved an average throughput of about 15 Mbps.
+The next highest bitrate, 24 Mbps, shows a large plummet in performance, down to a throughput of 1.17 Mbps.
+This steep drop is charateristic, and is also talked about in the SampleRate thesis~\cite{samplerate}.
+It demonstrates just how important bit rate selection is---the optimal bit rate has typically much higher throughput than the next-best rate, but if the algorithm strays too high the results can be devastating.
+
+Bit rate selection algorithms strive to have a throughput near or greater than the optimal constant rate.
+For this \texttt{clear\_1.dat} trace, SampleRate clearly meets this goal, achieving an average throughput more than 1 Mbps higher than the optimal constant rate.
+In noisier situations, though, Minstrel tends to be closer to the optimal constant rate.
 
@@
 
-Bit-rate selection algorithms naturally show variation between runs on the same data because probe rates are chosen randomly.
+Bit rate selection algorithms naturally show variation between runs on the same data because probe rates are chosen randomly.
 The variations were small, though.
@@
 SampleRate was similarly stable, it averaged 5.284 Mbps over 10 runs, with a maximum variation of about +/- 0.3 Mbps.
-The variation of results on other traces was similar.
+The variation of results on other traces (except the very short traces, where variance was naturally higher) was similar.
 
-Additionally, Minstrel uses a complex method of calculating when to send a sample packet that involves a heuristic to make sure that 10\% of packets were actually samples.
-This is because the mrr places sample packets second in the retry chain when the sample rate is slower than the rate with the best throughput.
-So even if 10\% of probes have a sample rate in the retry chain, a sample rate isn't actually sent 10\% of the time.
-We implemented this heuristic and it tended to cause Minstrel to send too many probe packets.
+The histograms of which rates the algorithms decided to send at provided some useful insight into their behavior.
+Table~\ref{table:2} is a histogram of a 35 second corner trace.
+In this situation Minstrel outperforms SampleRate.
+The constant rate throughputs demonstrate that 11, 12 and 18 Mbps were the optimal rates to send at, and that no packets were sucessful at any higher rates.
+SampleRate does not use these rates except for probe packets, which happens once every ten seconds.
+Minstrel, since it does not terminate transmission after four successive failures, spends a lot of time sending probe packets at these high rates, which are never successful.
+The only reason low rates have fewer probe packets is because Minstrel explicitly places them second in the retry chain, and because the main rate usually succeeds.
+Despite the fact that SampleRate spends less time probing high rates, it is less successful.
+This is because Minstrel send the most packets at the three most successful bitrates.
+SampleRate, however, spends too much time sampling low bit rates.
+Additionally, it spends a lot of time sending at 5.5 Mbps.
+TODO: why does SampleRate spend time at 5.5 Mbps???
+We suspect that this might be because the equation SampleRate uses to estimate average transmission time needs to be tweaked.
 
-In the next section we will talk about Minproved, an improved version of Minstrel where we attempt to fix problems with EWMA and too-frequent probing
+Our examination of the histograms showed us that Minstrel spends a lot of time sampling at high rates.
+In the specific case we examined, they were all probe packets.
+But sometimes Minstrel would choose 54Mbps as the rate with the best throughput, which made little sense.
+A closer investigation revealed that Minstrel would sometimes get lucky with a successful probe at 54Mbps.
+The probability of success would be 1, so then 54Mbps would be chosen as the rate with the best throughput.
+This is unfortunate, but the real harm comes in the next window: packets would now be send at 54Mbps, and nearly all would fail.
+However, since these hundreds of failed packets weight equally with the single successful probe packet, Minstrel still considers 54 Mbps the best rate.
+Thus the probability of success might drop from 1 to about 0.70, which is still high enough for the throughput equation to rank 54 Mbps highly.
+Fundamentally, the problem is that the EWMA does not account the second block having hundreds more packets than the first.
+In the next section we will talk about Minproved, an improved version of Minstrel where we attempt to fix this problem.
 
@@
   \hspace{-1em}\includegraphics[width=3.4in]{constant.png}\vspace{-0em}
-  \caption{Throughput of using a constant 802.11b/g bit-rate in clear line-of-sight from the access point.}
+  \caption{Throughput of using a constant 802.11b/g bitrate in clear line-of-sight from the access point.}
 \label{figure:2}
@@
 
-\subsection{Improvements to Minstrel}
-TODO: Minproved.
--make sure we mention how Minproved almost always does better than highest fixed rate
+\section{Improvements to Minstrel}
+
+In Section~\ref{sec:analysis}, we noted that the EWMA used by Minstrel causes problems when probe packets succeed.
+A natural next step was to modify Minstrel to avoid this problem, by weighing each 100ms block of statistics in proportion to how many packets it contains.
+To preserve the exponential weighing of statistics, we chose to weigh each 100ms block based on the number of packets sent, compared to the those sent in the average block.
+Parameters were chosen so that if all blocks were of average size, our ``balanced'' EWMA would behave identically to a normal EWMA.
+
+To achieve, this weighing, we first recast the usual EWMA algorithm.
+The usual algorithm computes $$p \gets \alpha p + (1 - \alpha) \frac{n_i}{d_i}),$$ where $p$ is the EWMA probability, $\alpha$ is a weight factor equal to $0.75$ in Minstrel, and $n_i$ and $d_i$ are the number of successful and total packets in the 100ms window.
+We next write this as $$p \gets \frac{\beta p + \frac{n_i}{d_i}}{\beta + 1},$$ where $\beta = \alpha / (1 - \alpha)$.
+We can now change the factor in front of $n_i / d_i$ to account for the total number of packets sent.
+In our case, we chose to compute $$p \gets \frac{\beta p + w_i \frac{n_i}{d_i}}{\beta + w_i},$$ where the weight $w_i$ is $d_i / (d / b)$, with $d$ the total number of packets ever send and $b$ the total number of 100ms windows ever seen.
+Then $d / b$ is the average number of packets per window and $d_i / (d / b)$ is the ratio between the current window and the average window.
+Note that if the current window is of average size, this is exactly equal to the usual EWMA result.
+Finally, we can simplify the above equation to $$S p \gets \frac{\beta \frac{d}{b} p + S n_i}{\beta \frac{d}{b} + d_i},$$ where $S$ is any constant.
+This equation is usable in fixed-precision arithmetic, making it simple to implement in the Linux kernel (which prefers to avoid floating point computations for portability reasons).
+Because packets are high rates rarely succeed, the average window size is a few packets; thus, the main qualitative difference between our balanced EWMA and the original Minstrel EWMA is that non-probe packets at probe bit-rates are weighed extraordinarily heavily, and this fact allows us to avoid multiple failing 100ms windows.
+
+We tested a modified Minstrel which used this more balanced EWMA algorithm on the same traces as Minstrel and SampleRate, and found a consistent improvement of approximately 1 Mbps over the normal Minstrel algorithm.
+
+We noted another curious fact about Minstrel: the kernel implementation of Minstrel samples rates very frequently.
+This is due to a flag the kernel uses to track which probe bit-rates were actually sampled.
+In cases where the probe packets are at a lower bit-rate than the current bit-rate, the probe packets might never be sent if the current bit-rate succeeds.
+However, the flagging seems to misbehave in certain cases, causing the extremely aggressive probing we see in some traces.
+For example, in one of the traces where the client was around a corner from the access point, Minstrel sent almost half of its packets at probe frequencies.
+We modified Minstrel to be less aggressive with probe packets, decreasing the sampling frequency to approximately 10\%.
+This lead to more improvement, especially on noisy traces such as the cases where the client was around a corner or moving.
+
+Overall, our improved Minstrel, which we call Minproved, often out-performs Minstrel by multiple Mbps and never doing any worse; and usually surpassing the best constant bit rate.
+In only one case did SampleRate achieve higher throughput than Minproved, and in this case both surpassed the best constant bit-rate.
+In cases where Minstrel out-performed SampleRate, the same behavior was true of Minproved.
+In most cases, Minproved was $30\%$ faster than Minstrel, and in some the improvement as great as $60\%$.
+
+However, Minproved still makes many poor choices.
+While a single successful probe packet does not cause many hundreds of milliseconds of failed packets, as it does in Minstrel, the single successful packet can still cause one window's worth of failed packets.
+Instead, it would be best if a successful probe packet at a rarely-successful bit-rate was treated more carefully.
+One can imagine an algorithm, for example, which would treat a single successful probe packet as a reason to send a dozen
+
 \begin{table*}[htb]
@@
 Minstrel and SampleRate perform similarly.
-The results suggest that SampleRate tends to do slightly better, except in very lossy links such as the client being far away from the AP or actively moving.
+The results suggest that SampleRate tends to do slightly better, except in very lossy links such as the client being far awat from the AP or actively moving.
 Minproved, our improved version of Minstrel, solidly outperforms both Minstrel and SampleRate.

### 2013-05-16 paper/paper.tex
Line breaks.

@@
 \subsection*{Abstract}
-In this paper we compare the performance SampleRate and Minstrel, two popular bit-rate selection algorithms that have widespread real-world usage on the Linux operating system.
-We use a novel trace-based approach that allows us avoid in-depth kernel programming and write simple Python scripts instead.
-We also introduce improvements to the Minstrel algorithm that allow for significant gains in throughput.
+In this paper we compare the performance SampleRate and Minstrel, two
+popular bit-rate selection algorithms that have widespread real-world
+usage on the Linux operating system. We use a novel trace-based
+approach that allows us avoid in-depth kernel programming and write
+simple Python scripts instead. We also introduce improvements to the
+Minstrel algorithm that allow for significant gains in throughput.
 \section{Introduction}
 
-One of the key ways that wireless networks differ from wired is that wireless networks have varying link rates.
-Link conditions vary with time due to interference from other devices, changing network geometry, and mobile clients.
-An optimal rate at some time $t$ may be different from the optimal rate just 30 seconds earlier.
-A good bit-rate selection algorithm has to detect and adapt to these conditions.
-If the chosen bit-rate is too slow, then the throughput will be unnecessarily low.
-If the rate is too high, then failures will be very frequent and throughput will again suffer.
+One of the key ways that wireless networks differ from wired is that
+wireless networks have varying link rates. Link conditions vary with
+time due to interference from other devices, changing network
+geometry, and mobile clients. An optimal rate at some time $t$ may be
+different from the optimal rate just 30 seconds earlier. A good
+bit-rate selection algorithm has to detect and adapt to these
+conditions. If the chosen bit-rate is too slow, then the throughput
+will be unnecessarily low. If the rate is too high, then failures will
+be very frequent and throughput will again suffer.
 
-There are three main classes of bit-rate adaptation protocols: frame-based, SNR-based, and cross-layer protocols.
-Frame-based protocols measure the fraction of successfully received packets.
-SNR protocols make decisions based on the estimated Signal to Noise ratio.
-Cross-layer protocols use SoftPHY data from the physical layer.
-The most commonly implemented protocols on today's networks are frame based, because SNR protocols perform poorly~\cite{samplerate}, and cross-layer protocols cannot be deployed on current networks as they violate the network layering abstraction.
+There are three main classes of bit-rate adaptation protocols:
+frame-based, SNR-based, and cross-layer protocols. Frame-based
+protocols measure the fraction of successfully received packets. SNR
+protocols make decisions based on the estimated Signal to Noise
+ratio. Cross-layer protocols use SoftPHY data from the physical
+layer. The most commonly implemented protocols on today's networks are
+frame based, because SNR protocols perform poorly~\cite{samplerate},
+and cross-layer protocols cannot be deployed on current networks as
+they violate the network layering abstraction.
 
-The two most popular frame-based protocols are SampleRate and Minstrel.
-Both were implemented in the MadWifi drivers for Linux wireless, and today Minstrel is the default bit-rate selection algorithm for all wireless drivers on Linux.
-Our research used a trace-based approach to analyze the performance of these two algorithms.
-We also created a modified version of Minstrel that provides significant throughput gains over the vanilla Minstrel implementation.
+The two most popular frame-based protocols are SampleRate and
+Minstrel. Both were implemented in the MadWifi drivers for Linux
+wireless, and today Minstrel is the default bit-rate selection
+algorithm for all wireless drivers on Linux. Our research used a
+trace-based approach to analyze the performance of these two
+algorithms. We also created a modified version of Minstrel that
+provides significant throughput gains over the vanilla Minstrel
+implementation.
 
-This paper will provide a general overview of SampleRate and Minstrel.
-We then discuss our testing methodology.
-We compare the performance of the two algorithms, and then discuss modifications we made to Minstrel and compare it's performance to the current kernel implementation.
+This paper will provide a general overview of SampleRate and
+Minstrel. We then discuss our testing methodology. We compare the
+performance of the two algorithms, and then discuss modifications we
+made to Minstrel and compare it's performance to the current kernel
+implementation.
+
 \section{SampleRate}
 
-SampleRate has three main functions: \texttt{ApplyRate()}, which returns a the bit-rate to send a packet at, \texttt{ProcessFeedback()}, which updates the statistics for a bitrate after an attempted packet send,  and \texttt{RemoveStaleResults()}, which removes all statistics older than 10 seconds.
+SampleRate has three main functions: \texttt{ApplyRate()}, which
+returns a the bit-rate to send a packet at,
+\texttt{ProcessFeedback()}, which updates the statistics for a bitrate
+after an attempted packet send, and \texttt{RemoveStaleResults()},
+which removes all statistics older than 10 seconds.
 
-SampleRate sends at the rate with the lowest average transmission time.
-Every ten packets, it sends out a sample packet at a random rate that has a smaller predicted transmission time than the current rate.
-SampleRate updates its transmission time estimates every ten seconds based on feedback from the card about which packets succeeded and at which rate they were sent.
-If the current best rate experiences four successive failures, SampleRate stops using that rate.
+SampleRate sends at the rate with the lowest average transmission
+time. Every ten packets, it sends out a sample packet at a random rate
+that has a smaller predicted transmission time than the current
+rate. SampleRate updates its transmission time estimates every ten
+seconds based on feedback from the card about which packets succeeded
+and at which rate they were sent. If the current best rate experiences
+four successive failures, SampleRate stops using that rate.
 
@@
 
-Minstrel utilizes the multi-rate retry chain (mrr), which is an array that tells the card which rates to try before reporting a failure.
-The retry chain makes failures extremely unlikely.
-Minstrel has a mrr of [\emph{highest throughput, next highest throughput, best probability, lowest base rate}] for most packets.
-In essence, Minstrel tells the card to send at the highest throughput rates but to fall back to reliable rates upon failure.
-The throughput and success probability are calculated for each rate every 100ms.
-To keep an accurate estimate of throughput and probability for each rate, Minstrel sends sample frames ten percent of the time.
-If the randomly chosen rate has a higher lossless throughput than the current optimal rate, the mrr is as follows: [\emph{random, highest throughput, best probability, lowest base rate}].
-If the random rate is slower than the optimal rate, it is moved further back in the retry chain: [\emph{best throughput, random, best probability, lowest base rate}].
-This ensures that Minstrel never samples rates worse than the current optimal rate unless the optimal rate experiences a failure.
-See table ~\ref{table:1} for an overview of the mrr's used in Minstrel.
+Minstrel utilizes the multi-rate retry chain (mrr), which is an array
+that tells the card which rates to try before reporting a failure. The
+retry chain makes failures extremely unlikely. Minstrel has a mrr of
+[\emph{highest throughput, next highest throughput, best probability,
+    lowest base rate}] for most packets. In essence, Minstrel tells
+the card to send at the highest throughput rates but to fall back to
+reliable rates upon failure. The throughput and success probability
+are calculated for each rate every 100ms.  To keep an accurate
+estimate of throughput and probability for each rate, Minstrel sends
+sample frames ten percent of the time. If the randomly chosen rate has
+a higher lossless throughput than the current optimal rate, the mrr is
+as follows: [\emph{random, highest throughput, best probability,
+    lowest base rate}]. If the random rate is slower than the optimal
+rate, it is moved further back in the retry chain: [\emph{best
+    throughput, random, best probability, lowest base rate}]. This
+ensures that Minstrel never samples rates worse than the current
+optimal rate unless the optimal rate experiences a failure. See table
+~\ref{table:1} for an overview of the mrr's used in Minstrel.
 
-Minstrel's implementation is based on SampleRate's, so it also has \texttt{ApplyRate()} and \texttt{ProcessFeedback()}, as well as a \texttt{UpdateStats()} method that runs every 100ms.
-\texttt{UpdateStats()} is home to some of the key differences between Minstrel and SampleRate.
-Instead of making decisions based on the average
-transmission time, Minstrel uses throughput: \begin{equation} T = \frac{p}{tx\_time}\end{equation} where $T$ is the throughput of some rate $r$, $p$ is the probability of successfully transmitting at $r$, and $tx\_time$ is the lossless transmission time at $r$.
+Minstrel's implementation is based on SampleRate's, so it also has
+\texttt{ApplyRate()} and \texttt{ProcessFeedback()}, as well as a
+\texttt{UpdateStats()} method that runs every
+100ms. \texttt{UpdateStats()} is home to some of the key differences
+between Minstrel and SampleRate. Instead of making decisions based on
+the average transmission time, Minstrel uses
+throughput: \begin{equation} T = \frac{p}{tx\_time}\end{equation}
+where $T$ is the throughput of some rate $r$, $p$ is the probability
+of successfully transmitting at $r$, and $tx\_time$ is the lossless
+transmission time at $r$.
 
-The probability $p$ of a successful transmission is calculated from statistics collected by the \texttt{ProcessFeedback()} method, and is done using an exponential weighted moving average, or EWMA.
-The EWMA creates a weighted average that weighs recent data more heavily than old data.
-The weighting for old data decreases exponentially.
-An EWMA is beneficial because it ensures that a sudden degradation in link quality will create a rapid response in the probabilities due to the increased weight of recent data.
-Minstrel will waste less time sending packets at a rate that no longer works, as compared to a simple probability calculation.
+The probability $p$ of a successful transmission is calculated from
+statistics collected by the \texttt{ProcessFeedback()} method, and is
+done using an exponential weighted moving average, or EWMA. The EWMA
+creates a weighted average that weighs recent data more heavily than
+old data. The weighting for old data decreases exponentially. An EWMA
+is beneficial because it ensures that a sudden degradation in link
+quality will create a rapid response in the probabilities due to the
+increased weight of recent data. Minstrel will waste less time sending
+packets at a rate that no longer works, as compared to a simple
+probability calculation.
 
@@
 
-We collected traces about the success rates of each bit rate through a modified Linux wireless driver; Minstrel and SampleRate were then re-implemented in Python and replayed on this collected data.
+We collected traces about the success rates of each bit rate through a
+modified Linux wireless driver; Minstrel and SampleRate were then
+re-implemented in Python and replayed on this collected data.
 
@@
 
-Our initial plan was to use the algorithms as implemented in the MadWifi drivers for the Atheros chipset.
-However, MadWifi is deprecated, and only runs on now-ancient network cards.
-In addition to the unavailability of compatible hardware, MadWifi would not be an accurate reflection of contemporary users.
-Both ath5k and ath9k, which were created to replace MadWifi, did not port the old bit-rate selection system from MadWifi and instead opted to use the Linux kernel's rate selection framework, which only implements Minstrel.
-Porting SampleRate over to the new drivers would not have been trivial, as there are vast differences in the assumed interfaces.
+Our initial plan was to use the algorithms as implemented in the
+MadWifi drivers for the Atheros chipset. However, MadWifi is
+deprecated, and only runs on now-ancient network cards. In addition to
+the unavailability of compatible hardware, MadWifi would not be an
+accurate reflection of contemporary users. Both ath5k and ath9k, which
+were created to replace MadWifi, did not port the old bit-rate
+selection system from MadWifi and instead opted to use the Linux
+kernel's rate selection framework, which only implements Minstrel.
+Porting SampleRate over to the new drivers would not have been
+trivial, as there are vast differences in the assumed interfaces.
 
-At this point, we decided to take a different approach and use an approach similar to that used in Sprout~\cite{sprout}.
-We modified the ath9k driver to try bit rates uniformly and report packet successes and failures, then analyze this data in user-space with re-implementations of SampleRate and Minstrel.
-Not only did this lessen the re-implementation cost (since the algorithms were re-implemented in user-space in Python, not in kernel-mode in C), but it also provided reproducibility and allowed us to investigate modifications to SampleRate and to Minstrel.
+At this point, we decided to take a different approach and use an
+approach similar to that used in Sprout~\cite{sprout}.  We modified
+the ath9k driver to try bit rates uniformly and report packet
+successes and failures, then analyze this data in user-space with
+re-implementations of SampleRate and Minstrel.  Not only did this
+lessen the re-implementation cost (since the algorithms were
+re-implemented in user-space in Python, not in kernel-mode in C), but
+it also provided reproducibility and allowed us to investigate
+modifications to SampleRate and to Minstrel.
 
@@
 
-Traces were collected from an ath9k driver modified to sample bit-rates uniformly (forgoing the usual bit rate selection algorithm) and to report successes and failure from each packet.
-Each bit-rate was sampled uniformly in time -- thus, more packets were sent at higher bit-rates, since those bit-rates otherwise deliver packets faster.
-For each packet sent, the time to send was recorded, and the bit rate for the next packet was chosen to be the bit-rate that had the smallest total transmissions time.
-This guarantees approximately uniform selection of bit-rates.
-The wireless card used was an Qualcomm Atheros \texttt{AR9285} Wireless Network Adapter.
+Traces were collected from an ath9k driver modified to sample
+bit-rates uniformly (forgoing the usual bit rate selection algorithm)
+and to report successes and failure from each packet.  Each bit-rate
+was sampled uniformly in time -- thus, more packets were sent at
+higher bit-rates, since those bit-rates otherwise deliver packets
+faster.  For each packet sent, the time to send was recorded, and the
+bit rate for the next packet was chosen to be the bit-rate that had
+the smallest total transmissions time.  This guarantees approximately
+uniform selection of bit-rates.  The wireless card used was an
+Qualcomm Atheros \texttt{AR9285} Wireless Network Adapter.
 
 Traces were collected at an unused access point in a basement of MIT.
-Before initiating trace collection, the wireless card was put in monitor mode for a minute to check that no other users were connected to the access point.
-None were, guaranteeing that traces would not include delays from carrier sensing or run into problems due to congestion at the access point.
-Traffic was simulated by sending 1500 byte packets to the access point's IP address.
-Packets were sent over UDP to avoid the TCP congestion window and retry loop; since no users were using the access point, there was no real risk of deteriorating service for other users.
-Figure~\ref{figure:1} shows a floor plan of the access point and basement.
+Before initiating trace collection, the wireless card was put in
+monitor mode for a minute to check that no other users were connected
+to the access point.  None were, guaranteeing that traces would not
+include delays from carrier sensing or run into problems due to
+congestion at the access point.  Traffic was simulated by sending 1500
+byte packets to the access point's IP address.  Packets were sent over
+UDP to avoid the TCP congestion window and retry loop; since no users
+were using the access point, there was no real risk of deteriorating
+service for other users.  Figure~\ref{figure:1} shows a floor plan of
+the access point and basement.
 
-Eighteen total traces were collected, grouped into a number of scenarios.
-Each trace was 30 seconds to a minute long, and each scenario was repeated twice.
-One scenario placed the wireless card with line of sight to the access point, two meters from it.
-Another created solid body interference by having one of the authors stand between the wireless device and the access point.
-One placed the device around a corner from the access point, creating some potential for multi-path interference and fading.
-In one scenario, we moved the wireless device throughout the trace, pacing back and forth in a 10 meter line in front of the access point.
-Finally, one five-minute trace was taken, which mixed the above scenarios, first moving around a corner, then standing still, then moving to line-of-sight with the access point.
-A few shorter (3 to 15 second long) traces were collected to investigate the startup behavior of bit-rate selection.
+Eighteen total traces were collected, grouped into a number of
+scenarios.  Each trace was 30 seconds to a minute long, and each
+scenario was repeated twice.  One scenario placed the wireless card
+with line of sight to the access point, two meters from it.  Another
+created solid body interference by having one of the authors stand
+between the wireless device and the access point.  One placed the
+device around a corner from the access point, creating some potential
+for multi-path interference and fading.  In one scenario, we moved the
+wireless device throughout the trace, pacing back and forth in a 10
+meter line in front of the access point.  Finally, one five-minute
+trace was taken, which mixed the above scenarios, first moving around
+a corner, then standing still, then moving to line-of-sight with the
+access point.  A few shorter (3 to 15 second long) traces were
+collected to investigate the startup behavior of bit-rate selection.
 
@@
 
-Once collected, traces were fed to a bit rate selection simulator to analyze the performance of Minstrel and SampleRate.
-Both algorithms were represented by two functions: \texttt{ApplyRate}, which returned a multi-rate retry chain of rates to attempt transmission at, and \texttt{ProcessFeedback}, which was passed the success or failure of the transmission and how many attempts were made at each bit rate (along with the current time and the total time to send the packet); this parallels the interface used by the Linux kernel itself.
-Helper functions, like SampleRate's \texttt{RemoveStaleResults} function, were instead called from one of \verb|apply_rate| or \verb|process_feedback|.
-For every packet transmission, the simulator would call \texttt{ApplyRate}; compute the probability of success based on packets sent in 10-millisecond window around the current time; and from that determine the success or failure of the packet, and the number of retries required.
-The results of a test run are thus non-deterministic; both algorithms tested are non-deterministic anyway, so this is no handicap.
-To improve re-implementation fidelity, both implementations were meticulously checked by both authors.
+Once collected, traces were fed to a bit rate selection simulator to
+analyze the performance of Minstrel and SampleRate.  Both algorithms
+were represented by two functions: \texttt{ApplyRate}, which returned
+a multi-rate retry chain of rates to attempt transmission at, and
+\texttt{ProcessFeedback}, which was passed the success or failure of
+the transmission and how many attempts were made at each bit rate
+(along with the current time and the total time to send the packet);
+this parallels the interface used by the Linux kernel itself.  Helper
+functions, like SampleRate's \texttt{RemoveStaleResults} function,
+were instead called from one of \verb|apply_rate| or
+\verb|process_feedback|.  For every packet transmission, the simulator
+would call \texttt{ApplyRate}; compute the probability of success
+based on packets sent in 10-millisecond window around the current
+time; and from that determine the success or failure of the packet,
+and the number of retries required.  The results of a test run are
+thus non-deterministic; both algorithms tested are non-deterministic
+anyway, so this is no handicap.  To improve re-implementation
+fidelity, both implementations were meticulously checked by both
+authors.
 
-We implemented SampleRate as outlined in John Bicket's Master's thesis~\cite{samplerate}.
-This implementation is slightly different from how it was implemented in the MadWifi kernel driver.
-The primary difference is that the kernel implementatiion uses a EWMA, while the thesis implementation computes average transmission times based on a 10-second sliding window.
-It would be worth looking at the performance of EWMA-based SampleRate in the future.
-Our implementation, \texttt{samplerate.py} contains \texttt{ApplyRate()}, \texttt{ProcessFeedback()} and  \texttt{RemoveStaleResults()}, as well as a few helper functions and data structures for tracking rate statistics.
+We implemented SampleRate as outlined in John Bicket's Master's
+thesis~\cite{samplerate}. This implementation is slightly different
+from how it was implemented in the MadWifi kernel driver. The primary
+difference is that the kernel implementatiion uses a EWMA, while the
+thesis implementation computes average transmission times based on a
+10-second sliding window. It would be worth looking at the performance
+of EWMA-based SampleRate in the future.  Our implementation,
+\texttt{samplerate.py} contains \texttt{ApplyRate()},
+\texttt{ProcessFeedback()} and \texttt{RemoveStaleResults()}, as well
+as a few helper functions and data structures for tracking rate
+statistics.
 
-We implemented Minstrel by porting the C code from the 3.3.8 version of the Linux kernel into Python.
-\texttt{minstrel.py} contains \texttt{ApplyRate()}, \texttt{ProcessFeedback()} and  \texttt{UpdateStats()}, as well as a few helper functions and data structures for tracking rate statistics.
+We implemented Minstrel by porting the C code from the 3.3.8 version
+of the Linux kernel into Python. \texttt{minstrel.py} contains
+\texttt{ApplyRate()}, \texttt{ProcessFeedback()} and
+\texttt{UpdateStats()}, as well as a few helper functions and data
+structures for tracking rate statistics.
 
@@
 
-There are a few parameters that can be adjusted in Minstrel and SampleRate, but our implementation is consistent with what the respective authors originally chose.
+There are a few parameters that can be adjusted in Minstrel and
+SampleRate, but our implementation is consistent with what the
+respective authors originally chose.
 
-In SampleRate, the number of successive retries, the frequency of probe packets, the window size, and the lossless $tx\_time$ estimations are all parameters that can be changed.
-The SampleRate thesis chose 4 successive retries, a probe frequency of every ten packets, and a window size of ten seconds.
-We used the same values in our implementation.
-Additionally, we use the same equation to calculate $tx\_time$.
+In SampleRate, the number of successive retries, the frequency of
+probe packets, the window size, and the lossless $tx\_time$
+estimations are all parameters that can be changed. The SampleRate
+thesis chose 4 successive retries, a probe frequency of every ten
+packets, and a window size of ten seconds. We used the same values in
+our implementation. Additionally, we use the same equation to
+calculate $tx\_time$.
 
-In Minstrel, the frequency of probe packets, the window size, and the EWMA weighting are the adjustable parameters.
-The 3.3.8 Linux kernel uses a probe frequency of 10\%, a window size of 100ms, and a EWMA weighting of 0.75.
-We use the same values.
-Additionally, Minstrel also estimates $tx\_time$ using the \texttt{ieee80211\_frame\_duration()} method from the \texttt{util.c} file in \texttt{net/mac80211}.
-We implemented the same function in Python.
+In Minstrel, the frequency of probe packets, the window size, and the
+EWMA weighting are the adjustable parameters. The 3.3.8 Linux kernel
+uses a probe frequency of 10\%, a window size of 100ms, and a EWMA
+weighting of 0.75. We use the same values. Additionally, Minstrel also
+estimates $tx\_time$ using the \texttt{ieee80211\_frame\_duration()}
+method from the \texttt{util.c} file in \texttt{net/mac80211}. We
+implemented the same function in Python.
 
@@
 
-Minstrel and SampleRate perform similarly, usually within 3 Mbps of each other.
-In especially noisy situations, such as when the client is moving or when the client is around the corner from the access point, Minstrel consistently outperforms SampleRate.
-Figure~\ref{figure:3} shows the results of a single run of both Minstrel and SampleRate over all the different types of traces we collected.
-This is consisten with the literature---the authors of Minstrel noted that they focused on making Minstrel robust to poor conditions~\cite{minstrel}, which means losing potential througput gains in stable situations.
-Conversely, John Bicket noted that SampleRate performs better in stable situations than noisy or highly varying conditions~\cite{samplerate}.
-We were slightly surprised to see how significantly SampleRate outperformed Minstrel in certain situations, though, with throughput being up to $67\%$ higher.
+Minstrel and SampleRate perform similarly, usually within 3 Mbps of
+each other. In especially noisy situations, such as when the client is
+moving or when the client is around the corner from the access point,
+Minstrel consistently outperforms SampleRate. Figure~\ref{figure:3}
+shows the results of a single run of both Minstrel and SampleRate over
+all the different types of traces we collected. This is consisten with
+the literature---the authors of Minstrel noted that they focused on
+making Minstrel robust to poor conditions~\cite{minstrel}, which means
+losing potential througput gains in stable situations. Conversely,
+John Bicket noted that SampleRate performs better in stable situations
+than noisy or highly varying conditions~\cite{samplerate}. We were
+slightly surprised to see how significantly SampleRate outperformed
+Minstrel in certain situations, though, with throughput being up to
+$67\%$ higher.
 
-In Figure~\ref{figure:2} we show the performance of constant bitrates on the \texttt{clear\_1.dat} trace, which is a 35 second trace in clear view of the access point.
-The optimal rate here was 18 Mbps, which achieved an average throughput of about 15 Mbps.
-The next highest bitrate, 24 Mbps, shows a large plummet in performance, down to a throughput of 1.17 Mbps.
-This steep drop is charateristic, and is also talked about in the SampleRate thesis~\cite{samplerate}.
-It demonstrates just how important bit rate selection is---the optimal bit rate has typically much higher throughput than the next-best rate, but if the algorithm strays too high the results can be devastating.
+In Figure~\ref{figure:2} we show the performance of constant bitrates
+on the \texttt{clear\_1.dat} trace, which is a 35 second trace in
+clear view of the access point. The optimal rate here was 18 Mbps,
+which achieved an average throughput of about 15 Mbps. The next
+highest bitrate, 24 Mbps, shows a large plummet in performance, down
+to a throughput of 1.17 Mbps. This steep drop is charateristic, and is
+also talked about in the SampleRate thesis~\cite{samplerate}. It
+demonstrates just how important bit rate selection is---the optimal
+bit rate has typically much higher throughput than the next-best rate,
+but if the algorithm strays too high the results can be devastating.
 
-Bit rate selection algorithms strive to have a throughput near or greater than the optimal constant rate.
-For this \texttt{clear\_1.dat} trace, SampleRate clearly meets this goal, achieving an average throughput more than 1 Mbps higher than the optimal constant rate.
-In noisier situations, though, Minstrel tends to be closer to the optimal constant rate.
+Bit rate selection algorithms strive to have a throughput near or
+greater than the optimal constant rate. For this \texttt{clear\_1.dat}
+trace, SampleRate clearly meets this goal, achieving an average
+throughput more than 1 Mbps higher than the optimal constant rate. In
+noisier situations, though, Minstrel tends to be closer to the optimal
+constant rate.
 
-We did notice that Minstrel occasionally has very poor performance on traces 1 seconds or shorter---sometimes as low as 2 Mbps.
-This is because it does not stop sending after four successive failures, but must wait until the end of the 100ms window for statistics to be compiled.
-Thus it wastes a lot of time sending at non-optimal rates; and a short trace has only a few 100ms windows.
-This behavior is not evident in long traces, however, as Minstrel has had time to both smooth out bad windows and collect more statistics (making it less likely to choose a bad rate).
+We did notice that Minstrel occasionally has very poor performance on
+traces 1 seconds or shorter---sometimes as low as 2 Mbps. This is
+because it does not stop sending after four successive failures, but
+must wait until the end of the 100ms window for statistics to be
+compiled. Thus it wastes a lot of time sending at non-optimal rates;
+and a short trace has only a few 100ms windows. This behavior is not
+evident in long traces, however, as Minstrel has had time to both
+smooth out bad windows and collect more statistics (making it less
+likely to choose a bad rate).
 
-Bit rate selection algorithms naturally show variation between runs on the same data because probe rates are chosen randomly.
-The variations were small, though.
-On ten runs on the \texttt{clear\_1.dat} trace, the average of Minstrel over 10 runs was 7.52 Mbps, with a maximum variation of +/- 0.12 Mbps.
-SampleRate was similarly stable, it averaged 5.284 Mbps over 10 runs, with a maximum variation of about +/- 0.3 Mbps.
-The variation of results on other traces (except the very short traces, where variance was naturally higher) was similar.
+Bit rate selection algorithms naturally show variation between runs on
+the same data because probe rates are chosen randomly. The variations
+were small, though. On ten runs on the \texttt{clear\_1.dat} trace,
+the average of Minstrel over 10 runs was 7.52 Mbps, with a maximum
+variation of +/- 0.12 Mbps. SampleRate was similarly stable, it
+averaged 5.284 Mbps over 10 runs, with a maximum variation of about
++/- 0.3 Mbps. The variation of results on other traces (except the
+very short traces, where variance was naturally higher) was similar.
 
-The histograms of which rates the algorithms decided to send at provided some useful insight into their behavior.
-Table~\ref{table:2} is a histogram of a 35 second corner trace.
-In this situation Minstrel outperforms SampleRate.
-The constant rate throughputs demonstrate that 11, 12 and 18 Mbps were the optimal rates to send at, and that no packets were sucessful at any higher rates.
-SampleRate does not use these rates except for probe packets, which happens once every ten seconds.
-Minstrel, since it does not terminate transmission after four successive failures, spends a lot of time sending probe packets at these high rates, which are never successful.
-The only reason low rates have fewer probe packets is because Minstrel explicitly places them second in the retry chain, and because the main rate usually succeeds.
-Despite the fact that SampleRate spends less time probing high rates, it is less successful.
-This is because Minstrel send the most packets at the three most successful bitrates.
-SampleRate, however, spends too much time sampling low bit rates.
-Additionally, it spends a lot of time sending at 5.5 Mbps.
-TODO: why does SampleRate spend time at 5.5 Mbps???
-We suspect that this might be because the equation SampleRate uses to estimate average transmission time needs to be tweaked.
+The histograms of which rates the algorithms decided to send at
+provided some useful insight into their behavior. Table~\ref{table:2}
+is a histogram of a 35 second corner trace. In this situation Minstrel
+outperforms SampleRate. The constant rate throughputs demonstrate that
+11, 12 and 18 Mbps were the optimal rates to send at, and that no
+packets were sucessful at any higher rates. SampleRate does not use
+these rates except for probe packets, which happens once every ten
+seconds. Minstrel, since it does not terminate transmission after four
+successive failures, spends a lot of time sending probe packets at
+these high rates, which are never successful. The only reason low
+rates have fewer probe packets is because Minstrel explicitly places
+them second in the retry chain, and because the main rate usually
+succeeds. Despite the fact that SampleRate spends less time probing
+high rates, it is less successful. This is because Minstrel send the
+most packets at the three most successful bitrates. SampleRate,
+however, spends too much time sampling low bit rates. Additionally, it
+spends a lot of time sending at 5.5 Mbps. TODO: why does SampleRate
+spend time at 5.5 Mbps??? We suspect that this might be because the
+equation SampleRate uses to estimate average transmission time needs
+to be tweaked.
 
-Our examination of the histograms showed us that Minstrel spends a lot of time sampling at high rates.
-In the specific case we examined, they were all probe packets.
-But sometimes Minstrel would choose 54Mbps as the rate with the best throughput, which made little sense.
-A closer investigation revealed that Minstrel would sometimes get lucky with a successful probe at 54Mbps.
-The probability of success would be 1, so then 54Mbps would be chosen as the rate with the best throughput.
-This is unfortunate, but the real harm comes in the next window: packets would now be send at 54Mbps, and nearly all would fail.
-However, since these hundreds of failed packets weight equally with the single successful probe packet, Minstrel still considers 54 Mbps the best rate.
-Thus the probability of success might drop from 1 to about 0.70, which is still high enough for the throughput equation to rank 54 Mbps highly.
-Fundamentally, the problem is that the EWMA does not account the second block having hundreds more packets than the first.
-In the next section we will talk about Minproved, an improved version of Minstrel where we attempt to fix this problem.
+Our examination of the histograms showed us that Minstrel spends a lot
+of time sampling at high rates. In the specific case we examined, they
+were all probe packets. But sometimes Minstrel would choose 54Mbps as
+the rate with the best throughput, which made little sense. A closer
+investigation revealed that Minstrel would sometimes get lucky with a
+successful probe at 54Mbps. The probability of success would be 1, so
+then 54Mbps would be chosen as the rate with the best throughput. This
+is unfortunate, but the real harm comes in the next window: packets
+would now be send at 54Mbps, and nearly all would fail.  However,
+since these hundreds of failed packets weight equally with the single
+successful probe packet, Minstrel still considers 54 Mbps the best
+rate. Thus the probability of success might drop from 1 to about 0.70,
+which is still high enough for the throughput equation to rank 54 Mbps
+highly. Fundamentally, the problem is that the EWMA does not account
+the second block having hundreds more packets than the first.  In the
+next section we will talk about Minproved, an improved version of
+Minstrel where we attempt to fix this problem.
 
@@
 
-In Section~\ref{sec:analysis}, we noted that the EWMA used by Minstrel causes problems when probe packets succeed.
-A natural next step was to modify Minstrel to avoid this problem, by weighing each 100ms block of statistics in proportion to how many packets it contains.
-To preserve the exponential weighing of statistics, we chose to weigh each 100ms block based on the number of packets sent, compared to the those sent in the average block.
-Parameters were chosen so that if all blocks were of average size, our ``balanced'' EWMA would behave identically to a normal EWMA.
+In Section~\ref{sec:analysis}, we noted that the EWMA used by Minstrel
+causes problems when probe packets succeed.  A natural next step was
+to modify Minstrel to avoid this problem, by weighing each 100ms block
+of statistics in proportion to how many packets it contains.  To
+preserve the exponential weighing of statistics, we chose to weigh
+each 100ms block based on the number of packets sent, compared to the
+those sent in the average block.  Parameters were chosen so that if
+all blocks were of average size, our ``balanced'' EWMA would behave
+identically to a normal EWMA.
 
 To achieve, this weighing, we first recast the usual EWMA algorithm.
-The usual algorithm computes $$p \gets \alpha p + (1 - \alpha) \frac{n_i}{d_i}),$$ where $p$ is the EWMA probability, $\alpha$ is a weight factor equal to $0.75$ in Minstrel, and $n_i$ and $d_i$ are the number of successful and total packets in the 100ms window.
-We next write this as $$p \gets \frac{\beta p + \frac{n_i}{d_i}}{\beta + 1},$$ where $\beta = \alpha / (1 - \alpha)$.
-We can now change the factor in front of $n_i / d_i$ to account for the total number of packets sent.
-In our case, we chose to compute $$p \gets \frac{\beta p + w_i \frac{n_i}{d_i}}{\beta + w_i},$$ where the weight $w_i$ is $d_i / (d / b)$, with $d$ the total number of packets ever send and $b$ the total number of 100ms windows ever seen.
-Then $d / b$ is the average number of packets per window and $d_i / (d / b)$ is the ratio between the current window and the average window.
-Note that if the current window is of average size, this is exactly equal to the usual EWMA result.
-Finally, we can simplify the above equation to $$S p \gets \frac{\beta \frac{d}{b} p + S n_i}{\beta \frac{d}{b} + d_i},$$ where $S$ is any constant.
-This equation is usable in fixed-precision arithmetic, making it simple to implement in the Linux kernel (which prefers to avoid floating point computations for portability reasons).
-Because packets are high rates rarely succeed, the average window size is a few packets; thus, the main qualitative difference between our balanced EWMA and the original Minstrel EWMA is that non-probe packets at probe bit-rates are weighed extraordinarily heavily, and this fact allows us to avoid multiple failing 100ms windows.
+The usual algorithm computes $$p \gets \alpha p + (1 - \alpha)
+\frac{n_i}{d_i}),$$ where $p$ is the EWMA probability, $\alpha$ is a
+weight factor equal to $0.75$ in Minstrel, and $n_i$ and $d_i$ are the
+number of successful and total packets in the 100ms window.  We next
+write this as $$p \gets \frac{\beta p + \frac{n_i}{d_i}}{\beta + 1},$$
+where $\beta = \alpha / (1 - \alpha)$.  We can now change the factor
+in front of $n_i / d_i$ to account for the total number of packets
+sent.  In our case, we chose to compute $$p \gets \frac{\beta p + w_i
+  \frac{n_i}{d_i}}{\beta + w_i},$$ where the weight $w_i$ is $d_i / (d
+/ b)$, with $d$ the total number of packets ever send and $b$ the
+total number of 100ms windows ever seen.  Then $d / b$ is the average
+number of packets per window and $d_i / (d / b)$ is the ratio between
+the current window and the average window.  Note that if the current
+window is of average size, this is exactly equal to the usual EWMA
+result.  Finally, we can simplify the above equation to $$S p \gets
+\frac{\beta \frac{d}{b} p + S n_i}{\beta \frac{d}{b} + d_i},$$ where
+$S$ is any constant.  This equation is usable in fixed-precision
+arithmetic, making it simple to implement in the Linux kernel (which
+prefers to avoid floating point computations for portability reasons).
+Because packets are high rates rarely succeed, the average window size
+is a few packets; thus, the main qualitative difference between our
+balanced EWMA and the original Minstrel EWMA is that non-probe packets
+at probe bit-rates are weighed extraordinarily heavily, and this fact
+allows us to avoid multiple failing 100ms windows.
 
-We tested a modified Minstrel which used this more balanced EWMA algorithm on the same traces as Minstrel and SampleRate, and found a consistent improvement of approximately 1 Mbps over the normal Minstrel algorithm.
+We tested a modified Minstrel which used this more balanced EWMA
+algorithm on the same traces as Minstrel and SampleRate, and found a
+consistent improvement of approximately 1 Mbps over the normal
+Minstrel algorithm.
 
-We noted another curious fact about Minstrel: the kernel implementation of Minstrel samples rates very frequently.
-This is due to a flag the kernel uses to track which probe bit-rates were actually sampled.
-In cases where the probe packets are at a lower bit-rate than the current bit-rate, the probe packets might never be sent if the current bit-rate succeeds.
-However, the flagging seems to misbehave in certain cases, causing the extremely aggressive probing we see in some traces.
-For example, in one of the traces where the client was around a corner from the access point, Minstrel sent almost half of its packets at probe frequencies.
-We modified Minstrel to be less aggressive with probe packets, decreasing the sampling frequency to approximately 10\%.
-This lead to more improvement, especially on noisy traces such as the cases where the client was around a corner or moving.
+We noted another curious fact about Minstrel: the kernel
+implementation of Minstrel samples rates very frequently.  This is due
+to a flag the kernel uses to track which probe bit-rates were actually
+sampled.  In cases where the probe packets are at a lower bit-rate
+than the current bit-rate, the probe packets might never be sent if
+the current bit-rate succeeds.  However, the flagging seems to
+misbehave in certain cases, causing the extremely aggressive probing
+we see in some traces.  For example, in one of the traces where the
+client was around a corner from the access point, Minstrel sent almost
+half of its packets at probe frequencies.  We modified Minstrel to be
+less aggressive with probe packets, decreasing the sampling frequency
+to approximately 10\%.  This lead to more improvement, especially on
+noisy traces such as the cases where the client was around a corner or
+moving.
 
-Overall, our improved Minstrel, which we call Minproved, often out-performs Minstrel by multiple Mbps and never doing any worse; and usually surpassing the best constant bit rate.
-In only one case did SampleRate achieve higher throughput than Minproved, and in this case both surpassed the best constant bit-rate.
-In cases where Minstrel out-performed SampleRate, the same behavior was true of Minproved.
-In most cases, Minproved was $30\%$ faster than Minstrel, and in some the improvement as great as $60\%$.
+Overall, our improved Minstrel, which we call Minproved, often
+out-performs Minstrel by multiple Mbps and never doing any worse; and
+usually surpassing the best constant bit rate.  In only one case did
+SampleRate achieve higher throughput than Minproved, and in this case
+both surpassed the best constant bit-rate.  In cases where Minstrel
+out-performed SampleRate, the same behavior was true of Minproved.  In
+most cases, Minproved was $30\%$ faster than Minstrel, and in some the
+improvement as great as $60\%$.
 
-However, Minproved still makes many poor choices.
-While a single successful probe packet does not cause many hundreds of milliseconds of failed packets, as it does in Minstrel, the single successful packet can still cause one window's worth of failed packets.
-Instead, it would be best if a successful probe packet at a rarely-successful bit-rate was treated more carefully.
-One can imagine an algorithm, for example, which would treat a single successful probe packet as a reason to send a dozen
+However, Minproved still makes many poor choices.  While a single
+successful probe packet does not cause many hundreds of milliseconds
+of failed packets, as it does in Minstrel, the single successful
+packet can still cause one window's worth of failed packets.  Instead,
+it would be best if a successful probe packet at a rarely-successful
+bit-rate was treated more carefully.  One can imagine an algorithm,
+for example, which would treat a single successful probe packet as a
+reason to send a dozen
 
@@
     \caption{Histograms of the number of packets sent at each rate.
-The trace was \texttt{corner\_1.dat} and was recorded around the corner from the access point.
-The best fixed rate for this trace was 11 Mbps, which achieved a throughput of 9.627 Mbps.
-Minproved achieves a higher throughput than the best fixed rate.}
+      The trace was \texttt{corner\_1.dat} and was recorded around the
+      corner from the access point. The best fixed rate for this trace
+      was 11 Mbps, which achieved a throughput of 9.627
+      Mbps. Minproved achieves a higher throughput than the best fixed
+      rate.}
 \label{table:2}
@@
 
-This repository contains the modified ath9k driver used to collect the traces, the traces we collected, as well as the Python simulation framework.
+This repository contains the modified ath9k driver used to collect the
+traces, the traces we collected, as well as the Python simulation
+framework.
 
@@
 
-Very little analysis has been done of the performance of bit-rate selection algorithms on 802.11n networks.
-802.11n introduces many new rates, as well as multiple-input/multiple output, or MIMO, capabilities.
-There is an 802.11n version of Minstrel, but there is no implementation of SampleRate.
+Very little analysis has been done of the performance of bit-rate
+selection algorithms on 802.11n networks. 802.11n introduces many new
+rates, as well as multiple-input/multiple output, or MIMO,
+capabilities. There is an 802.11n version of Minstrel, but there is no
+implementation of SampleRate.
 
-We implemented SampleRate as outlined in the thesis, but the MadWifi implementation has a few key differences from the thesis.
-We suspect that the MadWifi use of an EWMA instead of a window has more accurate rate statistics, which will potentially lead to better throughput.
-We did not have the time to implement this alternative version, but it is worth further examination.
+We implemented SampleRate as outlined in the thesis, but the MadWifi
+implementation has a few key differences from the thesis. We suspect
+that the MadWifi use of an EWMA instead of a window has more accurate
+rate statistics, which will potentially lead to better throughput. We
+did not have the time to implement this alternative version, but it is
+worth further examination.
 
-Hari Balakrishnan suggested that we use a broadcast approach when collecting our traces, instead of link layer acknowledgments.
-This would entail having multiple computers listening at the broadcast address, and compiling successes and failures from that information.
-This approach would free us from certain idiosyncrasies of reading data from the driver, and lead to possibly more accurate traces.
-We were unable to implement this due to a lack of time and ability to acquire compatible hardware.
+Hari Balakrishnan suggested that we use a broadcast approach when
+collecting our traces, instead of link layer acknowledgments. This
+would entail having multiple computers listening at the broadcast
+address, and compiling successes and failures from that
+information. This approach would free us from certain idiosyncrasies
+of reading data from the driver, and lead to possibly more accurate
+traces. We were unable to implement this due to a lack of time and
+ability to acquire compatible hardware.
 
-Finally, we are considering submitting our improvements to the Minstrel algorithm, namely the BEWMA, as a kernel patch. Currently is implemented in Python, so we would have to port the code into C.
+Finally, we are considering submitting our improvements to the
+Minstrel algorithm, namely the BEWMA, as a kernel patch. Currently is
+implemented in Python, so we would have to port the code into C.
 
@@
 
-Minstrel and SampleRate perform similarly.
-The results suggest that SampleRate tends to do slightly better, except in very lossy links such as the client being far awat from the AP or actively moving.
-Minproved, our improved version of Minstrel, solidly outperforms both Minstrel and SampleRate.
+Minstrel and SampleRate perform similarly. The results suggest that
+SampleRate tends to do slightly better, except in very lossy links
+such as the client being far awat from the AP or actively
+moving. Minproved, our improved version of Minstrel, solidly
+outperforms both Minstrel and SampleRate.
 
-It seems like the abandonment of SampleRate in the Linux kernel may have been a bit hasty, since it seems to perform similarly to Minstrel and it is a much simpler algorithm and thus less bug prone.
-We are not sure if there are adjustments to SampleRate that will bring it's performance near Minproved, though.
+It seems like the abandonment of SampleRate in the Linux kernel may
+have been a bit hasty, since it seems to perform similarly to Minstrel
+and it is a much simpler algorithm and thus less bug prone. We are not
+sure if there are adjustments to SampleRate that will bring it's
+performance near Minproved, though.
 
 \section{Acknowledgments}
-We would like to thank Jonathan Perry and Hari Balakrishnan for their valuable guidance, as well as Derek Smithies for answering our questions about Minstrel.
+
+We would like to thank Jonathan Perry and Hari Balakrishnan for their
+valuable guidance, as well as Derek Smithies for answering our
+questions about Minstrel.
 

### 2013-05-16 paper/paper.tex
tweak

@@
 
-We did notice that Minstrel occasionally has very poor performance on
-traces 1 seconds or shorter---sometimes as low as 2 Mbps. This is
-because it does not stop sending after four successive failures, but
-must wait until the end of the 100ms window for statistics to be
-compiled. Thus it wastes a lot of time sending at non-optimal rates;
-and a short trace has only a few 100ms windows. This behavior is not
-evident in long traces, however, as Minstrel has had time to both
-smooth out bad windows and collect more statistics (making it less
-likely to choose a bad rate).
-
 Bit rate selection algorithms naturally show variation between runs on
@@
 We implemented SampleRate as outlined in the thesis, but the MadWifi
-implementation has a few key differences from the thesis. We suspect
-that the MadWifi use of an EWMA instead of a window has more accurate
+implementation has a few key differences from the thesis. MadWifi 
+use of an EWMA instead of a window has more accurate
 rate statistics, which will potentially lead to better throughput. We

### 2013-05-16 paper/paper.tex
TeX cleanup

@@
-% TEMPLATE for Usenix papers, specifically to meet requirements of
-%  USENIX '05
-% originally a template for producing IEEE-format articles using LaTeX.
+% TEMPLATE for Usenix papers originally for producing IEEE-format articles
 % written by Matthew Ward, CS Department, Worcester Polytechnic Institute.
-% adapted by David Beazley for his excellent SWIG paper in Proceedings,
-%   Tcl 96
-% turned into a smartass generic template by De Clarke, with thanks to
-%   both the above pioneers
-% use at your own risk.  Complaints to /dev/null.
-% make it two column with no page numbering, default is 10 point
+% adapted by David Beazley for his paper in Proceedings, Tcl 96
+% turned into a smartass generic template by De Clarke
 
-% Munged by Fred Douglis <douglis@research.att.com> 10/97 to separate
-% the .sty file from the LaTeX source template, so that people can
-% more easily include the .sty file into an existing document.  Also
-% changed to more closely follow the style guidelines as represented
-% by the Word sample file.
+\documentclass[twocolumn,10pt]{article}
+\usepackage[margin=1in]{geometry}
+\usepackage{graphicx,hyperref}
+\usepackage{subcaption}
 
-% Note that since 2010, USENIX does not require endnotes. If you want
-% foot of page notes, don't include the endnotes package in the 
-% usepackage command, below.
-
-% This version uses the latex2e styles, not the very ancient 2.09 stuff.
-\documentclass[letterpaper,twocolumn,10pt]{article}
-\usepackage{epsfig,graphicx,usenix,fullpage, hyperref}
-
-%\usepackage{endnotes}
-\begin{document}
-
-%don't want date printed
-
-%make title bold and 14 pt font (Latex default is non-bold, 16 pt)
-\title{\Large \bf 6.829 Final Project: A comparison of bit-rate selection algorithms}
-%for single author (just remove % characters)
-\author{
-{\rm Colleen Josephson}\\
-cjoseph@mit.edu
-\and
-{\rm Pavel Panchekha}\\
-pavpan@mit.edu
-% copy the following lines to add more authors
-% \and
-% {\rm Name}\\
-%Name Institution
-} % end author
-
+\title{6.829 Final Project: A comparison of bit-rate selection algorithms}
+\author{Colleen Josephson \\ \texttt{cjoseph@mit.edu}
+  \and Pavel Panchekha \\ \texttt{pavpan@mit.edu}}
 \date{May 2013}
 
+\begin{document}
 \maketitle
 
-% Use the following at camera-ready time to suppress page numbers.
-% Comment it out when you first submit the paper for review.
-%\thispagestyle{empty}
+\begin{abstract}
+We compare the performance SampleRate and Minstrel, two popular
+bit-rate selection algorithms that have widespread real-world usage.
+We use a trace-based approach to avoid kernel programming and consider
+Python implementations of both SampleRate and Minstrel.  We test both
+algorithms in multiple real-world scenarios, including scenarios with
+mobile clients and noisy environments, to highlight differences
+between the two algorithms.  We also introduce improvements to the
+Minstrel algorithm that allow for significant gains in throughput.
+\end{abstract}
 
-\subsection*{Abstract}
-In this paper we compare the performance SampleRate and Minstrel, two
-popular bit-rate selection algorithms that have widespread real-world
-usage on the Linux operating system. We use a novel trace-based
-approach that allows us avoid in-depth kernel programming and write
-simple Python scripts instead. We also introduce improvements to the
-Minstrel algorithm that allow for significant gains in throughput.
 \section{Introduction}
 
-One of the key ways that wireless networks differ from wired is that
-wireless networks have varying link rates. Link conditions vary with
-time due to interference from other devices, changing network
-geometry, and mobile clients. An optimal rate at some time $t$ may be
+One of the key ways that wireless networks differ from wired links is
+that wireless networks have varying link rates.  Link conditions vary
+with time due to interference from other devices, changing network
+geometry, and mobile clients.  An optimal rate at one time may be
 different from the optimal rate just 30 seconds earlier.  A good
@@
 conditions.  If the chosen bit-rate is too slow, then the throughput
-will be unnecessarily low. If the rate is too high, then failures will
+will be unnecessarily low; if the rate is too high, failures will
 be very frequent and throughput will again suffer.
@@
 returns a the bit-rate to send a packet at,
-\texttt{ProcessFeedback()}, which updates the statistics for a bitrate
+\texttt{ProcessFeedback()}, which updates the statistics for a bit-rate
 after an attempted packet send, and \texttt{RemoveStaleResults()},
@@
 Our initial plan was to use the algorithms as implemented in the
-MadWifi drivers for the Atheros chipset. However, MadWifi is
+MadWifi drivers for the Atheros chip.  However, MadWifi is
 deprecated, and only runs on now-ancient network cards.  In addition to
@@
 
-\begin{figure}[hbt]
+\begin{figure}[htbp]
   \centering
-  \includegraphics[width=3.5in]{bsmt13floorplan.png}
+  \includegraphics[width=\linewidth]{bsmt13floorplan.png}
   \caption{We collected most traces at a remote access point in the basement of Building 13 at MIT.}
@@
 
-In Figure~\ref{figure:2} we show the performance of constant bitrates
+\begin{figure*}[htbp]
+  \centering
+  \begin{subfigure}{.45\linewidth}
+    \includegraphics[width=\linewidth]{constant.png}
+    \caption{Throughput of using a constant 802.11b/g bit-rate in
+      clear line-of-sight from the access point.}
+    \label{figure:2}
+  \end{subfigure}\hfill
+  \begin{subfigure}{.45\linewidth}
+    \includegraphics[width=\linewidth]{MinVSSam1.png}
+    \caption{Throughput of Minstrel and SampleRate in various
+      scenarios.}
+    \label{figure:3}
+  \end{subfigure}
+  \caption{Comparing Minstrel and Samplerate}
+\end{figure*}
+
+In Figure~\ref{figure:2} we show the performance of constant bit-rates
 on the \texttt{clear\_1.dat} trace, which is a 35 second trace in
@@
 which achieved an average throughput of about 15 Mbps.  The next
-highest bitrate, 24 Mbps, shows a large plummet in performance, down
+highest bit-rate, 24 Mbps, shows a large plummet in performance, down
 to a throughput of 1.17 Mbps.  This steep drop is charateristic, and is
@@
 Bit rate selection algorithms strive to have a throughput near or
-greater than the optimal constant rate. For this \texttt{clear\_1.dat}
-trace, SampleRate clearly meets this goal, achieving an average
-throughput more than 1 Mbps higher than the optimal constant rate. In
-noisier situations, though, Minstrel tends to be closer to the optimal
-constant rate.
+greater than the optimal constant rate.  For this
+\texttt{clear\_1.dat} trace, SampleRate clearly meets this goal,
+achieving an average throughput more than 1 Mbps higher than the
+optimal constant rate.  In noisier situations, though, Minstrel tends
+to be closer to the optimal constant rate.
 
@@
 provided some useful insight into their behavior.  Table~\ref{table:2}
-is a histogram of a 35 second corner trace. In this situation Minstrel
-outperforms SampleRate. The constant rate throughputs demonstrate that
-11, 12 and 18 Mbps were the optimal rates to send at, and that no
-packets were sucessful at any higher rates. SampleRate does not use
-these rates except for probe packets, which happens once every ten
-seconds. Minstrel, since it does not terminate transmission after four
-successive failures, spends a lot of time sending probe packets at
-these high rates, which are never successful. The only reason low
-rates have fewer probe packets is because Minstrel explicitly places
-them second in the retry chain, and because the main rate usually
-succeeds. Despite the fact that SampleRate spends less time probing
-high rates, it is less successful. This is because Minstrel send the
-most packets at the three most successful bitrates. SampleRate,
-however, spends too much time sampling low bit rates. Additionally, it
-spends a lot of time sending at 5.5 Mbps. TODO: why does SampleRate
-spend time at 5.5 Mbps??? We suspect that this might be because the
-equation SampleRate uses to estimate average transmission time needs
-to be tweaked.
+is a histogram of a 35 second corner trace.  In this situation
+Minstrel outperforms SampleRate.  The constant rate throughputs
+demonstrate that 11, 12 and 18 Mbps were the optimal rates to send at,
+and that no packets were sucessful at any higher rates.  SampleRate
+does not use these rates except for probe packets, which happens once
+every ten seconds.  Minstrel, since it does not terminate transmission
+after four successive failures, spends a lot of time sending probe
+packets at these high rates, which are never successful.  The only
+reason low rates have fewer probe packets is because Minstrel
+explicitly places them second in the retry chain, and because the main
+rate usually succeeds.  Despite the fact that SampleRate spends less
+time probing high rates, it is less successful.  This is because
+Minstrel send the most packets at the three most successful bit-rates.
+SampleRate, however, spends too much time sampling low bit rates.
+Additionally, it spends a lot of time sending at 5.5 Mbps.  We suspect
+that this might be because the equation SampleRate uses to estimate
+average transmission time needs to be tweaked.
+%   TODO: why does SampleRate spend time at 5.5 Mbps???
 
@@
 Minstrel where we attempt to fix this problem.
-
-\begin{figure}[htb]
-  \hspace{-1em}\includegraphics[width=3.4in]{constant.png}\vspace{-0em}
-  \caption{Throughput of using a constant 802.11b/g bitrate in clear line-of-sight from the access point.}
-\label{figure:2}
-\end{figure}
-
-\begin{figure}[htb]
-  \hspace{-.75em}\includegraphics[width=3.5in]{MinVSSam1.png}\vspace{-0.75em}
-  \caption{Throughput of Minstrel and SampleRate on various trace situations.}
-\label{figure:3}
-\end{figure}
 
@@
 
+\begin{table*}[htb]
+    \centering
+    \begin{tabular}{l|l|l|l|l}
+
+    \textbf{Rate in Mbps} & \textbf{SampleRate} &
+    \textbf{Minstrel} & \textbf{Minproved} &
+    \textbf{Constant Rate Throughput}\\ \hline
+
+    1   & 631  & 39    & 29    & 0.639 Mbps \\
+    2   & 863  & 82    & 20    & 1.425 Mbps \\
+    5.5 & 4814 & 110   & 805   & 4.406 Mbps \\
+    6   & 423  & 137   & 276   & 4.603 Mbps \\
+    9   & 553  & 327   & 57    & 4.630 Mbps \\
+    11  & 3326 & 4121  & 3356  & 9.627 Mbps \\
+    12  & 4649 & 9515  & 14285 & 9.444 Mbps \\
+    18  & 1154 & 13295 & 15795 & 8.458 Mbps \\
+    24  & 16   & 6909  & 1918  & 0 Mbps \\
+    36  & 16   & 6944  & 2030  & 0 Mbps \\
+    48  & 16   & 6888  & 2009  & 0 Mbps \\
+    54  & 16   & 7308  & 1953  & 0 Mbps \\ \hline
+
+    \textbf{Avg. Throughput:} & \emph{5.42 Mbps}  & \emph{7.578 Mbps} & \emph{11.107 Mbps} \\
+    \end{tabular}
+    \caption{Histograms of the number of packets sent at each rate.
+      The trace was \texttt{corner\_1.dat} and was recorded around the
+      corner from the access point.  The best fixed rate for this
+      trace was 11 Mbps, which achieved a throughput of 9.627 Mbps.
+      Minproved achieves a higher throughput than the best fixed
+      rate.}
+
+    \label{table:2}
+\end{table*}
+
+\begin{figure}[htb]
+  \includegraphics[width=\linewidth]{mnVSspVSmp1.png}
+  \caption{Throughput of Minproved as compared to SampleRate and
+    vanilla Minstrel.}
+\label{figure:4}
+\end{figure}
+
 Overall, our improved Minstrel, which we call Minproved, often
@@
 for example, which would treat a single successful probe packet as a
-reason to send a dozen
+reason to send a dozen more probe packets at that bit-rate, but does
+not yet commit to sending hundreds of packets at that bit-rate.
 
-\begin{table*}[htb]
-    \centering
-    \begin{tabular}{l|l|l|l|l}
-    \textbf{Rate in Mbps}     & \textbf{SampleRate} & \textbf{Minstrel}   & \textbf{Minproved}   & \textbf{Constant Rate Throughput}\\ \hline
-    1                & 631        & 39         & 29      & 0.639 Mbps    \\
-    2                & 863        & 82         & 20      & 1.425 Mbps    \\
-    5.5              & 4814       & 110        & 805     & 4.406 Mbps    \\
-    6                & 423       & 137        & 276     & 4.603 Mbps    \\
-    9                & 553         & 327        & 57      & 4.630 Mbps   \\
-    11               & 3326       & 4121       & 3356    & 9.627 Mbps   \\
-    12               & 4649       & 9515       & 14285   & 9.444 Mbps   \\
-    18               & 1154         & 13295      & 15795   & 8.458 Mbps    \\
-    24               & 16         & 6909       & 1918    & 0 Mbps   \\
-    36               & 16         & 6944       & 2030    & 0 Mbps   \\
-    48               & 16         & 6888       & 2009    & 0 Mbps   \\
-    54               & 16         & 7308       & 1953    & 0 Mbps   \\ \hline
-    \textbf{Avg. Throughput:} & \emph{5.42 Mbps}  & \emph{7.578 Mbps} & \emph{11.107 Mbps} \\
-    \end{tabular}
-    \caption{Histograms of the number of packets sent at each rate.
-      The trace was \texttt{corner\_1.dat} and was recorded around the
-      corner from the access point. The best fixed rate for this trace
-      was 11 Mbps, which achieved a throughput of 9.627
-      Mbps. Minproved achieves a higher throughput than the best fixed
-      rate.}
-\label{table:2}
-\end{table*}
-
-\begin{figure}[htb]
-  \hspace{-2em}\includegraphics[width=3.5in]{mnVSspVSmp1.png}\vspace{-0em}
-  \caption{Throughput of Minproved as compared to SampleRate and vanilla Minstrel.}
-\label{figure:4}
-\end{figure}
+However, these are only vague ideas, and we have no concrete
+implementations of such an algorithm.
 
@@
 would entail having multiple computers listening at the broadcast
-address, and compiling successes and failures from that
-information. This approach would free us from certain idiosyncrasies
-of reading data from the driver, and lead to possibly more accurate
-traces. We were unable to implement this due to a lack of time and
-ability to acquire compatible hardware.
+address, and compiling successes and failures from that information.
+This approach would free us from certain idiosyncrasies of reading
+data from the driver, and lead to possibly more accurate traces.  We
+were unable to implement this due to a lack of time and ability to
+acquire compatible hardware.
 
@@
 SampleRate tends to do slightly better, except in very lossy links
-such as the client being far awat from the AP or actively
-moving. Minproved, our improved version of Minstrel, solidly
-outperforms both Minstrel and SampleRate.
+such as the client being far awat from the AP or actively moving.
+Minproved, our improved version of Minstrel, solidly outperforms both
+Minstrel and SampleRate.
 
@@
 have been a bit hasty, since it seems to perform similarly to Minstrel
-and it is a much simpler algorithm and thus less bug prone. We are not
-sure if there are adjustments to SampleRate that will bring it's
+and it is a much simpler algorithm and thus less bug prone.  We are
+not sure if there are adjustments to SampleRate that will bring it's
 performance near Minproved, though.

### 2013-05-16 paper/paper.tex
More cleanup
Proofing

@@
 \usepackage[margin=1in]{geometry}
-\usepackage{graphicx,hyperref}
-\usepackage{subcaption}
+\usepackage{graphicx,subcaption,hyperref}
 
@@
 conditions.  If the chosen bit-rate is too slow, then the throughput
-will be unnecessarily low; if the rate is too high, failures will
-be very frequent and throughput will again suffer.
+will be unnecessarily low; if the rate is too high, failures will be
+very frequent and throughput will again suffer.
 
@@
 protocols measure the fraction of successfully received packets.  SNR
-protocols make decisions based on the estimated Signal to Noise
-ratio.  Cross-layer protocols use SoftPHY data from the physical
-layer.  The most commonly implemented protocols on today's networks are
-frame based, because SNR protocols perform poorly~\cite{samplerate},
-and cross-layer protocols cannot be deployed on current networks as
-they violate the network layering abstraction.
+protocols make decisions based on the estimated Signal to Noise ratio.
+Cross-layer protocols use SoftPHY data from the physical layer.  The
+most commonly implemented protocols on today's networks are frame
+based, because SNR protocols perform poorly~\cite{samplerate}, and
+cross-layer protocols cannot be deployed on current networks as they
+violate the network layering abstraction.
 
@@
 
-This paper will provide a general overview of SampleRate and
-Minstrel.  We then discuss our testing methodology.  We compare the
-performance of the two algorithms, and then discuss modifications we
-made to Minstrel and compare it's performance to the current kernel
+This paper will provide a general overview of SampleRate and Minstrel.
+We then discuss our testing methodology.  We compare the performance
+of the two algorithms, and then discuss modifications we made to
+Minstrel and compare it's performance to the current kernel
 implementation.
@@
 
+SampleRate, first introduced in \cite{samplerate}, is a bit rate
+selection algorithm which maintains estimates of average transmission
+time for each potential bit rate.To keep these estimates up to date,
+SampleRate periodically sends a packet at a randomly-selected bit
+rate, and uses the success or failure of this packet to update the
+average transmission time of that bit-rate.  SampleRate also abandons
+a bit rate after four successive failures at that rate.  All
+information about packets is maintained with a 10-second sliding window.
+
 SampleRate has three main functions: \texttt{ApplyRate()}, which
 returns a the bit-rate to send a packet at,
-\texttt{ProcessFeedback()}, which updates the statistics for a bit-rate
-after an attempted packet send, and \texttt{RemoveStaleResults()},
-which removes all statistics older than 10 seconds.
-
-SampleRate sends at the rate with the lowest average transmission
-time.  Every ten packets, it sends out a sample packet at a random rate
-that has a smaller predicted transmission time than the current
-rate.  SampleRate updates its transmission time estimates every ten
-seconds based on feedback from the card about which packets succeeded
-and at which rate they were sent.  If the current best rate experiences
-four successive failures, SampleRate stops using that rate.
+\texttt{ProcessFeedback()}, which updates the statistics for a
+bit-rate after an attempted packet send, and
+\texttt{RemoveStaleResults()}, which removes the effect of packets
+older than 10 seconds.
 
@@
 
-Minstrel utilizes the multi-rate retry chain (mrr), which is an array
-that tells the card which rates to try before reporting a failure.  The
-retry chain makes failures extremely unlikely.  Minstrel has a mrr of
-[\emph{highest throughput, next highest throughput, best probability,
-    lowest base rate}] for most packets.  In essence, Minstrel tells
+Minstrel, developed specifically for the Linux kernel, attempts to
+improve upon SampleRate to make it better suited for noisy or
+dynamic environments.  Unlike SampleRate, Minstrel tracks raw
+probabilities of successful transmission for each bit rate, computed
+based on an exponentially-weighted moving average of 100ms windows.
+
+Minstrel makes use the multi-rate retry chain (MRR), an array of bit rates
+and number of attempts to make at each bit rate, that tells the card
+which rates to try before reporting a failure.  The retry chain makes
+failures extremely unlikely and allows Minstrel to choose fallback
+rates when a packet does not succeed.  Minstrel sets the MRR to first
+try the rate with the highest throughput, then the next-highest
+throughput, then greatest probability of success, and finally attempts
+to send a packet at the lowest, base rate.  In essence, Minstrel tells
 the card to send at the highest throughput rates but to fall back to
-reliable rates upon failure.  The throughput and success probability
-are calculated for each rate every 100ms.  To keep an accurate
-estimate of throughput and probability for each rate, Minstrel sends
-sample frames ten percent of the time.  If the randomly chosen rate has
-a higher lossless throughput than the current optimal rate, the mrr is
-as follows: [\emph{random, highest throughput, best probability,
-    lowest base rate}].  If the random rate is slower than the optimal
-rate, it is moved further back in the retry chain: [\emph{best
-    throughput, random, best probability, lowest base rate}].  This
-ensures that Minstrel never samples rates worse than the current
-optimal rate unless the optimal rate experiences a failure.  See table
-~\ref{table:1} for an overview of the mrr's used in Minstrel.
+reliable rates upon failure.
 
-Minstrel's implementation is based on SampleRate's, so it also has
-\texttt{ApplyRate()} and \texttt{ProcessFeedback()}, as well as a
-\texttt{UpdateStats()} method that runs every
-100ms.  \texttt{UpdateStats()} is home to some of the key differences
-between Minstrel and SampleRate.  Instead of making decisions based on
-the average transmission time, Minstrel uses
-throughput: \begin{equation} T = \frac{p}{tx\_time}\end{equation}
-where $T$ is the throughput of some rate $r$, $p$ is the probability
-of successfully transmitting at $r$, and $tx\_time$ is the lossless
-transmission time at $r$.
+To keep an accurate estimate of throughput and probability for each
+rate, Minstrel sends sample frames ten percent of the time.  However,
+Minstrel tries to avoid sampling at a rate slower than the current
+best rate, if the current best rate is still successfully
+transmitting.  If the randomly chosen rate has a higher lossless
+throughput than the current optimal rate, the MRR first lists the
+sample rate, then the highest-throughput rate, then the
+best-probability rate, and finally the base rate.  On the other hand,
+if the random rate is slower than the optimal rate, the sample rate is
+placed lower in the retry chain: first, the best throughput rate is
+tried, and only if it fails does the sample rate then get attempted
+(followed, as usual, by the best probability rate and the base rate).
+This ensures that Minstrel never samples rates worse than the current
+optimal rate unless the optimal rate experiences a failure.  See
+Table~\ref{table:mrr} for an overview of the retry chains used in
+Minstrel.
+
+Minstrel implementation is based on SampleRate, so it also has
+\texttt{ApplyRate()} and \texttt{ProcessFeedback()} methods, as well
+as a \texttt{UpdateStats()} method that runs every 100ms.
+\texttt{UpdateStats()} is home to some of the key differences between
+Minstrel and SampleRate.  Instead of making decisions based on the
+average transmission time, Minstrel uses throughput, computed with $$T
+= \frac{p}{tx\_time}$$ where $T$ is the throughput of some rate $r$,
+$p$ is the probability of successfully transmitting at $r$, and
+$tx\_time$ is the computed lossless transmission time at rate $r$.
 
@@
 creates a weighted average that weighs recent data more heavily than
-old data.  The weighting for old data decreases exponentially.  An EWMA
-is beneficial because it ensures that a sudden degradation in link
-quality will create a rapid response in the probabilities due to the
-increased weight of recent data.  Minstrel will waste less time sending
-packets at a rate that no longer works, as compared to a simple
-probability calculation.
+old data, with the weighting for old data decreasing exponentially.
+An EWMA ensures that a sudden degradation in link quality will create
+a rapid response in the probabilities, making the EWMA-using Minstrel
+perform better in dynamic environments.  Minstrel will waste less time
+sending packets at a rate that no longer works, as compared to a
+simple sliding-window probability calculation.
 
@@
     \begin{tabular}[htb]{l|l|l|l}
-    \textbf{Try} & \textbf{Normal}               & \textbf{Random (slower than best)} & \textbf{Random (faster than best)} \\ \hline
+    \textbf{Try} & \textbf{Normal} &
+    \textbf{Random (slower than best)} &
+    \textbf{Random (faster than best)} \\ \hline
     1 & Best throughput      & Best throughput     & Random              \\
@@
     \caption{Multi-rate retry chains for Minstrel}
-\label{table:1}
+\label{table:mrr}
 \end{table*}
@@
 Our initial plan was to use the algorithms as implemented in the
-MadWifi drivers for the Atheros chip.  However, MadWifi is
-deprecated, and only runs on now-ancient network cards.  In addition to
-the unavailability of compatible hardware, MadWifi would not be an
-accurate reflection of contemporary users.  Both ath5k and ath9k, which
-were created to replace MadWifi, did not port the old bit-rate
-selection system from MadWifi and instead opted to use the Linux
-kernel's rate selection framework, which only implements Minstrel.
-Porting SampleRate over to the new drivers would not have been
-trivial, as there are vast differences in the assumed interfaces.
+MadWifi drivers for the Atheros chip.  However, MadWifi is deprecated,
+and only runs on now-ancient network cards.  In addition to the
+unavailability of compatible hardware, MadWifi would not be an
+accurate reflection of contemporary users.  Both ath5k and
+\texttt{ath9k}, which were created to replace MadWifi, did not port
+the old bit-rate selection system from MadWifi and instead opted to
+use the Linux kernel's rate selection framework, which only implements
+Minstrel.  Porting SampleRate over to the new drivers would not have
+been trivial, as there are vast differences in the assumed interfaces.
 
@@
 approach similar to that used in Sprout~\cite{sprout}.  We modified
-the ath9k driver to try bit rates uniformly and report packet
+the \texttt{ath9k} driver to try bit rates uniformly and report packet
 successes and failures, then analyze this data in user-space with
@@
 
-Traces were collected from an ath9k driver modified to sample
+Traces were collected from an \texttt{ath9k} driver modified to sample
 bit-rates uniformly (forgoing the usual bit rate selection algorithm)
@@
 thesis implementation computes average transmission times based on a
-10-second sliding window.  It would be worth looking at the performance
-of EWMA-based SampleRate in the future.  Our implementation,
-\texttt{samplerate.py} contains \texttt{ApplyRate()},
+10-second sliding window.  It would be worth looking at the
+performance of EWMA-based SampleRate in the future.  Our
+implementation, \texttt{samplerate.py} contains \texttt{ApplyRate()},
 \texttt{ProcessFeedback()} and \texttt{RemoveStaleResults()}, as well
@@
 uses a probe frequency of 10\%, a window size of 100ms, and a EWMA
-weighting of 0.75.  We use the same values.  Additionally, Minstrel also
-estimates $tx\_time$ using the \texttt{ieee80211\_frame\_duration()}
+weighting of 0.75.  We use the same values.  Additionally, Minstrel
+also estimates $tx\_time$ using the \verb|ieee80211_frame_duration()|
 method from the \texttt{util.c} file in \texttt{net/mac80211}.  We
@@
 Minstrel and SampleRate perform similarly, usually within 3 Mbps of
-each other.  In especially noisy situations, such as when the client is
-moving or when the client is around the corner from the access point,
-Minstrel consistently outperforms SampleRate.  Figure~\ref{figure:3}
-shows the results of a single run of both Minstrel and SampleRate over
-all the different types of traces we collected.  This is consisten with
-the literature---the authors of Minstrel noted that they focused on
-making Minstrel robust to poor conditions~\cite{minstrel}, which means
-losing potential througput gains in stable situations.  Conversely,
-John Bicket noted that SampleRate performs better in stable situations
-than noisy or highly varying conditions~\cite{samplerate}.  We were
-slightly surprised to see how significantly SampleRate outperformed
-Minstrel in certain situations, though, with throughput being up to
-$67\%$ higher.
+each other.  In especially noisy situations, such as when the client
+is moving or when the client is around the corner from the access
+point, Minstrel consistently outperforms SampleRate.
+Figure~\ref{figure:3} shows the results of a single run of both
+Minstrel and SampleRate over all the different types of traces we
+collected.  This is consisten with the literature---the authors of
+Minstrel noted that they focused on making Minstrel robust to poor
+conditions~\cite{minstrel}, which means losing potential througput
+gains in stable situations.  Conversely, John Bicket noted that
+SampleRate performs better in stable situations than noisy or highly
+varying conditions~\cite{samplerate}.  We were slightly surprised to
+see how significantly SampleRate outperformed Minstrel in certain
+situations, though, with throughput being up to $67\%$ higher.
 
@@
 highest bit-rate, 24 Mbps, shows a large plummet in performance, down
-to a throughput of 1.17 Mbps.  This steep drop is charateristic, and is
-also talked about in the SampleRate thesis~\cite{samplerate}.  It
+to a throughput of 1.17 Mbps.  This steep drop is charateristic, and
+is also talked about in the SampleRate thesis~\cite{samplerate}.  It
 demonstrates just how important bit rate selection is---the optimal
@@
 Bit rate selection algorithms strive to have a throughput near or
-greater than the optimal constant rate.  For this
-\texttt{clear\_1.dat} trace, SampleRate clearly meets this goal,
-achieving an average throughput more than 1 Mbps higher than the
-optimal constant rate.  In noisier situations, though, Minstrel tends
-to be closer to the optimal constant rate.
+greater than the optimal constant rate.  For this \verb|clear_1.dat|
+trace, SampleRate clearly meets this goal, achieving an average
+throughput more than 1 Mbps higher than the optimal constant rate.  In
+noisier situations, though, Minstrel tends to be closer to the optimal
+constant rate.
 
@@
 
-\subsection{Availability}
+\section{Availability}
+
 All of our code and collected trace data are available on GitHub:
@@
 
-This repository contains the modified ath9k driver used to collect the
-traces, the traces we collected, as well as the Python simulation
-framework.
+This repository contains the modified \texttt{ath9k} driver used to
+collect the traces, the traces we collected, as well as the Python
+simulation framework.
 
@@
 selection algorithms on 802.11n networks.  802.11n introduces many new
-rates, as well as multiple-input/multiple output, or MIMO,
-capabilities.  There is an 802.11n version of Minstrel, but there is no
+rates, as well as multiple-input/multiple output (MIMO) capabilities.
+There is an 802.11n version of Minstrel, but there is no similar
 implementation of SampleRate.
@@
 We implemented SampleRate as outlined in the thesis, but the MadWifi
-implementation has a few key differences from the thesis. MadWifi 
-use of an EWMA instead of a window has more accurate
-rate statistics, which will potentially lead to better throughput. We
-did not have the time to implement this alternative version, but it is
-worth further examination.
+implementation has a few key differences from the thesis.  MadWifi use
+of an EWMA instead of a window has more accurate rate statistics,
+which will potentially lead to better throughput.  We did not have the
+time to implement this alternative version, but it is worth further
+examination.
 
-Hari Balakrishnan suggested that we use a broadcast approach when
-collecting our traces, instead of link layer acknowledgments.  This
-would entail having multiple computers listening at the broadcast
-address, and compiling successes and failures from that information.
-This approach would free us from certain idiosyncrasies of reading
-data from the driver, and lead to possibly more accurate traces.  We
-were unable to implement this due to a lack of time and ability to
-acquire compatible hardware.
+Hari Balakrishnan suggested that we use 802.11 broadcasts to collect
+our traces, instead of link layer acknowledgments.  We would have
+multiple computers listening at the broadcast address, and recording
+all received packets.  If packets were sent over UDP, each computer's
+record of which packets it received would allow us to compute the
+success probability for each packet for each rate.  This approach
+would free us from certain idiosyncrasies of reading data from the
+driver, and lead to possibly more accurate traces, as well as the
+ability to test multiple scenarios at once.  We were unable to
+implement this due to a lack of time and inability to acquire
+compatible hardware.
 
 Finally, we are considering submitting our improvements to the
-Minstrel algorithm, namely the BEWMA, as a kernel patch.  Currently is
-implemented in Python, so we would have to port the code into C.
+Minstrel algorithm, namely the balanced EWMA, as a kernel patch.
+Currently our changes implemented in Python, so we would have to port
+our improvements to C.
 
@@
 Minstrel and SampleRate perform similarly.  The results suggest that
-SampleRate tends to do slightly better, except in very lossy links
-such as the client being far awat from the AP or actively moving.
-Minproved, our improved version of Minstrel, solidly outperforms both
-Minstrel and SampleRate.
+SampleRate tends to perform slightly better, except in very lossy
+links such as the client being far awat from the AP or actively
+moving.  Minproved, our improved version of Minstrel, solidly
+outperforms both Minstrel and SampleRate.
 
-It seems like the abandonment of SampleRate in the Linux kernel may
-have been a bit hasty, since it seems to perform similarly to Minstrel
-and it is a much simpler algorithm and thus less bug prone.  We are
-not sure if there are adjustments to SampleRate that will bring it's
-performance near Minproved, though.
+It seems that the replacement of SampleRate in the Linux kernel may
+have been a hasty, since it seems to perform similarly to Minstrel and
+it is a much simpler algorithm and thus simpler to adapt to changing
+wireless standards.  We are not sure if adjustments to SampleRate can
+bring its performance to match that of Minproved, however.
 
@@
 
-{\footnotesize \bibliographystyle{acm}
-\bibliography{paper}}
+\bibliographystyle{acm}
+\bibliography{paper}
 

### 2013-05-17 paper/paper.tex
Added paragraph about what bit rates are.

@@
 
-\title{6.829 Final Project: A comparison of bit-rate selection algorithms}
+\title{6.829 Final Project: A comparison of bit rate selection algorithms}
 \author{Colleen Josephson \\ \texttt{cjoseph@mit.edu}
@@
 We compare the performance SampleRate and Minstrel, two popular
-bit-rate selection algorithms that have widespread real-world usage.
+bit rate selection algorithms that have widespread real-world usage.
 We use a trace-based approach to avoid kernel programming and consider
@@
 different from the optimal rate just 30 seconds earlier.  A good
-bit-rate selection algorithm has to detect and adapt to these
-conditions.  If the chosen bit-rate is too slow, then the throughput
+bit rate selection algorithm has to detect and adapt to these
+conditions.  If the chosen bit rate is too slow, then the throughput
 will be unnecessarily low; if the rate is too high, failures will be
@@
 
-There are three main classes of bit-rate adaptation protocols:
+The wireless 802.11 standard includes a variety of different bit rates
+to account for a range of possible network speeds.  Later standard
+versions introduce more bit rates; we focus on bit rates available in
+802.11b and 802.11g, since 802.11n networks and hardware were not
+available for study.  Table~\ref{tbl:bitrates} lists the bit rates
+available in the wireless networks studied.  The job of a bit rate
+selection protocol is to choose a bit rate, from among the available
+rates, for each packet transmission.  Due to the variability and
+non-determinism of packet transmission, this is a challenging task.
+
+\begin{table}[bthp]
+  \begin{tabular}{r|l|l}
+    Throughput & Transmission & Modulation \\\hline
+
+    1   Mbps & dsss & bpsk \\
+    2   Mbps & dsss & qpsk \\
+    5.5 Mbps & dsss & cck \\
+    11  Mbps & dsss & cck \\\hline
+
+    6   Mbps & ofdm & bpsk \\
+    9   Mbps & ofdm & bpsk \\
+    12  Mbps & ofdm & qpsk \\
+    18  Mbps & ofdm & qpsk \\
+    24  Mbps & ofdm & qam-16 \\
+    36  Mbps & ofdm & qam-16 \\
+    48  Mbps & ofdm & qam-64 \\
+    54  Mbps & ofdm & qam-64 \\
+  \end{tabular}
+  \caption{Bit rates available in 802.11b and 802.11g.  The first
+    block, of four bit rates, contains bit rates introduced in
+    802.11b, while the last eight bit rates were introduced in
+    802.11g.}
+  \label{tbl:bitrates}
+\end{table}
+
+There are three main classes of bit rate selection protocols:
 frame-based, SNR-based, and cross-layer protocols.  Frame-based
@@
 protocols make decisions based on the estimated Signal to Noise ratio.
-Cross-layer protocols use SoftPHY data from the physical layer.  The
-most commonly implemented protocols on today's networks are frame
-based, because SNR protocols perform poorly~\cite{samplerate}, and
-cross-layer protocols cannot be deployed on current networks as they
-violate the network layering abstraction.
+Cross-layer protocols use SoftPHY data from the physical layer to make
+better estimates of bit rate properties.  The most commonly
+implemented protocols on today's networks are frame based; SNR
+protocols perform poorly~\cite{samplerate} and cross-layer protocols
+are hard to deploy due to their violation the standard network
+layering abstraction.  Due to their ubiquity, we analyze frame-based
+protocols.
 
@@
 Minstrel.  Both were implemented in the MadWifi drivers for Linux
-wireless, and today Minstrel is the default bit-rate selection
-algorithm for all wireless drivers on Linux.  Our research used a
-trace-based approach to analyze the performance of these two
-algorithms.  We also created a modified version of Minstrel that
-provides significant throughput gains over the vanilla Minstrel
-implementation.
+wireless, and today Minstrel is the default bit rate selection
+algorithm for all wireless drivers on Linux.  We use a trace-based
+approach to analyze the performance of these two algorithms.  We also
+created a modified version of Minstrel that provides significant
+throughput gains over the vanilla Minstrel implementation.
+
+We provide a high-level overview of SampleRate and Minstrel in
+Sections~\ref{sec:samplerate} and~\ref{sec:minstrel}.  Then we discuss
+our testing methodology in Section~\ref{sec:methodology} and analyze
+the results in Section~\ref{sec:analysis}.  Finally we introduce our
+modifications to the Minstrel algorithm in Section~\ref{sec:minproved}
+and analyze the performance of our improvements.
 
@@
 of the two algorithms, and then discuss modifications we made to
-Minstrel and compare it's performance to the current kernel
+Minstrel and compare its performance to the current kernel
 implementation.
 
-\section{SampleRate}
+\section{SampleRate} \label{sec:samplerate}
 
@@
 rate, and uses the success or failure of this packet to update the
-average transmission time of that bit-rate.  SampleRate also abandons
+average transmission time of that bit rate.  SampleRate also abandons
 a bit rate after four successive failures at that rate.  All
@@
 SampleRate has three main functions: \texttt{ApplyRate()}, which
-returns a the bit-rate to send a packet at,
+returns a the bit rate to send a packet at,
 \texttt{ProcessFeedback()}, which updates the statistics for a
-bit-rate after an attempted packet send, and
+bit rate after an attempted packet send, and
 \texttt{RemoveStaleResults()}, which removes the effect of packets
@@
 
-\section{Minstrel}
+\section{Minstrel} \label{sec:minstrel}
 
@@
 average transmission time, Minstrel uses throughput, computed with $$T
-= \frac{p}{tx\_time}$$ where $T$ is the throughput of some rate $r$,
-$p$ is the probability of successfully transmitting at $r$, and
-$tx\_time$ is the computed lossless transmission time at rate $r$.
+= \frac{p}{\mathtt{tx\_time}}$$ where $T$ is the throughput of some
+rate $r$, $p$ is the probability of successfully transmitting at $r$,
+and $tx\_time$ is the computed lossless transmission time at rate $r$.
 
@@
 
-\section{Methodology}
+\section{Methodology} \label{sec:methodology}
 
@@
 \texttt{ath9k}, which were created to replace MadWifi, did not port
-the old bit-rate selection system from MadWifi and instead opted to
+the old bit rate selection system from MadWifi and instead opted to
 use the Linux kernel's rate selection framework, which only implements
@@
 Traces were collected from an \texttt{ath9k} driver modified to sample
-bit-rates uniformly (forgoing the usual bit rate selection algorithm)
-and to report successes and failure from each packet.  Each bit-rate
+bit rates uniformly (forgoing the usual bit rate selection algorithm)
+and to report successes and failure from each packet.  Each bit rate
 was sampled uniformly in time -- thus, more packets were sent at
-higher bit-rates, since those bit-rates otherwise deliver packets
+higher bit rates, since those bit rates otherwise deliver packets
 faster.  For each packet sent, the time to send was recorded, and the
-bit rate for the next packet was chosen to be the bit-rate that had
+bit rate for the next packet was chosen to be the bit rate that had
 the smallest total transmissions time.  This guarantees approximately
-uniform selection of bit-rates.  The wireless card used was an
+uniform selection of bit rates.  The wireless card used was an
 Qualcomm Atheros \texttt{AR9285} Wireless Network Adapter.
@@
 access point.  A few shorter (3 to 15 second long) traces were
-collected to investigate the startup behavior of bit-rate selection.
+collected to investigate the startup behavior of bit rate selection.
 
@@
   \includegraphics[width=\linewidth]{bsmt13floorplan.png}
-  \caption{We collected most traces at a remote access point in the basement of Building 13 at MIT.}
+  \caption{We collected most traces at a remote access point in the
+    basement of MIT's Building 13.}
 \label{figure:1}
@@
 functions, like SampleRate's \texttt{RemoveStaleResults} function,
-were instead called from one of \verb|apply_rate| or
-\verb|process_feedback|.  For every packet transmission, the simulator
-would call \texttt{ApplyRate}; compute the probability of success
-based on packets sent in 10-millisecond window around the current
-time; and from that determine the success or failure of the packet,
-and the number of retries required.  The results of a test run are
-thus non-deterministic; both algorithms tested are non-deterministic
-anyway, so this is no handicap.  To improve re-implementation
-fidelity, both implementations were meticulously checked by both
-authors.
+were instead called from one of \texttt{ApplyRate} or
+\texttt{ProcessFeedback}.  For every packet transmission, the
+simulator would call \texttt{ApplyRate}; compute the probability of
+success based on packets sent in 10-millisecond window around the
+current time; and from that determine the success or failure of the
+packet, and the number of retries required.  The results of a test run
+are thus non-deterministic; both algorithms tested are
+non-deterministic anyway, so this is no handicap.  To improve
+re-implementation fidelity, both implementations were meticulously
+checked by both authors.
 
@@
     \includegraphics[width=\linewidth]{constant.png}
-    \caption{Throughput of using a constant 802.11b/g bit-rate in
+    \caption{Throughput of using a constant 802.11b/g bit rate in
       clear line-of-sight from the access point.}
@@
 
-In Figure~\ref{figure:2} we show the performance of constant bit-rates
+In Figure~\ref{figure:2} we show the performance of constant bit rates
 on the \texttt{clear\_1.dat} trace, which is a 35 second trace in
@@
 which achieved an average throughput of about 15 Mbps.  The next
-highest bit-rate, 24 Mbps, shows a large plummet in performance, down
+highest bit rate, 24 Mbps, shows a large plummet in performance, down
 to a throughput of 1.17 Mbps.  This steep drop is charateristic, and
@@
 time probing high rates, it is less successful.  This is because
-Minstrel send the most packets at the three most successful bit-rates.
+Minstrel send the most packets at the three most successful bit rates.
 SampleRate, however, spends too much time sampling low bit rates.
@@
 
-\section{Improvements to Minstrel}
+\section{Improvements to Minstrel} \label{sec:minproved}
 
@@
 result.  Finally, we can simplify the above equation to $$S p \gets
-\frac{\beta \frac{d}{b} p + S n_i}{\beta \frac{d}{b} + d_i},$$ where
-$S$ is any constant.  This equation is usable in fixed-precision
-arithmetic, making it simple to implement in the Linux kernel (which
-prefers to avoid floating point computations for portability reasons).
-Because packets are high rates rarely succeed, the average window size
-is a few packets; thus, the main qualitative difference between our
-balanced EWMA and the original Minstrel EWMA is that non-probe packets
-at probe bit-rates are weighed extraordinarily heavily, and this fact
-allows us to avoid multiple failing 100ms windows.
+\frac{\beta \frac{d}{b} S p + S n_i}{\beta \frac{d}{b} + d_i},$$ where
+$S$ is any constant.  This equation is usable in fixed-point
+arithmetic (where $S$ is the scaling factor for the fixed-point
+representation), making it suitable for implementation in the Linux
+kernel (which avoids floating point computations for
+portability reasons).  Because packets are high rates rarely succeed,
+the average window size is a few packets; thus, the main qualitative
+difference between our balanced EWMA and the original Minstrel EWMA is
+that non-probe packets at probe bit rates are weighed extraordinarily
+heavily, and this fact allows us to avoid multiple failing 100ms
+windows.
 
@@
 implementation of Minstrel samples rates very frequently.  This is due
-to a flag the kernel uses to track which probe bit-rates were actually
-sampled.  In cases where the probe packets are at a lower bit-rate
-than the current bit-rate, the probe packets might never be sent if
-the current bit-rate succeeds.  However, the flagging seems to
+to a flag the kernel uses to track which probe bit rates were actually
+sampled.  In cases where the probe packets are at a lower bit rate
+than the current bit rate, the probe packets might never be sent if
+the current bit rate succeeds.  However, the flagging seems to
 misbehave in certain cases, causing the extremely aggressive probing
@@
     \centering
-    \begin{tabular}{l|l|l|l|l}
+    \begin{tabular}{r|rrr|r}
 
-    \textbf{Rate in Mbps} & \textbf{SampleRate} &
+    \textbf{Maximum Throughput} & \textbf{SampleRate} &
     \textbf{Minstrel} & \textbf{Minproved} &
-    \textbf{Constant Rate Throughput}\\ \hline
+    \textbf{Static Throughput}\\ \hline
 
-    1   & 631  & 39    & 29    & 0.639 Mbps \\
-    2   & 863  & 82    & 20    & 1.425 Mbps \\
-    5.5 & 4814 & 110   & 805   & 4.406 Mbps \\
-    6   & 423  & 137   & 276   & 4.603 Mbps \\
-    9   & 553  & 327   & 57    & 4.630 Mbps \\
-    11  & 3326 & 4121  & 3356  & 9.627 Mbps \\
-    12  & 4649 & 9515  & 14285 & 9.444 Mbps \\
-    18  & 1154 & 13295 & 15795 & 8.458 Mbps \\
-    24  & 16   & 6909  & 1918  & 0 Mbps \\
-    36  & 16   & 6944  & 2030  & 0 Mbps \\
-    48  & 16   & 6888  & 2009  & 0 Mbps \\
-    54  & 16   & 7308  & 1953  & 0 Mbps \\ \hline
+    1   Mbps & 631  pkts & 39    pkts & 29    pkts & 0.639 Mbps \\
+    2   Mbps & 863  pkts & 82    pkts & 20    pkts & 1.425 Mbps \\
+    5.5 Mbps & 4814 pkts & 110   pkts & 805   pkts & 4.406 Mbps \\
+    6   Mbps & 423  pkts & 137   pkts & 276   pkts & 4.603 Mbps \\
+    9   Mbps & 553  pkts & 327   pkts & 57    pkts & 4.630 Mbps \\
+    11  Mbps & 3326 pkts & 4121  pkts & 3356  pkts & 9.627 Mbps \\
+    12  Mbps & 4649 pkts & 9515  pkts & 14285 pkts & 9.444 Mbps \\
+    18  Mbps & 1154 pkts & 13295 pkts & 15795 pkts & 8.458 Mbps \\
+    24  Mbps & 16   pkts & 6909  pkts & 1918  pkts & 0 Mbps \\
+    36  Mbps & 16   pkts & 6944  pkts & 2030  pkts & 0 Mbps \\
+    48  Mbps & 16   pkts & 6888  pkts & 2009  pkts & 0 Mbps \\
+    54  Mbps & 16   pkts & 7308  pkts & 1953  pkts & 0 Mbps \\ \hline
 
@@
 SampleRate achieve higher throughput than Minproved, and in this case
-both surpassed the best constant bit-rate.  In cases where Minstrel
+both surpassed the best constant bit rate.  In cases where Minstrel
 out-performed SampleRate, the same behavior was true of Minproved.  In
@@
 it would be best if a successful probe packet at a rarely-successful
-bit-rate was treated more carefully.  One can imagine an algorithm,
+bit rate was treated more carefully.  One can imagine an algorithm,
 for example, which would treat a single successful probe packet as a
-reason to send a dozen more probe packets at that bit-rate, but does
-not yet commit to sending hundreds of packets at that bit-rate.
+reason to send a dozen more probe packets at that bit rate, but does
+not yet commit to sending hundreds of packets at that bit rate.
 
@@
 
-\section{Availability}
+\section{Availability} \label{sec:availability}
 
@@
 
-\section{Future Work}
+\section{Future Work} \label{sec:future-work}
 
-Very little analysis has been done of the performance of bit-rate
+Very little analysis has been done of the performance of bit rate
 selection algorithms on 802.11n networks.  802.11n introduces many new
@@
 
-\section{Conclusion}
+\section{Conclusion} \label{sec:conclusion}
 
@@
 
-\section{Acknowledgments}
+\section*{Acknowledgments} \label{sec:acknowledgments}
 

### 2013-05-17 paper/paper.tex
Minor stuff.

@@
 \begin{abstract}
-We compare the performance SampleRate and Minstrel, two popular
-bit rate selection algorithms that have widespread real-world usage.
-We use a trace-based approach to avoid kernel programming and consider
-Python implementations of both SampleRate and Minstrel.  We test both
-algorithms in multiple real-world scenarios, including scenarios with
-mobile clients and noisy environments, to highlight differences
-between the two algorithms.  We also introduce improvements to the
-Minstrel algorithm that allow for significant gains in throughput.
+We compare the performance SampleRate and Minstrel, two popular bit
+rate selection algorithms that have widespread real-world usage.  We
+use a trace-based approach to avoid kernel programming and improve
+reproducibility and allow analysis.  We test both algorithms in
+multiple real-world scenarios, including scenarios with mobile clients
+and noisy environments, to highlight differences between the two.  We
+also introduce improvements to the Minstrel algorithm that allow for
+significant gains in throughput.
 \end{abstract}
@@
 Minstrel and SampleRate perform similarly.  The results suggest that
-SampleRate tends to perform slightly better, except in very lossy
-links such as the client being far awat from the AP or actively
+SampleRate tends to perform better, except in very lossy
+links such as the client being far away from the AP or actively
 moving.  Minproved, our improved version of Minstrel, solidly
@@
 It seems that the replacement of SampleRate in the Linux kernel may
-have been a hasty, since it seems to perform similarly to Minstrel and
-it is a much simpler algorithm and thus simpler to adapt to changing
-wireless standards.  We are not sure if adjustments to SampleRate can
-bring its performance to match that of Minproved, however.
+have been hasty, since it performs similarly to Minstrel in many
+situations and in most situations and is a much simpler algorithm, and
+thus easier to adapt to changing wireless standards.  We are not sure
+if adjustments to SampleRate can bring its performance to match that
+of Minproved, however.
 
@@
 
-We would like to thank Jonathan Perry and Hari Balakrishnan for their
-valuable guidance, as well as Derek Smithies for answering our
+We would like to thank Jonathan~Perry and Hari~Balakrishnan for their
+valuable guidance, as well as Derek~Smithies for answering our
 questions about Minstrel.

### 2013-05-17 paper/paper.tex
fixed typo

--- 2013-05-17 paper/paper.tex
@@
 It seems that the replacement of SampleRate in the Linux kernel may
-have been a hasty, since it seems to perform similarly to Minstrel and
-it is a much simpler algorithm and thus simpler to adapt to changing
-wireless standards.  We are not sure if adjustments to SampleRate can
-bring its performance to match that of Minproved, however.
+have been a hasty, since it performs similarly or better than Minstrel 
+and in most situations and is a much simpler algorithm, and thus simpler
+ to adapt to changing wireless standards.  We are not sure if 
+adjustments to SampleRate can bring its performance to match that of 
+Minproved, however.
 

### 2013-05-17 paper/paper.tex
readme adjustments and fixed typo

--- 2013-05-17 paper/paper.tex
@@
 which rates to try before reporting a failure.  The retry chain makes
-failures extremely unlikely and allows Minstrel to choose fallback
+failures extremely unlikely and allows Minstrel to choose fall-back
 rates when a packet does not succeed.  Minstrel sets the MRR to first
@@
 from how it was implemented in the MadWifi kernel driver.  The primary
-difference is that the kernel implementatiion uses a EWMA, while the
+difference is that the kernel implementation uses a EWMA, while the
 thesis implementation computes average transmission times based on a
@@
 Minstrel and SampleRate over all the different types of traces we
-collected.  This is consisten with the literature---the authors of
+collected.  This is consistent with the literature---the authors of
 Minstrel noted that they focused on making Minstrel robust to poor
-conditions~\cite{minstrel}, which means losing potential througput
+conditions~\cite{minstrel}, which means losing potential throughput
 gains in stable situations.  Conversely, John Bicket noted that
@@
 highest bit rate, 24 Mbps, shows a large plummet in performance, down
-to a throughput of 1.17 Mbps.  This steep drop is charateristic, and
+to a throughput of 1.17 Mbps.  This steep drop is characteristic, and
 is also talked about in the SampleRate thesis~\cite{samplerate}.  It
@@
 demonstrate that 11, 12 and 18 Mbps were the optimal rates to send at,
-and that no packets were sucessful at any higher rates.  SampleRate
+and that no packets were successful at any higher rates.  SampleRate
 does not use these rates except for probe packets, which happens once
@@
 of time sampling at high rates.  In the specific case we examined, they
-were all probe packets.  But sometimes Minstrel would choose 54Mbps as
+were all probe packets.  But sometimes Minstrel would choose 54 Mbps as
 the rate with the best throughput, which made little sense.  A closer
 investigation revealed that Minstrel would sometimes get lucky with a
-successful probe at 54Mbps.  The probability of success would be 1, so
-then 54Mbps would be chosen as the rate with the best throughput.  This
+successful probe at 54 Mbps.  The probability of success would be 1, so
+then 54 Mbps would be chosen as the rate with the best throughput.  This
 is unfortunate, but the real harm comes in the next window: packets
-would now be send at 54Mbps, and nearly all would fail.  However,
+would now be send at 54 Mbps, and nearly all would fail.  However,
 since these hundreds of failed packets weight equally with the single

### 2013-08-06 paper/paper.tex

@@
 
-One of the key ways that wireless networks differ from wired links is
+One of the key ways that wireless networks differ from wired is
 that wireless networks have varying link rates.  Link conditions vary
@@
 
-This paper will provide a general overview of SampleRate and Minstrel.
-We then discuss our testing methodology.  We compare the performance
-of the two algorithms, and then discuss modifications we made to
-Minstrel and compare its performance to the current kernel
-implementation.
-
 \section{SampleRate} \label{sec:samplerate}
@@
 selection algorithm which maintains estimates of average transmission
-time for each potential bit rate.To keep these estimates up to date,
+time for each potential bit rate. To keep these estimates up to date,
 SampleRate periodically sends a packet at a randomly-selected bit
@@
 failures extremely unlikely and allows Minstrel to choose fall-back
-rates when a packet does not succeed.  Minstrel sets the MRR to first
-try the rate with the highest throughput, then the next-highest
-throughput, then greatest probability of success, and finally attempts
-to send a packet at the lowest, base rate.  In essence, Minstrel tells
-the card to send at the highest throughput rates but to fall back to
-reliable rates upon failure.
+rates when a packet does not succeed.  When not sending a probe packet,
+Minstrel sets the MRR to first try the rate with the highest throughput, 
+then the next-highest throughput, then greatest probability of success, 
+and finally attempts to send a packet at the lowest base rate.  
+In essence, Minstrel tells the card to send at the highest throughput 
+rates but to fall back to reliable rates upon failure.
 
@@
 Minstrel tries to avoid sampling at a rate slower than the current
-best rate, if the current best rate is still successfully
-transmitting.  If the randomly chosen rate has a higher lossless
+best rate.  If the randomly chosen rate has a higher lossless
 throughput than the current optimal rate, the MRR first lists the
@@
 An EWMA ensures that a sudden degradation in link quality will create
-a rapid response in the probabilities, making the EWMA-using Minstrel
+a rapid response in the probabilities, making Minstrel
 perform better in dynamic environments.  Minstrel will waste less time
@@
 MadWifi drivers for the Atheros chip.  However, MadWifi is deprecated,
-and only runs on now-ancient network cards.  In addition to the
+and only runs on very old network cards.  In addition to the
 unavailability of compatible hardware, MadWifi would not be an
@@
 Minstrel.  Porting SampleRate over to the new drivers would not have
-been trivial, as there are vast differences in the assumed interfaces.
+been trivial, as there are vast differences in the interfaces.
 
@@
 Traces were collected from an \texttt{ath9k} driver modified to sample
-bit rates uniformly (forgoing the usual bit rate selection algorithm)
-and to report successes and failure from each packet.  Each bit rate
-was sampled uniformly in time -- thus, more packets were sent at
-higher bit rates, since those bit rates otherwise deliver packets
-faster.  For each packet sent, the time to send was recorded, and the
-bit rate for the next packet was chosen to be the bit rate that had
-the smallest total transmissions time.  This guarantees approximately
+bit rates uniformly and to report successes and failure from each packet.  
+Each bit rate was sampled uniformly in time --- thus, more packets 
+were sent at higher bit rates, since those bit rates deliver
+packets faster. For each packet sent, we recorded the time it was sent,
+the bit rate it was sent at, and whether or it it succeeded. The bit rate 
+for the next trace packet was chosen to be the bit rate that had
+the smallest total transmission time so far.  This guarantees approximately
 uniform selection of bit rates.  The wireless card used was an
@@
 byte packets to the access point's IP address.  Packets were sent over
-UDP to avoid the TCP congestion window and retry loop; since no users
-were using the access point, there was no real risk of deteriorating
-service for other users.  Figure~\ref{figure:1} shows a floor plan of
-the access point and basement.
+UDP to avoid the TCP congestion window and retry loop. Figure~\ref{figure:1} 
+shows a floor plan of the access point and basement.
 
@@
 Once collected, traces were fed to a bit rate selection simulator to 
-analyze the performance of Minstrel and SampleRate.  Both algorithms
-were represented by two functions: \texttt{ApplyRate}, which returned
-a multi-rate retry chain of rates to attempt transmission at, and
-\texttt{ProcessFeedback}, which was passed the success or failure of
-the transmission and how many attempts were made at each bit rate
-(along with the current time and the total time to send the packet);
-this parallels the interface used by the Linux kernel itself.  Helper
-functions, like SampleRate's \texttt{RemoveStaleResults} function,
-were instead called from one of \texttt{ApplyRate} or
-\texttt{ProcessFeedback}.  For every packet transmission, the
-simulator would call \texttt{ApplyRate}; compute the probability of
-success based on packets sent in 10-millisecond window around the
-current time; and from that determine the success or failure of the
-packet, and the number of retries required.  The results of a test run
-are thus non-deterministic; both algorithms tested are
-non-deterministic anyway, so this is no handicap.  To improve
-re-implementation fidelity, both implementations were meticulously
-checked by both authors.
+analyze the performance of Minstrel and SampleRate. 
+\texttt{harness.py} interfaced with the bit-rate selection algorithms using
+two functions: \texttt{ApplyRate()} and \texttt{ProcessFeedback()}.
+\texttt{ApplyRate} returns a multi-rate retry chain of rates to attempt 
+transmission at. The retry chain is one element long for SampleRate, since it
+does not use MRR. \texttt{ProcessFeedback} is a callback that is passed 
+the success or failure of the transmission and how many attempts were made
+at each bit rate. This parallels the interface used by the Linux kernel 
+itself. For every packet transmission, the simulator would call 
+\texttt{ApplyRate}; compute the probability of success based on packets 
+sent in 10-millisecond window around the current time; and from that 
+determine the success or failure of the packet, and the number of retries 
+it required.
 
@@
 structures for tracking rate statistics.
+
+To improve re-implementation fidelity, both implementations were meticulously checked by both authors.
 
@@
 Minstrel and SampleRate perform similarly, usually within 3 Mbps of
-each other.  In especially noisy situations, such as when the client
+each other.  We were slightly surprised to
+see how significantly SampleRate outperformed Minstrel in certain
+situations, though, with throughput being up to $67\%$ higher.
+However, in especially noisy situations, such as when the client
 is moving or when the client is around the corner from the access
@@
 SampleRate performs better in stable situations than noisy or highly
-varying conditions~\cite{samplerate}.  We were slightly surprised to
-see how significantly SampleRate outperformed Minstrel in certain
-situations, though, with throughput being up to $67\%$ higher.
+varying conditions~\cite{samplerate}.
 
@@
 is also talked about in the SampleRate thesis~\cite{samplerate}.  It
-demonstrates just how important bit rate selection is---the optimal
+demonstrates just how important bit rate selection is --- the optimal
 bit rate has typically much higher throughput than the next-best rate,
@@
 averaged 5.284 Mbps over 10 runs, with a maximum variation of about
-+/- 0.3 Mbps.  The variation of results on other traces (except the
-very short traces, where variance was naturally higher) was similar.
++/- 0.3 Mbps.  The variation of results on other traces (except for 
+traces 5s or less) was similar.
 
@@
 reason low rates have fewer probe packets is because Minstrel
-explicitly places them second in the retry chain, and because the main
-rate usually succeeds.  Despite the fact that SampleRate spends less
+explicitly places them second in the retry chain.
+Despite the fact that SampleRate spends less
 time probing high rates, it is less successful.  This is because
@@
 SampleRate, however, spends too much time sampling low bit rates.
-Additionally, it spends a lot of time sending at 5.5 Mbps.  We suspect
+Additionally, it spends a lot of time sending at 5.5 Mbps, which is not
+one of the optimal rates.  We suspect
 that this might be because the equation SampleRate uses to estimate
@@
 then 54 Mbps would be chosen as the rate with the best throughput.  This
-is unfortunate, but the real harm comes in the next window: packets
-would now be send at 54 Mbps, and nearly all would fail.  However,
+is unfortunate, but the real harm comes in the next window: all non-probe 
+packets would now be sent at 54 Mbps, and nearly all would fail.  However,
 since these hundreds of failed packets weight equally with the single
@@
 highly.  Fundamentally, the problem is that the EWMA does not account
-the second block having hundreds more packets than the first.  In the
-next section we will talk about Minproved, an improved version of
-Minstrel where we attempt to fix this problem.
+the second block having hundreds more packets than the first, so the 
+probabilities were not declining fast enough after a large number of
+   failures.
+
+Additionally, Minstrel uses a complex method of calculating when to send 
+a sample packet that involves a heuristic to make sure that 10\% of 
+packets were actually samples. This is because the MRR places sample 
+packets second in the retry chain when the sample rate is slower than 
+the rate with the best throughput. So even if 10\% of probes have a 
+sample rate in the retry chain, a sample rate is not actually sent 10\% 
+of the time. We implemented this heuristic and it tended to cause 
+Minstrel to send too many probe packets.
+
+In the next section we will talk about Minproved, an improved version 
+of Minstrel where we attempt to fix problems with EWMA and too-frequent
+ probing.
 
@@
 The usual algorithm computes $$p \gets \alpha p + (1 - \alpha)
-\frac{n_i}{d_i}),$$ where $p$ is the EWMA probability, $\alpha$ is a
+\frac{n_i}{d_i}),$$ where $p$ is the probability, $\alpha$ is a
 weight factor equal to $0.75$ in Minstrel, and $n_i$ and $d_i$ are the
@@
   \frac{n_i}{d_i}}{\beta + w_i},$$ where the weight $w_i$ is $d_i / (d
-/ b)$, with $d$ the total number of packets ever send and $b$ the
+/ b)$, with $d$ the total number of packets ever sent and $b$ the
 total number of 100ms windows ever seen.  Then $d / b$ is the average
@@
 kernel (which avoids floating point computations for
-portability reasons).  Because packets are high rates rarely succeed,
-the average window size is a few packets; thus, the main qualitative
+portability reasons).  Because packets at high rates rarely succeed,
+the average window size is a few packets.  The main qualitative
 difference between our balanced EWMA and the original Minstrel EWMA is
-that non-probe packets at probe bit rates are weighed extraordinarily
-heavily, and this fact allows us to avoid multiple failing 100ms
-windows.
+that windows with many more packets than the average for that bit-rate are
+weighted extraordinarily heavily. This fact allows us to avoid multiple 
+failing 100ms windows.
 
@@
 
-We noted another curious fact about Minstrel: the kernel
-implementation of Minstrel samples rates very frequently.  This is due
-to a flag the kernel uses to track which probe bit rates were actually
-sampled.  In cases where the probe packets are at a lower bit rate
-than the current bit rate, the probe packets might never be sent if
-the current bit rate succeeds.  However, the flagging seems to
-misbehave in certain cases, causing the extremely aggressive probing
+As noted in Section~\ref{sec:analysis}, Minstrel sends sample packets
+too frequently. This is due to a flag the kernel uses to track whether
+a probe bit rates was actually sampled. Recall that if the sample rate
+is slower than the current best rate, it is second in the retry chain, 
+so there is a good chance that the random rate will never be used to
+send a packet. This system is used as an attempt to make sure that 10\%
+of the sent packets are actually probes, as opposed to simply having 
+10\% of packets contain a random rate somewhere in the retry chain.
+ The flagging seems to misbehave, causing the extremely aggressive probing
 we see in some traces.  For example, in one of the traces where the
@@
 half of its packets at probe frequencies.  We modified Minstrel to be
-less aggressive with probe packets, decreasing the sampling frequency
-to approximately 10\%.  This lead to more improvement, especially on
-noisy traces such as the cases where the client was around a corner or
-moving.
+less aggressive with probe packets. We eliminated the flagging system,
+and instead simply put a random bit rate in 10\% of packets.
+This lead to a very large throughput improvement, bringing Minstrel to
+the same level of performance as SampleRate, and greatly exceeding it 
+on noisy traces such as the cases where the client was around a corner or
+moving. When this change was paired with the balanced EWMA change, it
+consistently outperformed both vanilla Minstrel and SampleRate.
 

### 2013-08-06 data/old_data/numpyify.py
Basic cleanup

@@ -1,67 +1,0 @@
-
-import numpy
-import collections
-import re
-import sys
-
-FIELDS = ["ts", "delay", "tries", "rateid", "kbps", "user_kbps", "i"]
-Record = collections.namedtuple("Record", FIELDS)
-
-NUM_RATES = 24
-
-LINE_RE = r"""Last\((\d+)\) took (\d+) ns / (\d+) tries with rate (\d+) at (\d+)\((\d+)\) kbps \[(\d+)\]\n"""
-LINE_RE = re.compile(LINE_RE)
-def parse_line(lines):
-    for line in lines:
-        match_info = LINE_RE.match(line)
-    
-        if not match_info:
-            print(ValueError("Cannot parse line", line))
-        else:
-            yield Record(*map(int, match_info.groups()))
-
-def clean_up(record_stream):
-    prev = next(record_stream)
-    yield prev
-
-    for record in record_stream:
-        # The kernel sets:
-        #   ts and rateid, kbps, user_kbs
-        # then it sends the packet
-        # then it sets:
-        #   delay, tries
-        # So we uniquify by the delay, which should be unique
-        if record.delay != prev.delay and record.rateid != prev.rateid and record.ts != prev.ts:
-            yield record
-            prev = record
-
-def numpyify(record_stream):
-    array = [[(0,0)] * NUM_RATES]
-
-    for record in record_stream:
-        idx = record.rateid
-
-        if array[-1][idx] != (0, 0):
-            array.append([(0, 0)] * NUM_RATES)
-
-        array[-1][idx] = (record.delay, record.tries)
-
-    return numpy.array(array)
-
-def save(array):
-    numpy.save(OUTFILE, array)
-
-def pipe(val, *fns):
-    for fn in fns:
-        val = fn(val)
-    return val
-
-OUTFILE = None
-
-if __name__ == "__main__":
-    OUTFILE = open(sys.argv[1], "wb")
-    pipe(sys.stdin,
-         parse_line,
-         clean_up,
-         numpyify,
-         save)

### 2013-08-06 raw/raw_old/collect.sh

@@ -1,10 +1,0 @@
-#!/bin/sh
-
-FILE=$1
-
-cat /proc/ath_rate >/dev/null
-
-while true; do
-    cat /proc/ath_rate
-    sleep .5
-done | tee "$FILE"

### 2013-08-06 raw/raw_old/reprobe.sh

@@ -1,6 +1,0 @@
-#!/bin/sh
-
-netctl stop mit
-rmmod ath9k ath9k_common ath9k_hw ath
-modprobe ath9k
-netctl start mit

### 2013-08-06 raw/raw_old/spew.py

@@ -1,16 +1,0 @@
-import socket
-import sys
-import random
-
-UDP_IP = sys.argv[1]
-UDP_PORT = 5001
-MESSAGE_LEN = 1500 - 64
-
-print("Spewing to {}:{}, {} bytes per packet".format(UDP_IP, UDP_PORT,
-                                                     MESSAGE_LEN))
-
-sock = socket.socket(socket.AF_INET, socket.SOCK_DGRAM)
-
-while True:
-    MESSAGE = bytes(random.randint(0, 255) for i in range(MESSAGE_LEN))
-    sock.sendto(MESSAGE, (UDP_IP, UDP_PORT))

### 2013-08-06 raw/raw_old/weps.py

@@ -1,16 +1,0 @@
-import socket
-import sys
-import random
-
-UDP_IP = sys.argv[1]
-UDP_PORT = 5001
-
-
-print("Listening on %r:%r." % (UDP_IP, UDP_PORT))
-
-sock = socket.socket(socket.AF_INET, socket.SOCK_DGRAM)
-sock.bind((UDP_IP, UDP_PORT))
-
-while True:
-    data, addr = sock.recvfrom(1024)
-    print("recieved message ", data)

### 2013-08-06 data/explore.py
More cleanup, moving around scripts.

@@ -1,33 +1,0 @@
-# This file is meant to be loaded from a iPython shell; run
-#
-#   %load explore.py
-#
-# to load it, then load a data file with
-#
-#   init("file.dat")
-
-start = 0
-data = [[] for i in range(12)]
-end = 0
-
-npkts = 0
-
-def init(fn):
-    global start, data, end, npkts
-    start, data, end = eval(open(fn, "rt").read())
-
-    npkts = sum(len(rdata) for rdata in data)
-
-def cum(s):
-     i = 0
-     for v in s:
-         i += v
-         yield i
-
-def rate_cdf(rate):
-    xs, ys = zip(*map(lambda x: x[:2], data[rate]))
-    return array(xs), array(list(cum(ys)))
-
-def plot_rates():
-    for i in range(len(data)):
-        plot(*rate_cdf(i))

### 2013-08-06 util/reprobe.sh

@@ -1,6 +1,7 @@
 #!/bin/sh
 
-netctl stop mit
+netctl store
+netctl stop-all
 rmmod ath9k ath9k_common ath9k_hw ath
 modprobe ath9k
-netctl start mit
+netctl restore

### 2013-08-06 util/collect.sh
Collect script has matured a lot.

@@ -1,10 +1,29 @@
 #!/bin/sh
 
-FILE=$1
+PORT=$1
+FILE=$2
+SEC=$3
+ITER=$((2*SEC))
 
+echo "Reprobing wifi driver (this may take a while)"
+./reprobe.sh
+echo "Starting packet spew"
+python ./spew.py "$PORT" &
+
+# Clear the stored buffer
+cat /proc/ath_rate >/dev/null
+
+echo "Starting capture"
 cat /proc/ath_rate
 
-while true; do
+while [ ITER -ne 0 ]; do
     cat /proc/ath_rate
     sleep .5
+    ITER=$((ITER-1))
 done | tee "$FILE"
+
+echo "Capture done"
+kill %python
+
+echo "Check: all numbers from 0 to 11 appear in the right hand column:"
+./packets.sh "$FILE"

### 2013-08-06 util/spew.py

@@ -19,5 +19,4 @@ while True:
     PREFIX = "Seq {}\n".format(i).encode("utf-8")
     MESSAGE = PREFIX + bytes(random.randint(0, 255) for i in range(MESSAGE_LEN - len(PREFIX)))
     sock.sendto(MESSAGE, (UDP_IP, UDP_PORT))
-    print(PREFIX)
     i += 1

### 2013-08-06 util/collect.sh
Bug fixes to collect.sh

@@ -6,7 +6,7 @@ SEC=$3
 ITER=$((2*SEC))
 
 echo "Reprobing wifi driver (this may take a while)"
-./reprobe.sh
+sudo ./reprobe.sh
 echo "Starting packet spew"
 python ./spew.py "$PORT" &
 
@@ -16,10 +16,10 @@ cat /proc/ath_rate >/dev/null
 echo "Starting capture"
 cat /proc/ath_rate
 
-while [ ITER -ne 0 ]; do
+while [ $ITER -ne 0 ]; do
     cat /proc/ath_rate
     sleep .5
-    ITER=$((ITER-1))
+    ITER=$((ITER - 1))
 done | tee "$FILE"
 
 echo "Capture done"

### 2013-08-06 util/munge.py
Made munge ignore the sanity-check lines.

@@ -9,16 +9,24 @@ Record = collections.namedtuple("Record", FIELDS)
 
 NUM_RATES = 12
 
+SANITY_RE = r"""\(\d+:\d+\s+\){""" + str(NUM_RATES) + "}"
+SANITY_RE = re.compile(SANITY_RE)
+
 LINE_RE = r"""Last\((\d+)\.(\d+)\) took (\d+) ns / (\d+) tries with rate (\d+) at (\d+)\((\d+)\) kbps \[(\d+)\]\n"""
 LINE_RE = re.compile(LINE_RE)
 def parse_line(lines):
     for line in lines:
         match_info = LINE_RE.match(line)
     
-        if not match_info:
-            print(ValueError("Cannot parse line", line))
-        else:
+        if match_info:
             yield Record(*map(int, match_info.groups()))
+            continue
+
+        match_info = SANITY_RE.match(line)
+        if match_info:
+            continue
+
+        print("Cannot parse line:", line, end="")
 
 def ts(record):
     return int(record.ts_s*1e9 + record.ts_ns)

### 2013-08-06 util/munge.py
Where are there so many different types of regexes?

@@ -9,7 +9,7 @@ Record = collections.namedtuple("Record", FIELDS)
 
 NUM_RATES = 12
 
-SANITY_RE = r"""\(\d+:\d+\s+\){""" + str(NUM_RATES) + "}"
+SANITY_RE = r"""(\d+:\d+\s+){""" + str(NUM_RATES) + "}"
 SANITY_RE = re.compile(SANITY_RE)
 
 LINE_RE = r"""Last\((\d+)\.(\d+)\) took (\d+) ns / (\d+) tries with rate (\d+) at (\d+)\((\d+)\) kbps \[(\d+)\]\n"""

### 2013-08-06 collect.sh
Moved collect.sh, since it is now the only entry-point.

@@ -5,10 +5,15 @@ FILE=$2
 SEC=$3
 ITER=$((2*SEC))
 
+if [ -f "raw/$FILE.trace" -o "$FILE" = "test"]; do
+    echo "File 'raw/$FILE.trace' already exists; will not overwrite"
+    exit 1
+done
+
 echo "Reprobing wifi driver (this may take a while)"
-sudo ./reprobe.sh
+sudo util/reprobe.sh
 echo "Starting packet spew"
-python ./spew.py "$PORT" &
+python util/spew.py "$PORT" &
 
 # Clear the stored buffer
 cat /proc/ath_rate >/dev/null
@@ -20,10 +25,13 @@ while [ $ITER -ne 0 ]; do
     cat /proc/ath_rate
     sleep .5
     ITER=$((ITER - 1))
-done | tee "$FILE"
+done | tee "raw/$FILE.trace"
 
 echo "Capture done"
 kill %python
 
 echo "Check: all numbers from 0 to 11 appear in the right hand column:"
-./packets.sh "$FILE"
+util/packets.sh "../raw/$FILE.trace"
+
+echo "Munging data"
+python util/munge.py <"../raw/$FILE.trace" "data/$FILE.dat"

### 2013-08-06 collect.sh
Woops

@@ -5,10 +5,10 @@ FILE=$2
 SEC=$3
 ITER=$((2*SEC))
 
-if [ -f "raw/$FILE.trace" -o "$FILE" = "test"]; do
+if [ -f "raw/$FILE.trace" -o "$FILE" = "test"]; then
     echo "File 'raw/$FILE.trace' already exists; will not overwrite"
     exit 1
-done
+fi
 
 echo "Reprobing wifi driver (this may take a while)"
 sudo util/reprobe.sh

### 2013-08-06 collect.sh
Bash!

@@ -5,7 +5,7 @@ FILE=$2
 SEC=$3
 ITER=$((2*SEC))
 
-if [ -f "raw/$FILE.trace" -o "$FILE" = "test"]; then
+if [ -f "raw/$FILE.trace" -o "$FILE" = "test" ]; then
     echo "File 'raw/$FILE.trace' already exists; will not overwrite"
     exit 1
 fi

### 2013-08-06 collect.sh
Got my conditional backwards

@@ -5,7 +5,7 @@ FILE=$2
 SEC=$3
 ITER=$((2*SEC))
 
-if [ -f "raw/$FILE.trace" -o "$FILE" = "test" ]; then
+if [ -f "raw/$FILE.trace" -a "$FILE" != "test" ]; then
     echo "File 'raw/$FILE.trace' already exists; will not overwrite"
     exit 1
 fi

### 2013-08-06 collect.sh
Forgot to change a few paths.

@@ -31,7 +31,7 @@ echo "Capture done"
 kill %python
 
 echo "Check: all numbers from 0 to 11 appear in the right hand column:"
-util/packets.sh "../raw/$FILE.trace"
+util/packets.sh "raw/$FILE.trace"
 
 echo "Munging data"
-python util/munge.py <"../raw/$FILE.trace" "data/$FILE.dat"
+python util/munge.py <"raw/$FILE.trace" "data/$FILE.dat"

### 2013-08-06 pysim/old/harness.py
Deleted old harness code

@@ -1,133 +1,0 @@
-import numpy
-import time
-import common
-import os, sys
-import random
-
-DEBUG = "DEBUG" in os.environ
-
-def load_data(source):
-    return numpy.load(source) / 1. # Convert to float
-
-def packet_stats(data, time, rate):
-    time = time % len(data)
-    
-    # First, we try to find a prior instance of such transmission
-    for i in range(time, -1, -1):
-        if data[i, rate].any():
-            delay, tries = data[i, rate]
-            return (delay, tries)
-
-    # Next, try to find a future instance
-    for i in range(len(data)-1, time, -1):
-        if data[i, rate].any():
-            delay, tries = data[i, rate]
-            return (delay, tries)
-    
-    # Otherwise, this seems like a big problem, so we cause an error
-    raise ValueError("No data available at bit-rate", rate)
-
-def baseline(data):
-    s = data.sum(0)[0]
-    return s[0] / s[1]
-
-class Harness:
-    def __init__(self, data, choose_rate, push_statistics):
-        # Keep the time somewhat believable
-        #self.clock = time.time()
-        self.clock = 0
-        self.packet_num = 0
-        self.baseline = baseline(data)
-        self.choose_rate = choose_rate
-        self.push_statistics = push_statistics
-
-        self.histogram = [0] * len(common.RATES)
-
-    def send_packet(self):
-        rate_arr = self.choose_rate(self.clock)
-
-        if DEBUG:
-            print("Sending packet at:", end=" ")
-            for (rate, tries) in rate_arr:
-                print("Rate {}/{}".format(rate, tries), end=" ")
-            print()
-
-        tot_delay = 0
-        tot_tries = []
-        tot_status = None
-        for (rate, tries) in rate_arr:
-            s_delay, s_tries = packet_stats(data, self.packet_num, rate)
-
-            # Correct for the extra packet transmission at the 0 rate
-            if s_tries >= 20:
-                s_delay = s_delay - self.baseline
-
-            # If it would take more tries than we have
-            if s_tries > tries:
-                tot_tries.append((rate, tries))
-                self.histogram[rate] += tries
-                tot_delay += s_delay * (tries / s_tries)
-            else: # We successfully transmit the packet
-                tot_tries.append((rate, s_tries))
-                self.histogram[rate] += s_tries
-                tot_delay += s_delay
-                tot_status = True # Success
[…]

### 2013-08-06 pysim/common.py
Stupid bug about OFDM bitrates

@@ -99,7 +99,7 @@ class BalancedEWMA:
 def tx_time(rix, length=1200): #rix is index to RATES, length in bytes
     mbps = RATES[rix].dot11_rate / 2000
 
-    if RATES[rix].phy == "odfm":
+    if RATES[rix].phy == "ofdm":
         '''* OFDM:
         *
         * N_DBPS = DATARATE x 4

### 2013-08-06 pysim/common.py
ieee80211_to_idx now returns one option, not a list of options.

@@ -35,8 +35,12 @@ RATES = [
 ]
 
 def ieee80211_to_idx(mbps):
-    return [i for i, rate in enumerate(RATES)
+    opts = [i for i, rate in enumerate(RATES)
             if rate.dot11_rate == int(round(2 * mbps))]
+    if opts:
+        return opts[0]
+    else:
+        raise ValueError("No bitrate with throughput {} Mbps exists".format(mbps))
 
 
 class EWMA:

### 2013-08-06 pysim/constant.py

@@ -3,16 +3,17 @@ import common
 
 RATE=float(os.environ["RATE"]) if "RATE" in os.environ else 1
 
-opts = common.ieee80211_to_idx(RATE)
-if opts:
-    FIXED_RATE = opts[0]
-    print("Running at rate %r Mbps..." % RATE)
-else:
-    print("Invalid rate.  Options are: {}".format(", ".join(str(rate.dot11_rate / 2) for rate in common.RATES)))
+try:
+    IDX = common.ieee80211_to_idx(RATE)
+except ValueError:
+    opts = [str(rate.dot11_rate / 2) for rate in common.RATES]
+    print("Invalid rate.  Options are: {}".format(", ".join(opts)))
     exit()
+else:
+    print("Running at rate %r Mbps..." % RATE)
 
 def apply_rate(time):
-    return [(FIXED_RATE, 4)] * 4
+    return [(IDX, 4)] * 4
 
 def process_feedback(status, timestamp, delay, tries):
     pass

### 2013-08-06 pysim/minproved.py

@@ -1,8 +1,6 @@
 # Colleen Josephson, 2013
 # This file attempts to implement the minstrel rate control algorithm
 
-# TODO : seed 1466742691193015552
-
 from random import randint
 from random import choice 
 import common
@@ -191,13 +189,13 @@ def apply_rate(cur_time): #cur_time is in nanoseconds
             random = choice(list(rates))
 
         if random < bestThruput:
-            r = [(ieee80211_to_idx(bestThruput)[0],
+            r = [(ieee80211_to_idx(bestThruput),
                      rates[bestThruput].adjusted_retry_count),
-                    (ieee80211_to_idx(random)[0],
+                    (ieee80211_to_idx(random),
                      rates[random].adjusted_retry_count), 
-                    (ieee80211_to_idx(bestProb)[0],
+                    (ieee80211_to_idx(bestProb),
                      rates[bestProb].adjusted_retry_count), 
-                    (ieee80211_to_idx(lowestRate)[0],
+                    (ieee80211_to_idx(lowestRate),
                      rates[lowestRate].adjusted_retry_count)]
         else:
             #TODO: understand the corresponding kernel code more 
@@ -206,23 +204,23 @@ def apply_rate(cur_time): #cur_time is in nanoseconds
                 if rates[random].sample_limit > 0:
                     rates[random].sample_limit -= 1
             
-            r = [(ieee80211_to_idx(random)[0],
+            r = [(ieee80211_to_idx(random),
                       rates[random].adjusted_retry_count), 
-                    (ieee80211_to_idx(bestThruput)[0],
+                    (ieee80211_to_idx(bestThruput),
                      rates[bestThruput].adjusted_retry_count),
-                    (ieee80211_to_idx(bestProb)[0],
+                    (ieee80211_to_idx(bestProb),
                      rates[bestProb].adjusted_retry_count), 
-                    (ieee80211_to_idx(lowestRate)[0],
+                    (ieee80211_to_idx(lowestRate),
                      rates[lowestRate].adjusted_retry_count)]
     
     else:     #normal
-        r = [(ieee80211_to_idx(bestThruput)[0],
+        r = [(ieee80211_to_idx(bestThruput),
               rates[bestThruput].adjusted_retry_count), 
-             (ieee80211_to_idx(nextThruput)[0],
+             (ieee80211_to_idx(nextThruput),
               rates[nextThruput].adjusted_retry_count), 
[…]

### 2013-08-06 pysim/minstrel.py

@@ -206,13 +206,13 @@ def apply_rate(cur_time): #cur_time is in nanoseconds
             random = choice(list(rates))
 
         if random < bestThruput:
-            r = [(ieee80211_to_idx(bestThruput)[0],
+            r = [(ieee80211_to_idx(bestThruput),
                      rates[bestThruput].adjusted_retry_count),
-                    (ieee80211_to_idx(random)[0],
+                    (ieee80211_to_idx(random),
                      rates[random].adjusted_retry_count), 
-                    (ieee80211_to_idx(bestProb)[0],
+                    (ieee80211_to_idx(bestProb),
                      rates[bestProb].adjusted_retry_count), 
-                    (ieee80211_to_idx(lowestRate)[0],
+                    (ieee80211_to_idx(lowestRate),
                      rates[lowestRate].adjusted_retry_count)]
             ''' FROM KERNEL:
             /* Only use IEEE80211_TX_CTL_RATE_CTRL_PROBE to mark        
@@ -231,23 +231,23 @@ def apply_rate(cur_time): #cur_time is in nanoseconds
                 if rates[random].sample_limit > 0:
                     rates[random].sample_limit -= 1
             
-            r = [(ieee80211_to_idx(random)[0],
+            r = [(ieee80211_to_idx(random),
                       rates[random].adjusted_retry_count), 
-                    (ieee80211_to_idx(bestThruput)[0],
+                    (ieee80211_to_idx(bestThruput),
                      rates[bestThruput].adjusted_retry_count),
-                    (ieee80211_to_idx(bestProb)[0],
+                    (ieee80211_to_idx(bestProb),
                      rates[bestProb].adjusted_retry_count), 
-                    (ieee80211_to_idx(lowestRate)[0],
+                    (ieee80211_to_idx(lowestRate),
                      rates[lowestRate].adjusted_retry_count)]
     
     else:     #normal
-        r = [(ieee80211_to_idx(bestThruput)[0],
+        r = [(ieee80211_to_idx(bestThruput),
               rates[bestThruput].adjusted_retry_count), 
-             (ieee80211_to_idx(nextThruput)[0],
+             (ieee80211_to_idx(nextThruput),
               rates[nextThruput].adjusted_retry_count), 
-             (ieee80211_to_idx(bestProb)[0],
+             (ieee80211_to_idx(bestProb),
               rates[bestProb].adjusted_retry_count), 
-             (ieee80211_to_idx(lowestRate)[0],
[…]

### 2013-08-06 pysim/samplerate.py

@@ -21,7 +21,7 @@ backoff = {0:0, 1:155, 2:315, 3:635, 4:1275, 5:2555, 6:5115, 7:5115, 8:5115, 9:5
            18:5115, 19:5115, 20:5115}
 
 def bitrate_type(bitrate):
-    return common.RATES[ieee80211_to_idx(bitrate)[0]].phy
+    return common.RATES[ieee80211_to_idx(bitrate)].phy
 
 
 
@@ -113,7 +113,7 @@ def apply_rate(cur_time):
         for i, r in sorted(rates.items(), reverse=True):
             if r.succFails < 4:
                 currRate = r.rate
-                return [(ieee80211_to_idx(currRate)[0], NRETRIES)]
+                return [(ieee80211_to_idx(currRate), NRETRIES)]
 
     # Every 10 packets, select a random non-failing bit rate w/ better avg tx
     #"If the number of packets sent over the link is a multiple of ten,"
@@ -129,11 +129,11 @@ def apply_rate(cur_time):
 
         if len(eligible) > 0:
             sampleRate = choice(eligible).rate #select random rate from eligible
-            return [(ieee80211_to_idx(sampleRate)[0], NRETRIES)]
+            return [(ieee80211_to_idx(sampleRate), NRETRIES)]
 
     #"Otherwise, send packet at the bit-rate that has the lowest avg transmission time"
     # Trusts that currRate is properly maintained to be lowest avgTX
-    return [(ieee80211_to_idx(currRate)[0], NRETRIES)]
+    return [(ieee80211_to_idx(currRate), NRETRIES)]
 
 
 #"When process f eedback() runs, it updates information that tracks

### 2013-08-06 pysim/samplerate_longer.py

@@ -23,7 +23,7 @@ backoff = {0:0, 1:155, 2:315, 3:635, 4:1275, 5:2555, 6:5115, 7:5115, 8:5115, 9:5
            18:5115, 19:5115, 20:5115}
 
 def bitrate_type(bitrate):
-    return common.RATES[ieee80211_to_idx(bitrate)[0]].phy
+    return common.RATES[ieee80211_to_idx(bitrate)].phy
 
 
 
@@ -115,7 +115,7 @@ def apply_rate(cur_time):
         for i, r in sorted(rates.items(), reverse=True):
             if r.succFails < MAXFAILS:
                 currRate = r.rate
-                return [(ieee80211_to_idx(currRate)[0], NRETRIES)]
+                return [(ieee80211_to_idx(currRate), NRETRIES)]
 
     # Every 10 packets, select a random non-failing bit rate w/ better avg tx
     #"If the number of packets sent over the link is a multiple of ten,"
@@ -131,11 +131,11 @@ def apply_rate(cur_time):
 
         if len(eligible) > 0:
             sampleRate = choice(eligible).rate #select random rate from eligible
-            return [(ieee80211_to_idx(sampleRate)[0], NRETRIES)]
+            return [(ieee80211_to_idx(sampleRate), NRETRIES)]
 
     #"Otherwise, send packet at the bit-rate that has the lowest avg transmission time"
     # Trusts that currRate is properly maintained to be lowest avgTX
-    return [(ieee80211_to_idx(currRate)[0], NRETRIES)]
+    return [(ieee80211_to_idx(currRate), NRETRIES)]
 
 
 #"When process f eedback() runs, it updates information that tracks

### 2013-08-06 pysim/samplerate_retry.py

@@ -21,7 +21,7 @@ backoff = {0:0, 1:155, 2:315, 3:635, 4:1275, 5:2555, 6:5115, 7:5115, 8:5115, 9:5
            18:5115, 19:5115, 20:5115}
 
 def bitrate_type(bitrate):
-    return common.RATES[ieee80211_to_idx(bitrate)[0]].phy
+    return common.RATES[ieee80211_to_idx(bitrate)].phy
 
 
 
@@ -116,7 +116,7 @@ def apply_rate(cur_time):
         for r in rrates:
             if r.succFails < 4:
                 currRate = r.rate
-                retry.append((ieee80211_to_idx(currRate)[0], NRETRIES))
+                retry.append((ieee80211_to_idx(currRate), NRETRIES))
         return retry
 
     # Every 10 packets, select a random non-failing bit rate w/ better avg tx
@@ -133,11 +133,11 @@ def apply_rate(cur_time):
 
         if len(eligible) > 0:
             sampleRate = choice(eligible).rate #select random rate from eligible
-            return [(ieee80211_to_idx(sampleRate)[0], NRETRIES)]
+            return [(ieee80211_to_idx(sampleRate), NRETRIES)]
 
     #"Otherwise, send packet at the bit-rate that has the lowest avg transmission time"
     # Trusts that currRate is properly maintained to be lowest avgTX
-    return [(ieee80211_to_idx(currRate)[0], NRETRIES)]
+    return [(ieee80211_to_idx(currRate), NRETRIES)]
 
 
 #"When process f eedback() runs, it updates information that tracks

### 2013-08-07 pysim/common.py
Cleanups of whitespace and some comments

@@ -119,7 +119,7 @@ def tx_time(rix, length=1200): #rix is index to RATES, length in bytes
         dur = 16 # SIFS + signal ext */
         dur += 16 # 17.3.2.3: T_PREAMBLE = 16 usec */
         dur += 4 # 17.3.2.3: T_SIGNAL = 4 usec */
-        dur += 4* (round((16+8*(length+4)+6)/(4*mbps))+1) # T_SYM x N_SYM 
+        dur += 4 * (round((16+8*(length+4)+6)/(4*mbps))+1) # T_SYM x N_SYM
 
     else:
         '''
@@ -132,7 +132,7 @@ def tx_time(rix, length=1200): #rix is index to RATES, length in bytes
         * aPreambleLength = 144 usec or 72 usec with short preamble
         * aPLCPHeaderLength = 48 usec or 24 usec with short preamble
         *'''
-        dur = 10 # aSIFSTime = 10 usec 
+        dur = 10 # aSIFSTime = 10 usec
         dur += (72 + 24) #using short preamble, otw we'd use (144 + 48)
         dur += round((8*(length + 4))/mbps)+1
     

### 2013-08-07 pysim/constant.py

@@ -3,6 +3,7 @@ import common
 
 RATE=float(os.environ["RATE"]) if "RATE" in os.environ else 1
 
+# Read the rate as a Mbps value and convert it to an index
 try:
     IDX = common.ieee80211_to_idx(RATE)
 except ValueError:

### 2013-08-07 pysim/minstrel.py

@@ -3,7 +3,6 @@
 # 3.3.8 linux kernel. We assume multi-rate retry capabilities, so we omit 
 # the code for the non-mrr case. 
 
-from random import randint
 from random import choice 
 import common
 from common import ieee80211_to_idx
@@ -44,41 +43,7 @@ odfm = set([6, 9, 12, 18, 24, 36, 48, 54])
 * also include SIFS.
 */'''
 def tx_time(rate, length): #rate is Mbps, length in bytes
-    if rate in odfm:
-        '''* OFDM:
-        *
-        * N_DBPS = DATARATE x 4
-        * N_SYM = Ceiling((16+8xLENGTH+6) / N_DBPS)
-        *	(16 = SIGNAL time, 6 = tail bits)
-        * TXTIME = T_PREAMBLE + T_SIGNAL + T_SYM x N_SYM + Signal Ext
-        *
-        * T_SYM = 4 usec
-        * 802.11a - 17.5.2: aSIFSTime = 16 usec
-        * 802.11g - 19.8.4: aSIFSTime = 10 usec +
-        *	signal ext = 6 usec
-        */'''
-        dur = 16 # SIFS + signal ext */
-        dur += 16 # 17.3.2.3: T_PREAMBLE = 16 usec */
-        dur += 4 # 17.3.2.3: T_SIGNAL = 4 usec */
-        dur += 4* (round((16+8*(length+4)+6)/(4*rate))+1) # T_SYM x N_SYM 
-
-    else:
-        '''
-        * 802.11b or 802.11g with 802.11b compatibility:
-        * 18.3.4: TXTIME = PreambleLength + PLCPHeaderTime +
-        * Ceiling(((LENGTH+PBCC)x8)/DATARATE). PBCC=0.
-        *
-        * 802.11 (DS): 15.3.3, 802.11b: 18.3.4
-        * aSIFSTime = 10 usec
-        * aPreambleLength = 144 usec or 72 usec with short preamble
-        * aPLCPHeaderLength = 48 usec or 24 usec with short preamble
-        *'''
-        dur = 10 # aSIFSTime = 10 usec 
-        dur += (72 + 24) #using short preamble, otw we'd use (144 + 48)
-        dur += round((8*(length + 4))/rate)+1
-    
-    return dur
-
+    return common.tx_time(ieee80211_to_idx(rate), length)
 
 class Packet:
     def __init__(self, time_sent, success, txTime, rate):
[…]

### 2013-08-07 pysim/minstrel.py
Clean up MRR handling in Minstrel

@@ -170,14 +170,7 @@ def apply_rate(cur_time): #cur_time is in nanoseconds
             random = choice(list(rates))
 
         if random < bestThruput:
-            r = [(ieee80211_to_idx(bestThruput),
-                     rates[bestThruput].adjusted_retry_count),
-                    (ieee80211_to_idx(random),
-                     rates[random].adjusted_retry_count), 
-                    (ieee80211_to_idx(bestProb),
-                     rates[bestProb].adjusted_retry_count), 
-                    (ieee80211_to_idx(lowestRate),
-                     rates[lowestRate].adjusted_retry_count)]
+            chain = [bestThruput, random, bestProb, lowestRate]
             ''' FROM KERNEL:
             /* Only use IEEE80211_TX_CTL_RATE_CTRL_PROBE to mark        
             * packets that have the sampling rate deferred to the      
@@ -195,24 +188,14 @@ def apply_rate(cur_time): #cur_time is in nanoseconds
                 if rates[random].sample_limit > 0:
                     rates[random].sample_limit -= 1
             
-            r = [(ieee80211_to_idx(random),
-                      rates[random].adjusted_retry_count), 
-                    (ieee80211_to_idx(bestThruput),
-                     rates[bestThruput].adjusted_retry_count),
-                    (ieee80211_to_idx(bestProb),
-                     rates[bestProb].adjusted_retry_count), 
-                    (ieee80211_to_idx(lowestRate),
-                     rates[lowestRate].adjusted_retry_count)]
+            chain = [random, bestThruput, bestProb, lowestRate]
     
     else:     #normal
-        r = [(ieee80211_to_idx(bestThruput),
-              rates[bestThruput].adjusted_retry_count), 
-             (ieee80211_to_idx(nextThruput),
-              rates[nextThruput].adjusted_retry_count), 
-             (ieee80211_to_idx(bestProb),
-              rates[bestProb].adjusted_retry_count), 
-             (ieee80211_to_idx(lowestRate),
-              rates[lowestRate].adjusted_retry_count)]
+        chain = [bestThruput, nextThruput, bestProb, lowestRate]
+
+    mrr = [(ieee80211_to_idx(rate), rates[rate].adjusted_retry_count)
+           for rate in chain]
+
     return r
         
 #status: true if packet was rcvd successfully

### 2013-08-07 pysim/common.py
Wrap comments properly.

@@ -97,7 +97,7 @@ class BalancedEWMA:
     def read(self):
         if self.val:
             return int(self.val * 18000)
-        else: 
+        else:
             return None
 
 def tx_time(rix, length=1200): #rix is index to RATES, length in bytes

### 2013-08-07 pysim/samplerate.py

@@ -14,8 +14,8 @@ NBYTES = 1500 #constant
 currRate = 54 #current best bitRate
 NRETRIES = 1
 
-# The average back-off period, in microseconds, for up to 8 attempts of a 802.11b unicast packet. 
-# TODO: find g data
+# The average back-off period, in microseconds, for up to 8 attempts
+# of a 802.11b unicast packet.  TODO: find g data
 backoff = {0:0, 1:155, 2:315, 3:635, 4:1275, 5:2555, 6:5115, 7:5115, 8:5115, 9:5115,
            10:5115, 11:5115, 12:5115, 13:5115, 14:5115, 15:5115, 16:5115, 17:5115,
            18:5115, 19:5115, 20:5115}
@@ -25,10 +25,11 @@ def bitrate_type(bitrate):
 
 
 
-#"To calculate the transmission time of a n-byte unicast packet given the bit-rate b and
-# number of retries r, SampleRate uses the following equation based on the 802.11 unicast
-# retransmission mechanism detailed in Section 2.2"
-#
+#"To calculate the transmission time of a n-byte unicast packet given
+# the bit-rate b and number of retries r, SampleRate uses the
+# following equation based on the 802.11 unicast retransmission
+# mechanism detailed in Section 2.2"
+
 # tx_time(b, r, n) =  difs + backoff[r] + (r + 1)*(sifs + ack + header + (n * 8/b))
 def tx_time(bitrate, retries, nbytes):
     # bitrate in MBPS, since 1*10^6 bps / 10-6 seconds/microseconds = 1 bit per microsecond
@@ -92,11 +93,13 @@ class Rate:
                    self.totalTX, self.avgTX, self.losslessTX))
 
 
-# The modulation scheme used in 802.11g is orthogonal frequency-division multiplexing (OFDM)
-# copied from 802.11a with data rates of 6, 9, 12, 18, 24, 36, 48, and 54 Mbit/s, and reverts 
-# to CCK (like the 802.11b standard) for 5.5 and 11 Mbit/s and DBPSK/DQPSK+DSSS for 1 and 2 Mbit/s.
-# Even though 802.11g operates in the same frequency band as 802.11b, it can achieve higher 
-# data rates because of its heritage to 802.11a.
+# The modulation scheme used in 802.11g is orthogonal
+# frequency-division multiplexing (OFDM) copied from 802.11a with data
[…]

### 2013-08-07 pysim/samplerate.py
Removed some imports

@@ -4,7 +4,7 @@
 
 from __future__ import division
 
-from random import choice 
+import random
 from common import ieee80211_to_idx
 import common
 
@@ -66,7 +66,6 @@ class Packet:
     def __repr__(self):
         return ("Pkt sent at time %r, rate %r was successful: %r\n" 
                 % (self.time_sent, self.rate, self.success))
-
 
 class Rate:
     def __init__(self, rate):
@@ -131,7 +130,7 @@ def apply_rate(cur_time):
                     if r.losslessTX < cavgTX and r.succFails < 4]
 
         if len(eligible) > 0:
-            sampleRate = choice(eligible).rate #select random rate from eligible
+            sampleRate = random.choice(eligible).rate #select random rate from eligible
             return [(ieee80211_to_idx(sampleRate), NRETRIES)]
 
     #"Otherwise, send packet at the bit-rate that has the lowest avg

### 2013-08-07 pysim/minstrel.py
Named Tuples are great

@@ -196,7 +196,7 @@ def apply_rate(cur_time): #cur_time is in nanoseconds
     mrr = [(ieee80211_to_idx(rate), rates[rate].adjusted_retry_count)
            for rate in chain]
 
-    return r
+    return mrr
         
 #status: true if packet was rcvd successfully
 #timestamp: time pkt was sent

### 2013-08-07 pysim/samplerate.py

@@ -5,8 +5,9 @@
 from __future__ import division
 
 import random
-from common import ieee80211_to_idx
 import common
+import collections
+from common import ieee80211_to_idx
 
 npkts = 0 #number of packets sent over link
 nsuccess = 0 #number of packets sent successfully 
@@ -56,16 +57,8 @@ def tx_time(bitrate, retries, nbytes):
 
     return difs + backoff[retries] + (retries+1)*(sifs + ack + header + (nbytes * 8/(bitrate)))
 
-class Packet:
-    def __init__(self, time_sent, success, txTime, rate):
-        self.time_sent = time_sent
-        self.success = success
-        self.txTime = txTime
-        self.rate = rate
-
-    def __repr__(self):
-        return ("Pkt sent at time %r, rate %r was successful: %r\n" 
-                % (self.time_sent, self.rate, self.success))
+Packet = collections.namedtuple("Packet", ["time_sent", "success",
+                                           "txTime", "rate"])
 
 class Rate:
     def __init__(self, rate):

### 2013-08-07 pysim/minstrel_bewma.py
Split out only the BEWMA change

@@ -1,0 +1,292 @@
+# Colleen Josephson, 2013
+# This file attempts to implement the minstrel rate control algorithm from the
+# 3.3.8 linux kernel. We assume multi-rate retry capabilities, so we omit 
+# the code for the non-mrr case. 
+
+from random import choice 
+import common
+from common import ieee80211_to_idx
+
+npkts = 0 #number of packets sent over link
+nsuccess = 0 #number of packets sent successfully 
+nlookaround = 0
+NBYTES = 1500 #constant
+currRate = 54  #current best bitRate
+NRETRIES = 20
+bestThruput = 12
+nextThruput = 11
+bestProb = 2
+lowestRate = 1
+time_last_called = 0
+cw_min = 15
+cw_max = 1023
+segment_size = 6000
+max_retry = 7 #safe default specified in kernel code
+np = 0
+nd = 0
+probeFlag = False
+
+# The average back-off period, in milliseconds, for up to 8 attempts of a 802.11b unicast packet. 
+# TODO: find g data
+backoff = {0:0, 1:.155, 2:.315, 3:.635, 4:1.275, 5:2.555, 6:5.115, 7:5.115, 8:5.115, 
+9:5.115, 10:5.115, 11:5.115, 12:5.115, 13:5.115, 14:5.115, 15:5.115, 16:5.115, 
+17:5.115, 18:5.115, 19:5.115, 20:5.115}
+#802.11g rates
+odfm = set([6, 9, 12, 18, 24, 36, 48, 54])
+
+'''
+/* Adapted from 802.11 util.c ieee80211_frame_duration()
+*
+* calculate duration (in microseconds, rounded up to next higher
+* integer if it includes a fractional microsecond) to send frame of
+* len bytes (does not include FCS) at the given rate. Duration will
+* also include SIFS.
+*/'''
+def tx_time(rate, length): #rate is Mbps, length in bytes
+    return common.tx_time(ieee80211_to_idx(rate), length)
+
+class Packet:
+    def __init__(self, time_sent, success, txTime, rate):
+        self.time_sent = time_sent
+        self.success = success
+        self.txTime = txTime
+        self.rate = rate
+
+    def __repr__(self):
+        return ("Pkt sent at time %r, rate %r was successful: %r\n" 
+                % (self.time_sent, self.rate, self.success))
+
+
+class Rate:
+    #OFDM = g
+    def __init__(self, rate):
+        self.rate = rate #in mbps
[…]

### 2013-08-07 pysim/common.py
Modified harness to use the SampleRate tx_time function
The new tx_time properly handles acks, backoff, and the DIFS.
This also lead to relabeling the 802.11 and 802.11b rates as DS and DSSS.

@@ -6,10 +6,10 @@ BitRate = collections.namedtuple("BitRate", ["phy", "kbps", "user_kbps",
                                              "code", "dot11_rate"])
 
 RATES = [
-    BitRate("cck", 1000, 900, 0, 2),
-    BitRate("cck", 2000, 1900, 1, 4),
-    BitRate("cck", 5500, 4900, 2, 11),
-    BitRate("cck", 11000, 8100, 3, 22),
+    BitRate("ds", 1000, 900, 0, 2),
+    BitRate("ds", 2000, 1900, 1, 4),
+    BitRate("dsss", 5500, 4900, 2, 11),
+    BitRate("dsss", 11000, 8100, 3, 22),
     BitRate("ofdm", 6000, 5400, 4, 12),
     BitRate("ofdm", 9000, 7800, 5, 18),
     BitRate("ofdm", 12000, 10100, 6, 24),
@@ -20,18 +20,18 @@ RATES = [
     BitRate("ofdm", 54000, 30900, 11, 108),
 
     # Ignored but here for completeness
-    BitRate("ht_ss", 6500, 6400, 0, 0),
-    BitRate("ht_ss", 13000, 12700, 1, 1),
-    BitRate("ht_ss", 19500, 18800, 2, 2),
-    BitRate("ht_ss", 26000, 25000, 3, 3),
-    BitRate("ht_ss", 39000, 36700, 4, 4),
-    BitRate("ht_ss", 52000, 48100, 5, 5),
-    BitRate("ht_ss", 58500, 53500, 6, 6),
-    BitRate("ht_ss", 65000, 59000, 7, 7),
-    BitRate("ht_hgi", 72200, 65400, 7, 7),
-    BitRate("ht_ds", 13000, 12700, 8, 8),
-    BitRate("ht_ds", 26000, 24800, 9, 9),
-    BitRate("ht_ds", 39000, 36600, 10, 10),
+    #BitRate("ht_ss", 6500, 6400, 0, 0),
+    #BitRate("ht_ss", 13000, 12700, 1, 1),
+    #BitRate("ht_ss", 19500, 18800, 2, 2),
+    #BitRate("ht_ss", 26000, 25000, 3, 3),
+    #BitRate("ht_ss", 39000, 36700, 4, 4),
+    #BitRate("ht_ss", 52000, 48100, 5, 5),
+    #BitRate("ht_ss", 58500, 53500, 6, 6),
[…]

### 2013-08-07 pysim/harness.py

@@ -82,7 +82,7 @@ class Harness:
 
             tot_tries.append((rate, s_tries))
             self.histogram[rate] += s_tries
-            tot_delay += s_tries * common.tx_time(rate, 1500)
+            tot_delay += common.tx_time(rate, s_tries, 1500)
 
             if succeeded:
                 tot_status = True

### 2013-08-07 pysim/minstrel.py

@@ -5,6 +5,7 @@
 
 from random import choice 
 import common
+import math
 from common import ieee80211_to_idx
 
 npkts = 0 #number of packets sent over link
@@ -26,24 +27,52 @@ np = 0
 nd = 0
 probeFlag = False
 
-# The average back-off period, in milliseconds, for up to 8 attempts of a 802.11b unicast packet. 
-# TODO: find g data
-backoff = {0:0, 1:.155, 2:.315, 3:.635, 4:1.275, 5:2.555, 6:5.115, 7:5.115, 8:5.115, 
-9:5.115, 10:5.115, 11:5.115, 12:5.115, 13:5.115, 14:5.115, 15:5.115, 16:5.115, 
-17:5.115, 18:5.115, 19:5.115, 20:5.115}
-#802.11g rates
-odfm = set([6, 9, 12, 18, 24, 36, 48, 54])
+def tx_time(mbps, length=1200): #rix is index to RATES, length in bytes
+    '''
+    Adapted from 802.11 util.c ieee80211_frame_duration()
 
-'''
-/* Adapted from 802.11 util.c ieee80211_frame_duration()
-*
-* calculate duration (in microseconds, rounded up to next higher
-* integer if it includes a fractional microsecond) to send frame of
-* len bytes (does not include FCS) at the given rate. Duration will
-* also include SIFS.
-*/'''
-def tx_time(rate, length): #rate is Mbps, length in bytes
-    return common.tx_time(ieee80211_to_idx(rate), length)
+    calculate duration (in microseconds, rounded up to next higher
+    integer if it includes a fractional microsecond) to send frame of
+    len bytes (does not include FCS) at the given rate. Duration will
+    also include SIFS.
+    '''
+
+    rateinfo = common.RATES[common.ieee80211_to_idx(mbps)]
+
+    if rateinfo.phy == "ofdm":
+        '''
+        OFDM:
+        N_DBPS = DATARATE x 4
+        N_SYM = Ceiling((16+8xLENGTH+6) / N_DBPS)
+        (16 = SIGNAL time, 6 = tail bits)
+        TXTIME = T_PREAMBLE + T_SIGNAL + T_SYM x N_SYM + Signal Ext
+
+        T_SYM = 4 usec
+        802.11a - 17.5.2: aSIFSTime = 16 usec
+        802.11g - 19.8.4: aSIFSTime = 10 usec +
+        signal ext = 6 usec
[…]

### 2013-08-07 pysim/samplerate.py

@@ -15,47 +15,41 @@ NBYTES = 1500 #constant
 currRate = 54 #current best bitRate
 NRETRIES = 1
 
-# The average back-off period, in microseconds, for up to 8 attempts
-# of a 802.11b unicast packet.  TODO: find g data
-backoff = {0:0, 1:155, 2:315, 3:635, 4:1275, 5:2555, 6:5115, 7:5115, 8:5115, 9:5115,
-           10:5115, 11:5115, 12:5115, 13:5115, 14:5115, 15:5115, 16:5115, 17:5115,
-           18:5115, 19:5115, 20:5115}
-
 def bitrate_type(bitrate):
     return common.RATES[ieee80211_to_idx(bitrate)].phy
 
-
+# The average back-off period, in microseconds, for up to 8 attempts
+# of a 802.11b unicast packet.
+# TODO: find g data
+backoff = [0, 155, 315, 635, 1275, 2555, 5115]
 
 #"To calculate the transmission time of a n-byte unicast packet given
 # the bit-rate b and number of retries r, SampleRate uses the
 # following equation based on the 802.11 unicast retransmission
 # mechanism detailed in Section 2.2"
 
-# tx_time(b, r, n) =  difs + backoff[r] + (r + 1)*(sifs + ack + header + (n * 8/b))
 def tx_time(bitrate, retries, nbytes):
-    # bitrate in MBPS, since 1*10^6 bps / 10-6 seconds/microseconds = 1 bit per microsecond
-    global currRate, npkts, nsuccess, NBYTES
+    # tx_time(b, r, n) = difs + backoff[r] + \
+    #                  + (r + 1)*(sifs + ack + header + (n * 8/b)
 
-    brtype = bitrate_type(bitrate)
-    if bitrate == 1:
-        difs = 50
-        sifs = 10
-        ack = 304
-        header = 192
-    elif brtype == "cck":
-        difs = 50
-        sifs = 10
-        ack = 304
-        header = 96
-    elif brtype == "ofdm":
-        difs = 28
-        sifs = 9
-        ack = 304 # Somehow 6mb acks aren't used
-        header = 20
-    else:
-        raise ValueError("Unknown bitrate type", brtype, bitrate)
+    #"where difs is 50 microseconds in 802.11b and 28 microseconds in
+    # 802.11a/g, sifs is 10 microseconds for 802.11b and 9 for
+    # 802.11a/g, and ack is 304 microseconds using 1 megabit
[…]

### 2013-08-07 pysim/harness.py
Print success percentage for each rate

@@ -53,7 +53,7 @@ class Harness:
         self.choose_rate = choose_rate
         self.push_statistics = push_statistics
 
-        self.histogram = [0] * len(common.RATES)
+        self.histogram = [[0, 0] for i in common.RATES]
 
     def send_packet(self):
         rate_arr = self.choose_rate(self.clock)
@@ -81,7 +81,8 @@ class Harness:
                     break
 
             tot_tries.append((rate, s_tries))
-            self.histogram[rate] += s_tries
+            self.histogram[rate][0] += s_tries
+            self.histogram[rate][1] += 1 if succeeded else 0
             tot_delay += common.tx_time(rate, s_tries, 1500)
 
             if succeeded:
@@ -160,8 +161,10 @@ if __name__ == "__main__":
     throughput = 1500 * 8 * good / (time / 1e9) / 1e6
     print("Average packet took {:.3f} ms / achieved {:.3f} Mbps".format(time / good / 1e6, throughput))
 
-    for rate_idx, tries in enumerate(harness.histogram):
+    for rate_idx, info in enumerate(harness.histogram):
+        tries, successes = info
         if not tries: continue
+
         mbps = common.RATES[rate_idx][-1] / 2
-        print("{} Mbps : {} tries".format(str(mbps).rjust(5),
-                                          str(int(tries)).rjust(4)))
+        print("{:>5} Mbps : {:>4} tries ({:.0%} success rate)".format(
+            mbps, tries, successes/tries))

### 2013-08-07 pysim/common.py
Modified harness to do backoff and difs better.
Now does backoff computation across MRRs.

@@ -103,27 +103,31 @@ class BalancedEWMA:
 # The average back-off period, in microseconds, for up to 7 attempts
 # of a 802.11b unicast packet.
 
-backoff = { "ofdm": [0], "ds": [0], "dsss": [0] }
+BACKOFF = { "ofdm": [0], "ds": [0], "dsss": [0] }
 for i in range(5, 11):
-    backoff["ds"].append(int(((2**i) - 1) * (20 / 2)))
+    BACKOFF["ds"].append(int(((2**i) - 1) * (20 / 2)))
 for i in range(5, 11):
-    backoff["dsss"].append(int(((2**i) - 1) * (9 / 2)))
+    BACKOFF["dsss"].append(int(((2**i) - 1) * (9 / 2)))
 for i in range(4, 11):
-    backoff["ofdm"].append(int(((2**i) - 1) * (9 / 2)))
+    BACKOFF["ofdm"].append(int(((2**i) - 1) * (9 / 2)))
+
+def backoff(rix, attempt):
+    return BACKOFF[RATES[rix].phy][min(attempt, len(BACKOFF) - 1)]
+
+def difs(rix):
+    version = "g" if RATES[rix].phy == "ofdm" else "b"
+    return 50 if version == "b" else 28
 
 def tx_time(rix, tries, nbytes):
     # From the SampleRate paper.  See samplerate.py for annotated version.
     bitrate = RATES[rix].dot11_rate / 2
     version = "g" if RATES[rix].phy == "ofdm" else "b"
-    difs = 50 if version == "b" else 28
     sifs = 10 if version == "b" else 9
     ack = 304 # Somehow 6mb acks aren't used
     header = 192 if bitrate == 1 else 96 if version == "b" else 20
     
     assert tries > 0, "Cannot try a non-positive number of times"
-    backoff_r = backoff[RATES[rix].phy][min(tries - 1, len(backoff) - 1)]
 
-    us = difs + backoff_r + \
-        tries * (sifs + ack + header + (nbytes * 8 / bitrate))
+    us = tries * (sifs + ack + header + (nbytes * 8 / bitrate))
     return us * 1000
 

### 2013-08-07 pysim/harness.py

@@ -55,6 +55,23 @@ class Harness:
 
         self.histogram = [[0, 0] for i in common.RATES]
 
+        self.attempts = 0
+
+    def send_one(self, rate, is_success):
+        delay = common.tx_time(rate, 1, 1500)
+
+        if is_success:
+            delay += common.difs(rate)
+            self.attempts = 0
+        else:
+            self.attempts += 1
+            delay += common.backoff(rate, self.attempts)
+
+        self.histogram[rate][0] += 1
+        self.histogram[rate][1] += 1 if is_success else 0
+
+        return delay
+
     def send_packet(self):
         rate_arr = self.choose_rate(self.clock)
 
@@ -75,15 +92,13 @@ class Harness:
             for i in range(tries):
                 success = random.random() < p_success
                 s_tries += 1
+                tot_delay += self.send_one(rate, success)
 
                 if success:
                     succeeded = True
                     break
 
             tot_tries.append((rate, s_tries))
-            self.histogram[rate][0] += s_tries
-            self.histogram[rate][1] += 1 if succeeded else 0
-            tot_delay += common.tx_time(rate, s_tries, 1500)
 
             if succeeded:
                 tot_status = True

### 2013-08-07 pysim/minstrel_bewma.py

@@ -5,6 +5,7 @@
 
 from random import choice 
 import common
+import math
 from common import ieee80211_to_idx
 
 npkts = 0 #number of packets sent over link
@@ -26,24 +27,52 @@ np = 0
 nd = 0
 probeFlag = False
 
-# The average back-off period, in milliseconds, for up to 8 attempts of a 802.11b unicast packet. 
-# TODO: find g data
-backoff = {0:0, 1:.155, 2:.315, 3:.635, 4:1.275, 5:2.555, 6:5.115, 7:5.115, 8:5.115, 
-9:5.115, 10:5.115, 11:5.115, 12:5.115, 13:5.115, 14:5.115, 15:5.115, 16:5.115, 
-17:5.115, 18:5.115, 19:5.115, 20:5.115}
-#802.11g rates
-odfm = set([6, 9, 12, 18, 24, 36, 48, 54])
+def tx_time(mbps, length=1200): #rix is index to RATES, length in bytes
+    '''
+    Adapted from 802.11 util.c ieee80211_frame_duration()
 
-'''
-/* Adapted from 802.11 util.c ieee80211_frame_duration()
-*
-* calculate duration (in microseconds, rounded up to next higher
-* integer if it includes a fractional microsecond) to send frame of
-* len bytes (does not include FCS) at the given rate. Duration will
-* also include SIFS.
-*/'''
-def tx_time(rate, length): #rate is Mbps, length in bytes
-    return common.tx_time(ieee80211_to_idx(rate), length)
+    calculate duration (in microseconds, rounded up to next higher
+    integer if it includes a fractional microsecond) to send frame of
+    len bytes (does not include FCS) at the given rate. Duration will
+    also include SIFS.
+    '''
+
+    rateinfo = common.RATES[common.ieee80211_to_idx(mbps)]
+
+    if rateinfo.phy == "ofdm":
+        '''
+        OFDM:
+        N_DBPS = DATARATE x 4
+        N_SYM = Ceiling((16+8xLENGTH+6) / N_DBPS)
+        (16 = SIGNAL time, 6 = tail bits)
+        TXTIME = T_PREAMBLE + T_SIGNAL + T_SYM x N_SYM + Signal Ext
+
+        T_SYM = 4 usec
+        802.11a - 17.5.2: aSIFSTime = 16 usec
+        802.11g - 19.8.4: aSIFSTime = 10 usec +
+        signal ext = 6 usec
+        '''
[…]

### 2013-08-07 pysim/samplerate.py
Minor shuffles in samplerate

@@ -9,11 +9,13 @@ import common
 import collections
 from common import ieee80211_to_idx
 
-npkts = 0 #number of packets sent over link
-nsuccess = 0 #number of packets sent successfully 
-NBYTES = 1500 #constant
-currRate = 54 #current best bitRate
+# Constants: send 1500 bytes at a time, with 1 try each in the MRR
+NBYTES = 1500
 NRETRIES = 1
+
+npkts = 0 # number of packets sent over link
+nsuccess = 0 #number of packets sent successfully 
+currRate = 54 #current best bitRate
 
 def bitrate_type(bitrate):
     return common.RATES[ieee80211_to_idx(bitrate)].phy
@@ -90,7 +92,7 @@ rates = dict((r, Rate(r)) for r in [1, 2, 5.5, 6, 9, 11, 12, 18, 24, 36, 48, 54]
 
 #multi-rate retry returns an array of (rate, ntries) for the next n packets
 def apply_rate(cur_time):
-    global currRate, npkts, nsuccess, NBYTES, NRETRIES
+    global currRate, npkts, nsuccess
     remove_stale_results(cur_time)
     
     #"Increment the number of packets sent over the link"
@@ -237,7 +239,7 @@ def remove_stale_results(cur_time):
         
 
 def calculateMin():
-    global currRate, npkts, nsuccess, NBYTES
+    global currRate, npkts, nsuccess
 
     #set current rate to the one w/ min avg tx time
     c = rates[currRate]

### 2013-08-07 pysim/common.py
Make dot11_rate a float with the actual rate

@@ -5,38 +5,24 @@ import math
 BitRate = collections.namedtuple("BitRate", ["phy", "kbps", "user_kbps",
                                              "code", "dot11_rate"])
 
+# The 802.11, 802.11b, and 802.11g rates
 RATES = [
-    BitRate("ds", 1000, 900, 0, 2),
-    BitRate("ds", 2000, 1900, 1, 4),
-    BitRate("dsss", 5500, 4900, 2, 11),
-    BitRate("dsss", 11000, 8100, 3, 22),
-    BitRate("ofdm", 6000, 5400, 4, 12),
-    BitRate("ofdm", 9000, 7800, 5, 18),
-    BitRate("ofdm", 12000, 10100, 6, 24),
-    BitRate("ofdm", 18000, 14100, 7, 36),
-    BitRate("ofdm", 24000, 17700, 8, 48),
-    BitRate("ofdm", 36000, 23700, 9, 72),
-    BitRate("ofdm", 48000, 27400, 10, 96),
-    BitRate("ofdm", 54000, 30900, 11, 108),
-
-    # Ignored but here for completeness
-    #BitRate("ht_ss", 6500, 6400, 0, 0),
-    #BitRate("ht_ss", 13000, 12700, 1, 1),
-    #BitRate("ht_ss", 19500, 18800, 2, 2),
-    #BitRate("ht_ss", 26000, 25000, 3, 3),
-    #BitRate("ht_ss", 39000, 36700, 4, 4),
-    #BitRate("ht_ss", 52000, 48100, 5, 5),
-    #BitRate("ht_ss", 58500, 53500, 6, 6),
-    #BitRate("ht_ss", 65000, 59000, 7, 7),
-    #BitRate("ht_hgi", 72200, 65400, 7, 7),
-    #BitRate("ht_ds", 13000, 12700, 8, 8),
-    #BitRate("ht_ds", 26000, 24800, 9, 9),
-    #BitRate("ht_ds", 39000, 36600, 10, 10),
+    BitRate("ds",    1000,   900,  0, 1.0),
+    BitRate("ds",    2000,  1900,  1, 2.0),
+    BitRate("dsss",  5500,  4900,  2, 5.5),
+    BitRate("dsss", 11000,  8100,  3, 11.0),
+    BitRate("ofdm",  6000,  5400,  4, 6.0),
+    BitRate("ofdm",  9000,  7800,  5, 9.0),
[…]

### 2013-08-07 pysim/constant.py

@@ -7,7 +7,7 @@ RATE=float(os.environ["RATE"]) if "RATE" in os.environ else 1
 try:
     IDX = common.ieee80211_to_idx(RATE)
 except ValueError:
-    opts = [str(rate.dot11_rate / 2) for rate in common.RATES]
+    opts = [str(rate.dot11_rate) for rate in common.RATES]
     print("Invalid rate.  Options are: {}".format(", ".join(opts)))
     exit()
 else:

### 2013-08-07 pysim/harness.py

@@ -180,6 +180,6 @@ if __name__ == "__main__":
         tries, successes = info
         if not tries: continue
 
-        mbps = common.RATES[rate_idx][-1] / 2
+        mbps = common.RATES[rate_idx].dot11_rate
         print("{:>5} Mbps : {:>4} tries ({:.0%} success rate)".format(
             mbps, tries, successes/tries))

### 2013-08-07 pysim/minproved.py

@@ -234,7 +234,7 @@ def process_feedback(status, timestamp, delay, tries):
     for t in range(len(tries)):
         (bitrate, br_tries) = tries[t]
         if br_tries > 0:
-            bitrate = common.RATES[bitrate][-1]/2.0
+            bitrate = common.RATES[bitrate].dot11_rate
             #if bitrate == 1:
             
             br = rates[bitrate]

### 2013-08-07 pysim/minstrel_bewma.py

@@ -237,7 +237,7 @@ def process_feedback(status, timestamp, delay, tries):
     for t in range(len(tries)):
         (bitrate, br_tries) = tries[t]
         if br_tries > 0:
-            bitrate = common.RATES[bitrate][-1]/2.0
+            bitrate = common.RATES[bitrate].dot11_rate
             #if bitrate == 1:
             
             br = rates[bitrate]

### 2013-08-07 pysim/samplerate.py

@@ -140,7 +140,7 @@ def process_feedback(status, timestamp, delay, tries):
     global currRate, npkts, nsuccess, NBYTES
     (bitrate, nretries) = tries[0]
     nretries -= 1
-    bitrate = common.RATES[bitrate].dot11_rate/2.0
+    bitrate = common.RATES[bitrate].dot11_rate
 
     #"Calculate the transmission time for the packet based on the
     # bit-rate and number of retries using Equation 5.1 below."

### 2013-08-07 pysim/samplerate_longer.py

@@ -150,7 +150,7 @@ def process_feedback(status, timestamp, delay, tries):
     global currRate, npkts, nsuccess, NBYTES
     (bitrate, nretries) = tries[0]
     nretries -= 1
-    bitrate = common.RATES[bitrate].dot11_rate/2.0
+    bitrate = common.RATES[bitrate].dot11_rate
 
     #"Calculate the transmission time for the packet based on the
     # bit-rate and number of retries using Equation 5.1 below."

### 2013-08-07 pysim/samplerate_retry.py

@@ -152,7 +152,7 @@ def process_feedback(status, timestamp, delay, tries):
     global currRate, npkts, nsuccess, NBYTES
     (bitrate, nretries) = tries[0]
     nretries -= 1
-    bitrate = common.RATES[bitrate].dot11_rate/2.0
+    bitrate = common.RATES[bitrate].dot11_rate
 
     #"Calculate the transmission time for the packet based on the
     # bit-rate and number of retries using Equation 5.1 below."

### 2013-08-07 pysim/minstrel.py
Cleanups to minstrel algorithm

@@ -3,17 +3,17 @@
 # 3.3.8 linux kernel. We assume multi-rate retry capabilities, so we omit 
 # the code for the non-mrr case. 
 
-from random import choice 
+from __future__ import division
+
+import random
 import common
 import math
 from common import ieee80211_to_idx
 
-npkts = 0 #number of packets sent over link
+packet_count = 0 #number of packets sent over link
 nsuccess = 0 #number of packets sent successfully 
 nlookaround = 0
-NBYTES = 1500 #constant
 currRate = 54  #current best bitRate
-NRETRIES = 20
 bestThruput = 12
 nextThruput = 11
 bestProb = 2
@@ -23,9 +23,12 @@ cw_min = 15
 cw_max = 1023
 segment_size = 6000
 max_retry = 7 #safe default specified in kernel code
-np = 0
-nd = 0
 probeFlag = False
+
+sample_count = 0
+sample_deferred = 0
+
+SAMPLING_RATIO = 10
 
 def tx_time(mbps, length=1200): #rix is index to RATES, length in bytes
     '''
@@ -49,8 +52,7 @@ def tx_time(mbps, length=1200): #rix is index to RATES, length in bytes
 
         T_SYM = 4 usec
         802.11a - 17.5.2: aSIFSTime = 16 usec
-        802.11g - 19.8.4: aSIFSTime = 10 usec +
-        signal ext = 6 usec
+        802.11g - 19.8.4: aSIFSTime = 10 usec + signal ext = 6 usec
         '''
         dur = 16 # SIFS + signal ext */
         dur += 16 # 17.3.2.3: T_PREAMBLE = 16 usec */
@@ -74,18 +76,6 @@ def tx_time(mbps, length=1200): #rix is index to RATES, length in bytes
 
     return dur
 
-class Packet:
-    def __init__(self, time_sent, success, txTime, rate):
-        self.time_sent = time_sent
-        self.success = success
-        self.txTime = txTime
-        self.rate = rate
-
-    def __repr__(self):
-        return ("Pkt sent at time %r, rate %r was successful: %r\n" 
-                % (self.time_sent, self.rate, self.success))
-
-
 class Rate:
     #OFDM = g
     def __init__(self, rate):
@@ -100,7 +90,6 @@ class Rate:
         self.attempts = 0 #number of attempted transmissions in cur time interval
[…]

### 2013-08-07 pysim/minstrel.py
More Minstrel cleanups

@@ -145,22 +145,25 @@ def apply_rate(cur_time): #cur_time is in nanoseconds
         update_stats(cur_time)
         time_last_called = cur_time
 
-    #Minstrel spends a particular percentage of frames, doing "look around" i.e. 
-    #randomly trying other rates, to gather statistics. The percentage of 
-    #"look around" frames defaults to 10%. The distribution of lookaround frames is
-    #also randomized somewhat to avoid any potential "strobing" of lookaround 
-    #between similar nodes.
+    #"Minstrel spends a particular percentage of frames, doing "look
+    # around" i.e.  randomly trying other rates, to gather
+    # statistics. The percentage of "look around" frames defaults to
+    # 10%. The distribution of lookaround frames is also randomized
+    # somewhat to avoid any potential "strobing" of lookaround between
+    # similar nodes."
     
-    #Try |         Lookaround rate              | Normal rate
-    #    | random < best    | random > best     |
-    #--------------------------------------------------------------
-    # 1  | Best throughput  | Random rate       | Best throughput
-    # 2  | Random rate      | Best throughput   | Next best throughput
-    # 3  | Best probability | Best probability  | Best probability
-    # 4  | Lowest Baserate  | Lowest baserate   | Lowest baserate
-    delta = (packet_count * SAMPLING_RATIO / 100) - (sample_count + sample_deferred // 2)
+    # Try |         Lookaround rate              | Normal rate
+    #     | random < best    | random > best     |
+    # --------------------------------------------------------------
+    #  1  | Best throughput  | Random rate       | Best throughput
+    #  2  | Random rate      | Best throughput   | Next best throughput
+    #  3  | Best probability | Best probability  | Best probability
+    #  4  | Lowest Baserate  | Lowest baserate   | Lowest baserate
 
-    if delta > 0:
+    delta = (packet_count * SAMPLING_RATIO // 100) - \
+            (sample_count + sample_deferred // 2)
+
+    if delta > 0: # In this case we attempt to sample a rate
         #"Analysis of information showed that the system was sampling
         # too hard at some rates. For those rates that never work
         # (54mb, 500m range) there is no point in sending 10 sample
@@ -182,26 +185,29 @@ def apply_rate(cur_time): #cur_time is in nanoseconds
             # out lots of sampling frames, which would result          
             # in a large throughput loss."
             sample_count += delta  - (len(rates)*2)
-            
 
-        randrate = 1
[…]

### 2013-08-07 pysim/minstrel.py
Added in the sample_skipped mechanism in the kernel

@@ -92,6 +92,7 @@ class Rate:
         self.ack = tx_time(rate, 10) #microseconds, assumes 1mbps ack rate
 
         #what is the difference between these?
+        self.sample_skipped = 0
         self.sample_limit = -1
         self.retry_count = 1
 
@@ -190,13 +191,14 @@ def apply_rate(cur_time): #cur_time is in nanoseconds
         # here, instead using a pregenerated table of "random" numbers.
         # See net/mac80211/rc80211_minstrel.c :: init_sample_table
 
-        if randrate < bestThruput:
         # TODO: Use the mechanism the kernel uses
         randrate = random.choice(rates.keys())
         
 	#"Decide if direct ( 1st mrr stage) or indirect (2nd mrr
 	# stage) rate sampling method should be used.  Respect such
 	# rates that are not sampled for 20 interations."
+
+        if randrate < bestThruput and rates[randrate].sample_skipped < 20:
             #"Only use IEEE80211_TX_CTL_RATE_CTRL_PROBE to mark
             # packets that have the sampling rate deferred to the
             # second MRR stage. Increase the sample counter only if
@@ -264,9 +266,12 @@ def update_stats(timestamp):
     for i, br in rates.items():
         if br.attempts: #prevents divide by 0
             p = br.success * 18000 // br.attempts
+            br.sample_skipped = 0
             br.succ_hist += br.success
             br.att_hist += br.attempts
             br.ewma.feed(timestamp, p)
+        else:
+            br.sample_skipped += 1
         p = br.ewma.read()
         
 

### 2013-08-07 pysim/minstrel.py
Minstrel gets major overhaul: more commenting, cleaner code, much faster.
Based on the source in net/mac80211/rc80211_minstrel.h, from Kernel 3.10.5

@@ -8,16 +8,15 @@ from __future__ import division
 import random
 import common
 import math
+import collections
 from common import ieee80211_to_idx
 
 packet_count = 0 #number of packets sent over link
-nsuccess = 0 #number of packets sent successfully 
-nlookaround = 0
-currRate = 54  #current best bitRate
-bestThruput = 12
-nextThruput = 11
-bestProb = 2
-lowestRate = 1
+
+# Rates
+rate_struct = collections.namedtuple("Rates", ["best", "next", "prob", "base"])
+choices = rate_struct(12, 11, 2, 1)
+
 time_last_called = 0
 cw_min = 15
 cw_max = 1023
@@ -29,6 +28,15 @@ sample_count = 0
 sample_deferred = 0
 
 SAMPLING_RATIO = 10
+MINSTREL_SCALE = 16
+EWMA_LEVEL = 96
+EWMA_DIV = 128
+
+def MINSTREL_FRAC(val, div):
+    return (val << MINSTREL_SCALE) // div
+
+def MINSTREL_TRUNC(val):
+    return val >> MINSTREL_SCALE
 
 def tx_time(mbps, length=1200): #rix is index to RATES, length in bytes
     '''
@@ -82,8 +90,7 @@ class Rate:
         self.rate = rate #in mbps
         self.throughput = 0 #in bits per second
         self.last_update = 0.0 #timestamp of last update, ns(?)
-        #ewma obj. can return probability of success. if you ask nicely.
-        self.ewma = common.EWMA(0.0, 100e6, 0.75) #100e6 = 100 ms,
+        self.probability = 0
         self.succ_hist = 0 #total successes ever
         self.att_hist = 0 #total xmission attempts ever
         self.success = 0  #number of successful xmissions in cur time interval
@@ -115,6 +122,10 @@ class Rate:
                                                       max_retry)
         self.adjusted_retry_count = self.retry_count #Max retrans. used for probing
 
+    def ewma(self, new, weight):
+        old = self.probability
+        self.probability = (new * (EWMA_DIV - weight) + old * weight) // EWMA_DIV;
+
     def __repr__(self):
         return ("Bitrate %r mbps: \n"
                 "  attempts: %r \n"
@@ -123,7 +134,7 @@ class Rate:
                 "  probSuccess: %r \n"
                 "  losslessTX: %r microseconds"
                 % (self.rate, self.attempts, self.success, 
-                   self.throughput, self.ewma.read(), self.losslessTX))
[…]

### 2013-08-07 pysim/minstrel.py
Minor Minstrel cleanups and fixes

@@ -1,7 +1,6 @@
-# Colleen Josephson, 2013
-# This file attempts to implement the minstrel rate control algorithm from the
-# 3.3.8 linux kernel. We assume multi-rate retry capabilities, so we omit 
-# the code for the non-mrr case. 
+# This file attempts to implement the minstrel rate control algorithm
+# from the 3.10.5 Linux kernel. We assume multi-rate retry
+# capabilities, so we omit the code for the non-mrr case.
 
 from __future__ import division
 
@@ -11,22 +10,18 @@ import math
 import collections
 from common import ieee80211_to_idx
 
-packet_count = 0 #number of packets sent over link
+packet_count = 0
+sample_count = 0
+sample_deferred = 0
 
 # Rates
 rate_struct = collections.namedtuple("Rates", ["best", "next", "prob", "base"])
 choices = rate_struct(12, 11, 2, 1)
 
 time_last_called = 0
-cw_min = 15
-cw_max = 1023
-segment_size = 6000
-max_retry = 7 #safe default specified in kernel code
 probeFlag = False
 
-sample_count = 0
-sample_deferred = 0
-
+MAX_RETRY = 7
 SAMPLING_RATIO = 10
 MINSTREL_SCALE = 16
 EWMA_LEVEL = 96
@@ -111,15 +106,15 @@ class Rate:
             tx_time_single = self.ack + self.losslessTX
 
             #contention window
-            cw = cw_min
+            cw = 15 # 15 = CWmin
             tx_time_single += (9* cw) >> 1;
-            cw = min((cw << 1) | 1, cw_max)
-            
+            cw = min((cw << 1) | 1, 1023) # 1023 = CWmax
+
             tx_time_ += tx_time_single
 
             self.retry_count += 1
-            condition = (tx_time_ < segment_size) and (self.retry_count  < 
-                                                      max_retry)
+            # 6000 == segment size
+            condition = (tx_time_ < 6000) and (self.retry_count  < MAX_RETRY)
         self.adjusted_retry_count = self.retry_count #Max retrans. used for probing
 
     def ewma(self, new, weight):
@@ -133,12 +128,16 @@ class Rate:
                 "  thruput: %r microseconds \n"
                 "  probSuccess: %r \n"
                 "  losslessTX: %r microseconds"
-                % (self.rate, self.attempts, self.success, 
+                % (self.rate, self.attempts, self.success,
                    self.throughput, self.probability, self.losslessTX))
 
[…]

### 2013-08-07 pysim/common.py
Moved tx_time into harness
This way the harness's tx_time is independent of the algorithms themselves.

@@ -85,35 +85,3 @@ class BalancedEWMA:
             return int(self.val * 18000)
         else:
             return None
-
-# The average back-off period, in microseconds, for up to 7 attempts
-# of a 802.11b unicast packet.
-
-BACKOFF = { "ofdm": [0], "ds": [0], "dsss": [0] }
-for i in range(5, 11):
-    BACKOFF["ds"].append(int(((2**i) - 1) * (20 / 2)))
-for i in range(5, 11):
-    BACKOFF["dsss"].append(int(((2**i) - 1) * (9 / 2)))
-for i in range(4, 11):
-    BACKOFF["ofdm"].append(int(((2**i) - 1) * (9 / 2)))
-
-def backoff(rix, attempt):
-    return BACKOFF[RATES[rix].phy][min(attempt, len(BACKOFF) - 1)]
-
-def difs(rix):
-    version = "g" if RATES[rix].phy == "ofdm" else "b"
-    return 50 if version == "b" else 28
-
-def tx_time(rix, tries, nbytes):
-    # From the SampleRate paper.  See samplerate.py for annotated version.
-    bitrate = RATES[rix].dot11_rate
-    version = "g" if RATES[rix].phy == "ofdm" else "b"
-    sifs = 10 if version == "b" else 9
-    ack = 304 # Somehow 6mb acks aren't used
-    header = 192 if bitrate == 1 else 96 if version == "b" else 20
-    
-    assert tries > 0, "Cannot try a non-positive number of times"
-
-    us = tries * (sifs + ack + header + (nbytes * 8 / bitrate))
-    return us * 1000
-

### 2013-08-07 pysim/harness.py

@@ -7,25 +7,24 @@ import random
 
 DEBUG = "DEBUG" in os.environ
 
-WINDOW = 1e7 # 10ms
-
 def load_data(source):
     return eval(open(source, "rt").read())
 
+WINDOW = 1e7 # 10ms
 LUOPS = 0
 def packet_stats(data, cache, time, rate):
     global LUOPS
-    
+
     txs = []
 
     while not txs:
         idx, w = cache[rate]
         cache[rate][0] = 0
-        
+
         for i in range(idx, len(data[rate])):
             t, success, delay = data[rate][i]
             LUOPS += 1
-            
+
             if abs(t - time) < w:
                 txs.append((success, delay))
                 if not cache[rate][0]:
@@ -37,10 +36,37 @@ def packet_stats(data, cache, time, rate):
 
         cache[rate][1] *= 2
 
-    successful = [tx[1] for tx in txs if tx[0]]
+    successful = [tx for tx in txs if tx[0]]
 
-    return (len(successful) / len(txs), 
-            sum(successful) / sum([tx[1] for tx in txs]))
+    return len(successful) / len(txs)
+
+
+# The test harness uses a calculation of transmission time based on
+# that in the SampleRate paper
+BACKOFF = { "ofdm": [0], "ds": [0], "dsss": [0] }
+for i in range(5, 11):
+    BACKOFF["ds"].append(int(((2**i) - 1) * (20 / 2)))
+for i in range(5, 11):
+    BACKOFF["dsss"].append(int(((2**i) - 1) * (9 / 2)))
+for i in range(4, 11):
+    BACKOFF["ofdm"].append(int(((2**i) - 1) * (9 / 2)))
+
+def backoff(rix, attempt):
+    return BACKOFF[common.RATES[rix].phy][min(attempt, len(BACKOFF) - 1)]
+
+def difs(rix):
+    version = "g" if common.RATES[rix].phy == "ofdm" else "b"
+    return 50 if version == "b" else 28
+
+def tx_time(rix, nbytes):
+    # From the SampleRate paper.  See samplerate.py for annotated version.
+    bitrate = common.RATES[rix].dot11_rate
+    version = "g" if common.RATES[rix].phy == "ofdm" else "b"
+    sifs = 10 if version == "b" else 9
+    ack = 304 # Somehow 6mb acks aren't used
+    header = 192 if bitrate == 1 else 96 if version == "b" else 20
+
[…]

### 2013-08-07 pysim/common.py
Rebase minstrel_bewma; remove EWMA implementations from common.

@@ -28,60 +28,3 @@ def ieee80211_to_idx(mbps):
     else:
         raise ValueError("No bitrate with throughput {} Mbps exists".format(mbps))
 
-
-class EWMA:
-    def __init__(self, time, time_step, pval):
-        self.p = pval
-        self.time = time
-        self.step = time_step
-        self.val = None
-
-    def feed(self, time, val):
-        if self.val is None:
-            newval = val
-        else:
-            p = self.p
-            newval = self.val * p + val * (1 - p)
-
-        self.val = int(newval)
-        self.time = time
-
-    def read(self):
-        if self.val is not None:
-            return self.val
-        else: 
-            return None
-
-class BalancedEWMA:
-    def __init__(self, time, time_step, pval):
-        self.p = pval
-        self.time = time
-        self.step = time_step
-
-        self.blocks = 0
-        self.denom = 0
-        self.val = None
-
-    def feed(self, time, num, denom):
-        if self.blocks == 0:
-            self.denom = denom
-            newval = num / denom
-        else:
-            avg_block = self.denom / self.blocks
-            block_weight = denom / avg_block
-            relweight = self.p / (1 - self.p)
-
-            prob = num / denom
-
-            newval = (self.val * relweight + prob * block_weight) / \
-                     (relweight + block_weight)
-
-        self.blocks += 1
-        self.val = newval
-        self.time = time
-
-    def read(self):
-        if self.val is not None:
-            return int(self.val * 18000)
-        else:
-            return None

### 2013-08-07 pysim/minproved.py

@@ -6,6 +6,40 @@ from random import choice
 import common
 from common import ieee80211_to_idx
 
+class BalancedEWMA:
+    def __init__(self, time, time_step, pval):
+        self.p = pval
+        self.time = time
+        self.step = time_step
+
+        self.blocks = 0
+        self.denom = 0
+        self.val = None
+
+    def feed(self, time, num, denom):
+        if self.blocks == 0:
+            self.denom = denom
+            newval = num / denom
+        else:
+            avg_block = self.denom / self.blocks
+            block_weight = denom / avg_block
+            relweight = self.p / (1 - self.p)
+
+            prob = num / denom
+
+            newval = (self.val * relweight + prob * block_weight) / \
+                     (relweight + block_weight)
+
+        self.blocks += 1
+        self.val = newval
+        self.time = time
+
+    def read(self):
+        if self.val is not None:
+            return int(self.val * 18000)
+        else:
+            return None
+
 npkts = 0 #number of packets sent over link
 nsuccess = 0 #number of packets sent successfully 
 nlookaround = 0
@@ -94,7 +128,7 @@ class Rate:
         self.throughput = 0 #in bits per second
         self.last_update = 0.0 #timestamp of last update, ns(?)
         #ewma obj. can return probability of success. if you ask nicely.
-        self.ewma = common.BalancedEWMA(0.0, 100e6, 0.75) #100e6 = 100 ms
+        self.ewma = BalancedEWMA(0.0, 100e6, 0.75) #100e6 = 100 ms
         self.succ_hist = 0 #total successes ever
         self.att_hist = 0 #total xmission attempts ever
         self.success = 0  #number of successful xmissions in cur time interval

### 2013-08-07 pysim/minstrel.py

@@ -34,45 +34,39 @@ def MINSTREL_TRUNC(val):
     return val >> MINSTREL_SCALE
 
 def tx_time(mbps, length=1200): #rix is index to RATES, length in bytes
-    '''
-    Adapted from 802.11 util.c ieee80211_frame_duration()
+    # Adapted from 802.11 util.c ieee80211_frame_duration()
 
-    calculate duration (in microseconds, rounded up to next higher
-    integer if it includes a fractional microsecond) to send frame of
-    len bytes (does not include FCS) at the given rate. Duration will
-    also include SIFS.
-    '''
+    #"calculate duration (in microseconds, rounded up to next higher
+    # integer if it includes a fractional microsecond) to send frame of
+    # len bytes (does not include FCS) at the given rate. Duration will
+    # also include SIFS."
 
     rateinfo = common.RATES[common.ieee80211_to_idx(mbps)]
 
     if rateinfo.phy == "ofdm":
-        '''
-        OFDM:
-        N_DBPS = DATARATE x 4
-        N_SYM = Ceiling((16+8xLENGTH+6) / N_DBPS)
-        (16 = SIGNAL time, 6 = tail bits)
-        TXTIME = T_PREAMBLE + T_SIGNAL + T_SYM x N_SYM + Signal Ext
+        #"OFDM:
+        # N_DBPS = DATARATE x 4
+        # N_SYM = Ceiling((16+8xLENGTH+6) / N_DBPS)
+        # (16 = SIGNAL time, 6 = tail bits)
+        # TXTIME = T_PREAMBLE + T_SIGNAL + T_SYM x N_SYM + Signal Ext
 
-        T_SYM = 4 usec
-        802.11a - 17.5.2: aSIFSTime = 16 usec
-        802.11g - 19.8.4: aSIFSTime = 10 usec + signal ext = 6 usec
-        '''
+        # T_SYM = 4 usec
+        # 802.11a - 17.5.2: aSIFSTime = 16 usec
+        # 802.11g - 19.8.4: aSIFSTime = 10 usec + signal ext = 6 usec"
         dur = 16 # SIFS + signal ext */
         dur += 16 # 17.3.2.3: T_PREAMBLE = 16 usec */
         dur += 4 # 17.3.2.3: T_SIGNAL = 4 usec */
         dur += 4 * (math.ceil((16+8*(length+4)+6)/(4*mbps))+1) # T_SYM x N_SYM
 
     else:
-        '''
-        802.11b or 802.11g with 802.11b compatibility:
[…]

### 2013-08-07 pysim/minstrel_bewma.py

@@ -1,31 +1,37 @@
-# Colleen Josephson, 2013
-# This file attempts to implement the minstrel rate control algorithm from the
-# 3.3.8 linux kernel. We assume multi-rate retry capabilities, so we omit 
-# the code for the non-mrr case. 
+# This file attempts to implement the minstrel rate control algorithm
+# from the 3.10.5 Linux kernel. We assume multi-rate retry
+# capabilities, so we omit the code for the non-mrr case.
 
-from random import choice 
+from __future__ import division
+
+import random
 import common
 import math
+import collections
 from common import ieee80211_to_idx
 
-npkts = 0 #number of packets sent over link
-nsuccess = 0 #number of packets sent successfully 
-nlookaround = 0
-NBYTES = 1500 #constant
-currRate = 54  #current best bitRate
-NRETRIES = 20
-bestThruput = 12
-nextThruput = 11
-bestProb = 2
-lowestRate = 1
+packet_count = 0
+sample_count = 0
+sample_deferred = 0
+
+# Rates
+rate_struct = collections.namedtuple("Rates", ["best", "next", "prob", "base"])
+choices = rate_struct(12, 11, 2, 1)
+
 time_last_called = 0
-cw_min = 15
-cw_max = 1023
-segment_size = 6000
-max_retry = 7 #safe default specified in kernel code
-np = 0
-nd = 0
 probeFlag = False
+
+MAX_RETRY = 7
+SAMPLING_RATIO = 10
+MINSTREL_SCALE = 16
+EWMA_LEVEL = 96
+EWMA_DIV = 128
+
+def MINSTREL_FRAC(val, div):
+    return (val << MINSTREL_SCALE) // div
+
+def MINSTREL_TRUNC(val):
+    return val >> MINSTREL_SCALE
 
 def tx_time(mbps, length=1200): #rix is index to RATES, length in bytes
     '''
@@ -49,8 +55,7 @@ def tx_time(mbps, length=1200): #rix is index to RATES, length in bytes
 
         T_SYM = 4 usec
         802.11a - 17.5.2: aSIFSTime = 16 usec
-        802.11g - 19.8.4: aSIFSTime = 10 usec +
-        signal ext = 6 usec
+        802.11g - 19.8.4: aSIFSTime = 10 usec + signal ext = 6 usec
         '''
         dur = 16 # SIFS + signal ext */
         dur += 16 # 17.3.2.3: T_PREAMBLE = 16 usec */
[…]

### 2013-08-07 pysim/common.py
Rename Bitrate.dot11_rate to Bitrate.mbps

@@ -3,7 +3,7 @@ import collections
 import math
 
 BitRate = collections.namedtuple("BitRate", ["phy", "kbps", "user_kbps",
-                                             "code", "dot11_rate"])
+                                             "code", "mbps"])
 
 # The 802.11, 802.11b, and 802.11g rates
 RATES = [
@@ -22,7 +22,7 @@ RATES = [
 ]
 
 def ieee80211_to_idx(mbps):
-    opts = [i for i, rate in enumerate(RATES) if rate.dot11_rate == mbps]
+    opts = [i for i, rate in enumerate(RATES) if rate.mbps == mbps]
     if opts:
         return opts[0]
     else:

### 2013-08-07 pysim/constant.py

@@ -7,7 +7,7 @@ RATE=float(os.environ["RATE"]) if "RATE" in os.environ else 1
 try:
     IDX = common.ieee80211_to_idx(RATE)
 except ValueError:
-    opts = [str(rate.dot11_rate) for rate in common.RATES]
+    opts = [str(rate.mbps) for rate in common.RATES]
     print("Invalid rate.  Options are: {}".format(", ".join(opts)))
     exit()
 else:

### 2013-08-07 pysim/harness.py

@@ -60,7 +60,7 @@ def difs(rix):
 
 def tx_time(rix, nbytes):
     # From the SampleRate paper.  See samplerate.py for annotated version.
-    bitrate = common.RATES[rix].dot11_rate
+    bitrate = common.RATES[rix].mbps
     version = "g" if common.RATES[rix].phy == "ofdm" else "b"
     sifs = 10 if version == "b" else 9
     ack = 304 # Somehow 6mb acks aren't used
@@ -206,6 +206,6 @@ if __name__ == "__main__":
         tries, successes = info
         if not tries: continue
 
-        mbps = common.RATES[rate_idx].dot11_rate
+        mbps = common.RATES[rate_idx].mbps
         print("{:>5} Mbps : {:>4} tries ({:.0%} success rate)".format(
             mbps, tries, successes/tries))

### 2013-08-07 pysim/minproved.py

@@ -22,7 +22,7 @@ class BalancedEWMA:
             newval = num / denom
         else:
             avg_block = self.denom / self.blocks
-            block_weight = denom / avg_block
+            block_weight = denom / avg_block if avg_block else 1
             relweight = self.p / (1 - self.p)
 
             prob = num / denom
@@ -268,7 +268,7 @@ def process_feedback(status, timestamp, delay, tries):
     for t in range(len(tries)):
         (bitrate, br_tries) = tries[t]
         if br_tries > 0:
-            bitrate = common.RATES[bitrate].dot11_rate
+            bitrate = common.RATES[bitrate].mbps
             #if bitrate == 1:
             
             br = rates[bitrate]

### 2013-08-07 pysim/minstrel.py

@@ -223,7 +223,7 @@ def process_feedback(status, timestamp, delay, tries):
     for t in range(len(tries)):
         (bitrate, br_tries) = tries[t]
         if br_tries > 0:
-            bitrate = common.RATES[bitrate].dot11_rate
+            bitrate = common.RATES[bitrate].mbps
             #if bitrate == 1:
 
             br = rates[bitrate]

### 2013-08-07 pysim/minstrel_bewma.py

@@ -238,7 +238,7 @@ def process_feedback(status, timestamp, delay, tries):
     for t in range(len(tries)):
         (bitrate, br_tries) = tries[t]
         if br_tries > 0:
-            bitrate = common.RATES[bitrate].dot11_rate
+            bitrate = common.RATES[bitrate].mbps
             #if bitrate == 1:
 
             br = rates[bitrate]

### 2013-08-07 pysim/samplerate.py

@@ -140,7 +140,7 @@ def process_feedback(status, timestamp, delay, tries):
     global currRate, npkts, nsuccess, NBYTES
     (bitrate, nretries) = tries[0]
     nretries -= 1
-    bitrate = common.RATES[bitrate].dot11_rate
+    bitrate = common.RATES[bitrate].mbps
 
     #"Calculate the transmission time for the packet based on the
     # bit-rate and number of retries using Equation 5.1 below."

### 2013-08-07 pysim/samplerate_longer.py

@@ -150,7 +150,7 @@ def process_feedback(status, timestamp, delay, tries):
     global currRate, npkts, nsuccess, NBYTES
     (bitrate, nretries) = tries[0]
     nretries -= 1
-    bitrate = common.RATES[bitrate].dot11_rate
+    bitrate = common.RATES[bitrate].mbps
 
     #"Calculate the transmission time for the packet based on the
     # bit-rate and number of retries using Equation 5.1 below."

### 2013-08-07 pysim/samplerate_retry.py

@@ -152,7 +152,7 @@ def process_feedback(status, timestamp, delay, tries):
     global currRate, npkts, nsuccess, NBYTES
     (bitrate, nretries) = tries[0]
     nretries -= 1
-    bitrate = common.RATES[bitrate].dot11_rate
+    bitrate = common.RATES[bitrate].mbps
 
     #"Calculate the transmission time for the packet based on the
     # bit-rate and number of retries using Equation 5.1 below."

### 2013-08-07 pysim/samplerate.py
Make samplerate not call ieee80211_to_idx

@@ -15,10 +15,6 @@ NRETRIES = 1
 
 npkts = 0 # number of packets sent over link
 nsuccess = 0 #number of packets sent successfully 
-currRate = 54 #current best bitRate
-
-def bitrate_type(bitrate):
-    return common.RATES[ieee80211_to_idx(bitrate)].phy
 
 # The average back-off period, in microseconds, for up to 8 attempts
 # of a 802.11b unicast packet.
@@ -30,7 +26,7 @@ backoff = [0, 155, 315, 635, 1275, 2555, 5115]
 # following equation based on the 802.11 unicast retransmission
 # mechanism detailed in Section 2.2"
 
-def tx_time(bitrate, retries, nbytes):
+def tx_time(rix, retries, nbytes):
     # tx_time(b, r, n) = difs + backoff[r] + \
     #                  + (r + 1)*(sifs + ack + header + (n * 8/b)
 
@@ -42,16 +38,18 @@ def tx_time(bitrate, retries, nbytes):
     # 802.11b packets, 96 for other 802.11b bit-rates, and 20 for
     # 802.11a/g bit-rates. backoff(r) is calculated using the table"
 
-    version = "g" if bitrate_type(bitrate) == "ofdm" else "b"
+    rate = common.RATES[rix]
+    version = "g" if rate.phy == "ofdm" else "b"
+
     difs = 50 if version == "b" else 28
     sifs = 10 if version == "b" else 9
     ack = 304 # Somehow 6mb acks aren't used
-    header = 192 if bitrate == 1 else 96 if version == "b" else 20
+    header = 192 if rate.code == 0 else 96 if version == "b" else 20
 
     backoff_r = backoff[retries] if retries < len(backoff) else backoff[-1]
 
     return difs + backoff_r + \
-        (retries + 1) * (sifs + ack + header + (nbytes * 8 / bitrate))
+        (retries + 1) * (sifs + ack + header + (nbytes * 8 / rate.mbps))
 
 Packet = collections.namedtuple("Packet", ["time_sent", "success",
                                            "txTime", "rate"])
@@ -59,6 +57,7 @@ Packet = collections.namedtuple("Packet", ["time_sent", "success",
 class Rate:
     def __init__(self, rate):
         self.rate = rate #in mbps
+        self.idx = ieee80211_to_idx(rate)
         self.success = 0
         self.tries = 0
         self.pktAcked = 0
@@ -66,19 +65,8 @@ class Rate:
[…]

### 2013-08-07 pysim/samplerate.py
Rename poorly-named constant

@@ -11,7 +11,7 @@ from common import ieee80211_to_idx
 
 # Constants: send 1500 bytes at a time, with 1 try each in the MRR
 NBYTES = 1500
-NRETRIES = 1
+NTRIES = 1
 
 npkts = 0 # number of packets sent over link
 nsuccess = 0 #number of packets sent successfully 
@@ -93,7 +93,7 @@ def apply_rate(cur_time):
         for i, r in sorted(rates.items(), reverse=True):
             if r.succFails < 4:
                 currRate = r
-                return [(r.idx, NRETRIES)]
+                return [(r.idx, NTRIES)]
 
     # Every 10 packets, select a random non-failing bit rate w/ better avg tx
     #"If the number of packets sent over the link is a multiple of ten,"
@@ -109,12 +109,12 @@ def apply_rate(cur_time):
 
         if len(eligible) > 0:
             sampleRate = random.choice(eligible)
-            return [(sampleRate.idx, NRETRIES)]
+            return [(sampleRate.idx, NTRIES)]
 
     #"Otherwise, send packet at the bit-rate that has the lowest avg
     # transmission time" Trusts that currRate is properly maintained
     # to be lowest avgTX
-    return [(currRate.idx, NRETRIES)]
+    return [(currRate.idx, NTRIES)]
 
 
 #"When process f eedback() runs, it updates information that tracks

### 2013-08-07 pysim/samplerate.py
SampleRate rates table indexed by rix, not by mbps

@@ -55,9 +55,11 @@ Packet = collections.namedtuple("Packet", ["time_sent", "success",
                                            "txTime", "rate"])
 
 class Rate:
-    def __init__(self, rate):
-        self.rate = rate #in mbps
-        self.idx = ieee80211_to_idx(rate)
+    def __init__(self, rix):
+        self.info = common.RATES[rix]
+        self.idx = rix
+        self.rate = self.info.mbps #in mbps
+
         self.success = 0
         self.tries = 0
         self.pktAcked = 0
@@ -76,8 +78,8 @@ class Rate:
 # DBPSK/DQPSK+DSSS for 1 and 2 Mbit/s.  Even though 802.11g operates
 # in the same frequency band as 802.11b, it can achieve higher data
 # rates because of its heritage to 802.11a.
-rates = dict((r, Rate(r)) for r in [1, 2, 5.5, 6, 9, 11, 12, 18, 24, 36, 48, 54])
-currRate = rates[54] #current best bitRate
+rates = [Rate(i) for i in range(len(common.RATES))]
+currRate = rates[-1] #current best bitRate
 
 #multi-rate retry returns an array of (rate, ntries) for the next n packets
 def apply_rate(cur_time):
@@ -90,21 +92,21 @@ def apply_rate(cur_time):
     #"If no packets have been successfully acknowledged, return the
     # highest bit-rate that has not had 4 successive failures."
     if nsuccess == 0:
-        for i, r in sorted(rates.items(), reverse=True):
-            if r.succFails < 4:
-                currRate = r
-                return [(r.idx, NTRIES)]
+        for rate in sorted(rates, key=lambda rate: rate.rate, reverse=True):
+            if rate.succFails < 4:
+                currRate = rate
+                return [(rate.idx, NTRIES)]
 
     # Every 10 packets, select a random non-failing bit rate w/ better avg tx
     #"If the number of packets sent over the link is a multiple of ten,"
     if (nsuccess != 0) and (npkts%10 == 0):
         #"select a random bit-rate from the bit-rates"
-        cavgTX = rates[currRate.rate].avgTX
+        cavgTX = rates[currRate.idx].avgTX
 
         #"that have not failed four successive times and that have a
         # minimum packet transmission time lower than the current
         # bit-rate's average transmission time."
-        eligible = [r for i, r in rates.items()
+        eligible = [r for r in rates
                     if r.losslessTX < cavgTX and r.succFails < 4]
[…]
